### 2023-08-21 doc/paper.tex
Add inital doc files
Signed-off-by: Jonatan Hanssen <jonatan.hanssen@outlook.com>

\documentclass[onecolumn,10pt,cleanfoot]{asme2ej}

\usepackage{graphicx} %% for loading jpg figures
\usepackage{bm}
\usepackage{nicefrac}
\usepackage{mathtools}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{parskip}
\usepackage{listings}
\usepackage{tablefootnote}
\usepackage{float}
\usepackage{xcolor}
\usepackage{xurl}

\title{Neural networks for solving regression and classification problems}

%%% first author
\author{Jonatan H. Hanssen
    \affiliation{
	Master Student, Robotics and \\
	Intelligent Systems\\ \\[-10pt]
	Department of Informatics\\ \\[-10pt]
	The faculty of Mathematics and \\
	Natural Sciences\\ \\[-10pt]
    Email: jonatahh@ifi.uio.no
    }
}

\author{Adrian Duric
    \affiliation{
	Master Student, Robotics and \\
	Intelligent Systems\\ \\[-10pt]
	Department of Informatics\\ \\[-10pt]
	The faculty of Mathematics and \\
	Natural Sciences\\ \\[-10pt]
    Email: adriandu@ifi.uio.no
    }
}

\author{Gregor Kajda
    \affiliation{
	Master Student, Robotics and \\
	Intelligent Systems\\ \\[-10pt]
	Department of Informatics\\ \\[-10pt]
	The faculty of Mathematics and \\
	Natural Sciences\\ \\[-10pt]
    Email: grzegork@ifi.uio.no
    }
}

\begin{document}

\maketitle

\section{Abstract}

\section{Introduction}

\section{Method}

\subsection{Theory}

This sections covers the theoretic background which underlies the methods used in this paper.

\subsubsection{Linear Regression}

\subsection{Implementation}

Our implementation can be found at \texttt{github.com/jonatan-hanssen/in5490-project.git}.

\section{Results}

\section{Bibliography}

\end{document}

### 2023-08-21 doc/paper.tex
Update title
Signed-off-by: Jonatan Hanssen <jonatan.hanssen@outlook.com>
Update to recommended template
Signed-off-by: Jonatan Hanssen <jonatan.hanssen@outlook.com>

@@
-\documentclass[onecolumn,10pt,cleanfoot]{asme2ej}
+\documentclass[conference]{IEEEtran}
+\IEEEoverridecommandlockouts
+% The preceding line is only needed to identify funding in the first footnote. If that is unneeded, please comment it out.
+\usepackage{cite}
+\usepackage{amsmath,amssymb,amsfonts}
+\usepackage{algorithmic}
+\usepackage{graphicx}
+\usepackage{textcomp}
+\usepackage{xcolor}
+\def\BibTeX{{\rm B\kern-.05em{\sc i\kern-.025em b}\kern-.08em
+    T\kern-.1667em\lower.7ex\hbox{E}\kern-.125emX}}
+\begin{document}
 
-\usepackage{graphicx} %% for loading jpg figures
-\usepackage{bm}
-\usepackage{nicefrac}
-\usepackage{mathtools}
-\usepackage{amssymb}
-\usepackage{amsmath}
-\usepackage{parskip}
-\usepackage{listings}
-\usepackage{tablefootnote}
-\usepackage{float}
-\usepackage{xcolor}
-\usepackage{xurl}
-
-\title{Neural networks for solving regression and classification problems}
-
-%%% first author
-\author{Jonatan H. Hanssen
-    \affiliation{
-	Master Student, Robotics and \\
-	Intelligent Systems\\ \\[-10pt]
-	Department of Informatics\\ \\[-10pt]
-	The faculty of Mathematics and \\
-	Natural Sciences\\ \\[-10pt]
-    Email: jonatahh@ifi.uio.no
-    }
+\title{Conference Paper Title*\\
+{\footnotesize \textsuperscript{*}Note: Sub-titles are not captured in Xplore and
+should not be used}
+\thanks{Identify applicable funding agency here. If none, delete this.}
 }
 
-\author{Adrian Duric
-    \affiliation{
-	Master Student, Robotics and \\
-	Intelligent Systems\\ \\[-10pt]
-	Department of Informatics\\ \\[-10pt]
-	The faculty of Mathematics and \\
-	Natural Sciences\\ \\[-10pt]
-    Email: adriandu@ifi.uio.no
+\author{\IEEEauthorblockN{1\textsuperscript{st} Given Name Surname}
+\IEEEauthorblockA{\textit{dept. name of organization (of Aff.)} \\
+\textit{name of organization (of Aff.)}\\
+City, Country \\
+email address or ORCID}
+\and
+\IEEEauthorblockN{2\textsuperscript{nd} Given Name Surname}
+\IEEEauthorblockA{\textit{dept. name of organization (of Aff.)} \\
+\textit{name of organization (of Aff.)}\\
+City, Country \\
+email address or ORCID}
+\and
+\IEEEauthorblockN{3\textsuperscript{rd} Given Name Surname}
+\IEEEauthorblockA{\textit{dept. name of organization (of Aff.)} \\
+\textit{name of organization (of Aff.)}\\
+City, Country \\
+email address or ORCID}
+\and
+\IEEEauthorblockN{4\textsuperscript{th} Given Name Surname}
+\IEEEauthorblockA{\textit{dept. name of organization (of Aff.)} \\
+\textit{name of organization (of Aff.)}\\
+City, Country \\
+email address or ORCID}
+\and
+\IEEEauthorblockN{5\textsuperscript{th} Given Name Surname}
+\IEEEauthorblockA{\textit{dept. name of organization (of Aff.)} \\
+\textit{name of organization (of Aff.)}\\
+City, Country \\
+email address or ORCID}
+\and
+\IEEEauthorblockN{6\textsuperscript{th} Given Name Surname}
+\IEEEauthorblockA{\textit{dept. name of organization (of Aff.)} \\
+\textit{name of organization (of Aff.)}\\
+City, Country \\
+email address or ORCID}
 }
-}
-
-\author{Gregor Kajda
-    \affiliation{
-	Master Student, Robotics and \\
-	Intelligent Systems\\ \\[-10pt]
-	Department of Informatics\\ \\[-10pt]
-	The faculty of Mathematics and \\
-	Natural Sciences\\ \\[-10pt]
-    Email: grzegork@ifi.uio.no
-    }
-}
-
-\begin{document}
 
@@
 
-\section{Abstract}
+\begin{abstract}
+This document is a model and instructions for \LaTeX.
+This and the IEEEtran.cls file define the components of your paper [title, text, heads, etc.]. *CRITICAL: Do Not Use Symbols, Special Characters, Footnotes, 
+or Math in Paper Title or Abstract.
+\end{abstract}
+
+\begin{IEEEkeywords}
+component, formatting, style, styling, insert
+\end{IEEEkeywords}
 
 \section{Introduction}
+This document is a model and instructions for \LaTeX.
+Please observe the conference page limits.
 
-\section{Method}
+\section{Ease of Use}
 
-\subsection{Theory}
+\subsection{Maintaining the Integrity of the Specifications}
 
-This sections covers the theoretic background which underlies the methods used in this paper.
+The IEEEtran class file is used to format your paper and style the text. All margins, 
+column widths, line spaces, and text fonts are prescribed; please do not 
+alter them. You may note peculiarities. For example, the head margin
+measures proportionately more than is customary. This measurement 
+and others are deliberate, using specifications that anticipate your paper 
+as one part of the entire proceedings, and not as an independent document. 
+Please do not revise any of the current designations.
 
-\subsubsection{Linear Regression}
+\section{Prepare Your Paper Before Styling}
+Before you begin to format your paper, first write and save the content as a 
+separate text file. Complete all content and organizational editing before 
+formatting. Please note sections \ref{AA}--\ref{SCM} below for more information on 
+proofreading, spelling and grammar.
 
-\subsection{Implementation}
+Keep your text and graphic files separate until after the text has been 
+formatted and styled. Do not number text heads---{\LaTeX} will do that 
+for you.
 
-Our implementation can be found at \texttt{github.com/jonatan-hanssen/in5490-project.git}.
+\subsection{Abbreviations and Acronyms}\label{AA}
+Define abbreviations and acronyms the first time they are used in the text, 
+even after they have been defined in the abstract. Abbreviations such as 
+IEEE, SI, MKS, CGS, ac, dc, and rms do not have to be defined. Do not use 
+abbreviations in the title or heads unless they are unavoidable.
 
-\section{Results}
+\subsection{Units}
+\begin{itemize}
+\item Use either SI (MKS) or CGS as primary units.
+(SI units are encouraged.) English units may be used as secondary units (in parentheses).
+An exception would be the use of English units as identifiers in trade, such as ``3.5-inch disk drive''.
+\item Avoid combining SI and CGS units, such as current in amperes and magnetic field in oersteds.
+This often leads to confusion because equations do not balance dimensionally.
+If you must use mixed units, clearly state the units for each quantity that you use in an equation.
+\item Do not mix complete spellings and abbreviations of units: ``Wb/m\textsuperscript{2}'' or ``webers per square meter'', not ``webers/m\textsuperscript{2}''.
+Spell out units when they appear in text: ``.
+.
+.
+a few henries'', not ``.
+.
+.
+a few H''.
+\item Use a zero before decimal points: ``0.25'', not ``.25''. Use ``cm\textsuperscript{3}'', not ``cc''.)
+\end{itemize}
 
-\section{Bibliography}
+\subsection{Equations}
+Number equations consecutively. To make your 
+equations more compact, you may use the solidus (~/~), the exp function, or 
+appropriate exponents. Italicize Roman symbols for quantities and variables, 
+but not Greek symbols. Use a long dash rather than a hyphen for a minus 
+sign. Punctuate equations with commas or periods when they are part of a 
+sentence, as in:
+\begin{equation}
+a+b=\gamma\label{eq}
+\end{equation}
+
+Be sure that the 
+symbols in your equation have been defined before or immediately following 
+the equation. Use ``\eqref{eq}'', not ``Eq.~\eqref{eq}'' or ``equation \eqref{eq}'', except at 
+the beginning of a sentence: ``Equation \eqref{eq} is . . .''
+
+\subsection{\LaTeX-Specific Advice}
+
+Please use ``soft'' (e.g., \verb|\eqref{Eq}|) cross references instead
+of ``hard'' references (e.g., \verb|(1)|). That will make it possible
+to combine sections, add equations, or change the order of figures or
+citations without having to go through the file line by line.
+
+Please don't use the \verb|{eqnarray}| equation environment. Use
+\verb|{align}| or \verb|{IEEEeqnarray}| instead. The \verb|{eqnarray}|
+environment leaves unsightly spaces around relation symbols.
+
+Please note that the \verb|{subequations}| environment in {\LaTeX}
+will increment the main equation counter even when there are no
+equation numbers displayed. If you forget that, you might write an
+article in which the equation numbers skip from (17) to (20), causing
+the copy editors to wonder if you've discovered a new method of
+counting.
+
+{\BibTeX} does not work by magic. It doesn't get the bibliographic
+data from thin air but from .bib files. If you use {\BibTeX} to produce a
+bibliography you must send the .bib files.
+
+{\LaTeX} can't read your mind. If you assign the same label to a
+subsubsection and a table, you might find that Table I has been cross
+referenced as Table IV-B3.
+
+{\LaTeX} does not have precognitive abilities. If you put a
+\verb|\label| command before the command that updates the counter it's
+supposed to be using, the label will pick up the last counter to be
+cross referenced instead. In particular, a \verb|\label| command
+should not go before the caption of a figure or a table.
+
+Do not use \verb|\nonumber| inside the \verb|{array}| environment. It
+will not stop equation numbers inside \verb|{array}| (there won't be
+any anyway) and it might stop a wanted equation number in the
+surrounding equation.
+
+\subsection{Some Common Mistakes}\label{SCM}
+\begin{itemize}
+\item The word ``data'' is plural, not singular.
+\item The subscript for the permeability of vacuum $\mu_{0}$, and other common scientific constants, is zero with subscript formatting, not a lowercase letter ``o''.
+\item In American English, commas, semicolons, periods, question and exclamation marks are located within quotation marks only when a complete thought or name is cited, such as a title or full quotation.
+When quotation marks are used, instead of a bold or italic typeface, to highlight a word or phrase, punctuation should appear outside of the quotation marks.
+A parenthetical phrase or statement at the end of a sentence is punctuated outside of the closing parenthesis (like this).
+(A parenthetical sentence is punctuated within the parentheses.)
+\item A graph within a graph is an ``inset'', not an ``insert''. The word alternatively is preferred to the word ``alternately'' (unless you really mean something that alternates).
+\item Do not use the word ``essentially'' to mean ``approximately'' or ``effectively''.
+\item In your paper title, if the words ``that uses'' can accurately replace the word ``using'', capitalize the ``u''; if not, keep using lower-cased.
+\item Be aware of the different meanings of the homophones ``affect'' and ``effect'', ``complement'' and ``compliment'', ``discreet'' and ``discrete'', ``principal'' and ``principle''.
+\item Do not confuse ``imply'' and ``infer''.
+\item The prefix ``non'' is not a word; it should be joined to the word it modifies, usually without a hyphen.
+\item There is no period after the ``et'' in the Latin abbreviation ``et al.''.
+\item The abbreviation ``i.e.'' means ``that is'', and the abbreviation ``e.g.'' means ``for example''.
+\end{itemize}
+An excellent style manual for science writers is \cite{b7}.
+
+\subsection{Authors and Affiliations}
+\textbf{The class file is designed for, but not limited to, six authors.} A 
+minimum of one author is required for all conference articles. Author names 
+should be listed starting from left to right and then moving down to the 
+next line. This is the author sequence that will be used in future citations 
+and by indexing services. Names should not be listed in columns nor group by 
+affiliation. Please keep your affiliations as succinct as possible (for 
+example, do not differentiate among departments of the same organization).
+
+\subsection{Identify the Headings}
+Headings, or heads, are organizational devices that guide the reader through 
+your paper. There are two types: component heads and text heads.
+
+Component heads identify the different components of your paper and are not 
+topically subordinate to each other. Examples include Acknowledgments and 
+References and, for these, the correct style to use is ``Heading 5''. Use 
+``figure caption'' for your Figure captions, and ``table head'' for your 
+table title. Run-in heads, such as ``Abstract'', will require you to apply a 
+style (in this case, italic) in addition to the style provided by the drop 
+down menu to differentiate the head from the text.
+
+Text heads organize the topics on a relational, hierarchical basis. For 
+example, the paper title is the primary text head because all subsequent 
+material relates and elaborates on this one topic. If there are two or more 
+sub-topics, the next level head (uppercase Roman numerals) should be used 
+and, conversely, if there are not at least two sub-topics, then no subheads 
+should be introduced.
+
+\subsection{Figures and Tables}
+\paragraph{Positioning Figures and Tables} Place figures and tables at the top and 
+bottom of columns. Avoid placing them in the middle of columns. Large 
+figures and tables may span across both columns. Figure captions should be 
+below the figures; table heads should appear above the tables. Insert 
+figures and tables after they are cited in the text. Use the abbreviation 
+``Fig.~\ref{fig}'', even at the beginning of a sentence.
+
+\begin{table}[htbp]
+\caption{Table Type Styles}
+\begin{center}
+\begin{tabular}{|c|c|c|c|}
+\hline
+\textbf{Table}&\multicolumn{3}{|c|}{\textbf{Table Column Head}} \\
+\cline{2-4} 
+\textbf{Head} & \textbf{\textit{Table column subhead}}& \textbf{\textit{Subhead}}& \textbf{\textit{Subhead}} \\
+\hline
+copy& More table copy$^{\mathrm{a}}$& &  \\
+\hline
+\multicolumn{4}{l}{$^{\mathrm{a}}$Sample of a Table footnote.}
+\end{tabular}
+\label{tab1}
+\end{center}
+\end{table}
+
+Figure Labels: Use 8 point Times New Roman for Figure labels. Use words 
+rather than symbols or abbreviations when writing Figure axis labels to 
+avoid confusing the reader. As an example, write the quantity 
+``Magnetization'', or ``Magnetization, M'', not just ``M''. If including 
+units in the label, present them within parentheses. Do not label axes only 
+with units. In the example, write ``Magnetization (A/m)'' or ``Magnetization 
+\{A[m(1)]\}'', not just ``A/m''. Do not label axes with a ratio of 
+quantities and units. For example, write ``Temperature (K)'', not 
+``Temperature/K''.
+
+\section*{Acknowledgment}
+
+The preferred spelling of the word ``acknowledgment'' in America is without 
+an ``e'' after the ``g''. Avoid the stilted expression ``one of us (R. B. 
+G.) thanks $\ldots$''. Instead, try ``R. B. G. thanks$\ldots$''. Put sponsor 
+acknowledgments in the unnumbered footnote on the first page.
+
+\section*{References}
+
+Please number citations consecutively within brackets \cite{b1}. The 
+sentence punctuation follows the bracket \cite{b2}. Refer simply to the reference 
+number, as in \cite{b3}---do not use ``Ref. \cite{b3}'' or ``reference \cite{b3}'' except at 
+the beginning of a sentence: ``Reference \cite{b3} was the first $\ldots$''
+
+Number footnotes separately in superscripts. Place the actual footnote at 
+the bottom of the column in which it was cited. Do not put footnotes in the 
+abstract or reference list. Use letters for table footnotes.
+
+Unless there are six authors or more give all authors' names; do not use 
+``et al.''. Papers that have not been published, even if they have been 
+submitted for publication, should be cited as ``unpublished'' \cite{b4}. Papers 
+that have been accepted for publication should be cited as ``in press'' \cite{b5}. 
+Capitalize only the first word in a paper title, except for proper nouns and 
+element symbols.
+
+For papers published in translation journals, please give the English 
+citation first, followed by the original foreign-language citation \cite{b6}.
+
+\begin{thebibliography}{00}
+\bibitem{b1} G.
+Eason, B.
+Noble, and I.
+N.
+Sneddon, ``On certain integrals of Lipschitz-Hankel type involving products of Bessel functions,'' Phil.
+Trans.
+Roy.
+Soc.
+London, vol.
+A247, pp.
+529--551, April 1955.
+\bibitem{b2} J. Clerk Maxwell, A Treatise on Electricity and Magnetism, 3rd ed., vol. 2. Oxford: Clarendon, 1892, pp.68--73.
+\bibitem{b3} I. S. Jacobs and C. P. Bean, ``Fine particles, thin films and exchange anisotropy,'' in Magnetism, vol. III, G. T. Rado and H. Suhl, Eds. New York: Academic, 1963, pp. 271--350.
+\bibitem{b4} K. Elissa, ``Title of paper if known,'' unpublished.
+\bibitem{b5} R. Nicole, ``Title of paper with only first word capitalized,'' J. Name Stand. Abbrev., in press.
+\bibitem{b6} Y.
+Yorozu, M.
+Hirano, K.
+Oka, and Y.
+Tagawa, ``Electron spectroscopy studies on magneto-optical media and plastic substrate interface,'' IEEE Transl.
+J.
+Magn.
+Japan, vol.
+2, pp.
+740--741, August 1987 [Digests 9th Annual Conf.
+Magnetics Japan, p.
+301, 1982].
+\bibitem{b7} M. Young, The Technical Writer's Handbook. Mill Valley, CA: University Science, 1989.
+\end{thebibliography}
+\vspace{12pt}
+\color{red}
+IEEE conference templates contain guidance text for composing and formatting conference papers.
+Please ensure that all template text is removed from your conference paper prior to submission to the conference.
+Failure to remove the template text from your paper may result in your paper not being published.
 

### 2023-08-21 doc/paper.tex
Update paper
Signed-off-by: Jonatan Hanssen <jonatan.hanssen@outlook.com>

@@
 
-\title{Conference Paper Title*\\
-{\footnotesize \textsuperscript{*}Note: Sub-titles are not captured in Xplore and
-should not be used}
-\thanks{Identify applicable funding agency here. If none, delete this.}
-}
+\title{IN5490 Project}
 
-\author{\IEEEauthorblockN{1\textsuperscript{st} Given Name Surname}
-\IEEEauthorblockA{\textit{dept. name of organization (of Aff.)} \\
-\textit{name of organization (of Aff.)}\\
-City, Country \\
-email address or ORCID}
+\author{\IEEEauthorblockN{Adrian Duric}
+\IEEEauthorblockA{\textit{Master Student, Dept. Informatics} \\
+\textit{The faculty of Mathematics}\\
+Oslo, Norway \\
+adriandu@ifi.uio.no}
 \and
-\IEEEauthorblockN{2\textsuperscript{nd} Given Name Surname}
-\IEEEauthorblockA{\textit{dept. name of organization (of Aff.)} \\
-\textit{name of organization (of Aff.)}\\
-City, Country \\
-email address or ORCID}
+\IEEEauthorblockN{Gregor Kajda}
+\IEEEauthorblockA{\textit{Master Student, Dept. of Informatics} \\
+\textit{The faculty of Mathematics}\\
+Oslo, Norway \\
+grzegork@ifi.uio.no}
 \and
-\IEEEauthorblockN{3\textsuperscript{rd} Given Name Surname}
-\IEEEauthorblockA{\textit{dept. name of organization (of Aff.)} \\
-\textit{name of organization (of Aff.)}\\
-City, Country \\
-email address or ORCID}
-\and
-\IEEEauthorblockN{4\textsuperscript{th} Given Name Surname}
-\IEEEauthorblockA{\textit{dept. name of organization (of Aff.)} \\
-\textit{name of organization (of Aff.)}\\
-City, Country \\
-email address or ORCID}
-\and
-\IEEEauthorblockN{5\textsuperscript{th} Given Name Surname}
-\IEEEauthorblockA{\textit{dept. name of organization (of Aff.)} \\
-\textit{name of organization (of Aff.)}\\
-City, Country \\
-email address or ORCID}
-\and
-\IEEEauthorblockN{6\textsuperscript{th} Given Name Surname}
-\IEEEauthorblockA{\textit{dept. name of organization (of Aff.)} \\
-\textit{name of organization (of Aff.)}\\
-City, Country \\
-email address or ORCID}
+\IEEEauthorblockN{Jonatan Hoffmann Hanssen}
+\IEEEauthorblockA{\textit{Master Student, Dept. of Informatics} \\
+\textit{The faculty of Mathematics}\\
+Oslo, Norway \\
+jonatahh@ifi.uio.no}
 }
@@
 \begin{abstract}
-This document is a model and instructions for \LaTeX.
-This and the IEEEtran.cls file define the components of your paper [title, text, heads, etc.]. *CRITICAL: Do Not Use Symbols, Special Characters, Footnotes, 
-or Math in Paper Title or Abstract.
+
+    This is gonna be so epic xD
+
 \end{abstract}
@@
 \begin{IEEEkeywords}
-component, formatting, style, styling, insert
+epic, lol, reddit
 \end{IEEEkeywords}
@@
 \section{Introduction}
-This document is a model and instructions for \LaTeX.
-Please observe the conference page limits.
 
-\section{Ease of Use}
-
-\subsection{Maintaining the Integrity of the Specifications}
-
-The IEEEtran class file is used to format your paper and style the text. All margins, 
-column widths, line spaces, and text fonts are prescribed; please do not 
-alter them. You may note peculiarities. For example, the head margin
-measures proportionately more than is customary. This measurement 
-and others are deliberate, using specifications that anticipate your paper 
-as one part of the entire proceedings, and not as an independent document. 
-Please do not revise any of the current designations.
-
-\section{Prepare Your Paper Before Styling}
-Before you begin to format your paper, first write and save the content as a 
-separate text file. Complete all content and organizational editing before 
-formatting. Please note sections \ref{AA}--\ref{SCM} below for more information on 
-proofreading, spelling and grammar.
-
-Keep your text and graphic files separate until after the text has been 
-formatted and styled. Do not number text heads---{\LaTeX} will do that 
-for you.
-
-\subsection{Abbreviations and Acronyms}\label{AA}
-Define abbreviations and acronyms the first time they are used in the text, 
-even after they have been defined in the abstract. Abbreviations such as 
-IEEE, SI, MKS, CGS, ac, dc, and rms do not have to be defined. Do not use 
-abbreviations in the title or heads unless they are unavoidable.
-
-\subsection{Units}
-\begin{itemize}
-\item Use either SI (MKS) or CGS as primary units.
-(SI units are encouraged.) English units may be used as secondary units (in parentheses).
-An exception would be the use of English units as identifiers in trade, such as ``3.5-inch disk drive''.
-\item Avoid combining SI and CGS units, such as current in amperes and magnetic field in oersteds.
-This often leads to confusion because equations do not balance dimensionally.
-If you must use mixed units, clearly state the units for each quantity that you use in an equation.
-\item Do not mix complete spellings and abbreviations of units: ``Wb/m\textsuperscript{2}'' or ``webers per square meter'', not ``webers/m\textsuperscript{2}''.
-Spell out units when they appear in text: ``.
-.
-.
-a few henries'', not ``.
-.
-.
-a few H''.
-\item Use a zero before decimal points: ``0.25'', not ``.25''. Use ``cm\textsuperscript{3}'', not ``cc''.)
-\end{itemize}
-
-\subsection{Equations}
-Number equations consecutively. To make your 
-equations more compact, you may use the solidus (~/~), the exp function, or 
-appropriate exponents. Italicize Roman symbols for quantities and variables, 
-but not Greek symbols. Use a long dash rather than a hyphen for a minus 
-sign. Punctuate equations with commas or periods when they are part of a 
-sentence, as in:
-\begin{equation}
-a+b=\gamma\label{eq}
-\end{equation}
-
-Be sure that the 
-symbols in your equation have been defined before or immediately following 
-the equation. Use ``\eqref{eq}'', not ``Eq.~\eqref{eq}'' or ``equation \eqref{eq}'', except at 
-the beginning of a sentence: ``Equation \eqref{eq} is . . .''
+We do a little bit of trolling
 

### 2023-09-14 doc/paper.tex
Remove placeholder text
Signed-off-by: Jonatan Hanssen <jonatan.hanssen@outlook.com>
Update paper
Signed-off-by: Jonatan Hanssen <jonatan.hanssen@outlook.com>

@@
 
-    This is gonna be so epic xD
-
 \end{abstract}
@@
 \begin{IEEEkeywords}
-epic, lol, reddit
 \end{IEEEkeywords}
@@
 
-We do a little bit of trolling
+One of the central challenges of reinforcement learning is that rewards are often both extremely rare and delayed, which means that optimizing a reinforcement learning algorithm requires significant trial and error \cite[423]{brunton}.
 
@@
 
-\begin{thebibliography}{00}
-\bibitem{b1} G.
-Eason, B.
-Noble, and I.
-N.
-Sneddon, ``On certain integrals of Lipschitz-Hankel type involving products of Bessel functions,'' Phil.
-Trans.
-Roy.
-Soc.
-London, vol.
-A247, pp.
-529--551, April 1955.
-\bibitem{b2} J. Clerk Maxwell, A Treatise on Electricity and Magnetism, 3rd ed., vol. 2. Oxford: Clarendon, 1892, pp.68--73.
-\bibitem{b3} I. S. Jacobs and C. P. Bean, ``Fine particles, thin films and exchange anisotropy,'' in Magnetism, vol. III, G. T. Rado and H. Suhl, Eds. New York: Academic, 1963, pp. 271--350.
-\bibitem{b4} K. Elissa, ``Title of paper if known,'' unpublished.
-\bibitem{b5} R. Nicole, ``Title of paper with only first word capitalized,'' J. Name Stand. Abbrev., in press.
-\bibitem{b6} Y.
-Yorozu, M.
-Hirano, K.
-Oka, and Y.
-Tagawa, ``Electron spectroscopy studies on magneto-optical media and plastic substrate interface,'' IEEE Transl.
-J.
-Magn.
-Japan, vol.
-2, pp.
-740--741, August 1987 [Digests 9th Annual Conf.
-Magnetics Japan, p.
-301, 1982].
-\bibitem{b7} M. Young, The Technical Writer's Handbook. Mill Valley, CA: University Science, 1989.
-\end{thebibliography}
-\vspace{12pt}
-\color{red}
-IEEE conference templates contain guidance text for composing and formatting conference papers.
-Please ensure that all template text is removed from your conference paper prior to submission to the conference.
-Failure to remove the template text from your paper may result in your paper not being published.
+\bibliographystyle{apalike}
+\bibliography{bibliography}
 

### 2023-09-15 src/test.py
Update prompting code
Signed-off-by: Jonatan Hanssen <jonatan.hanssen@outlook.com>

@@ -12,15 +12,16 @@ env = gym.make("MiniGrid-UnlockPickup-v0", render_mode="human")
 # observation, info = env.reset(seed=1)
 observation, info = env.reset()
 
-policy = llama2_70b_policy(rl_temp=0.1)
+policy = llama2_70b_policy(rl_temp=0.0)
 
 action_list = list()
 
 for _ in range(1000):
     if not action_list:
-        prompt = obs_to_string(observation["image"])
-        print(prompt)
-        policy(prompt, action_list, env)
+        obs_string = obs_to_string(observation["image"])
+        print(obs_string)
+        policy(obs_string, action_list, env)
+        input()
 
     print([constants.IDX_TO_ACTION[idx] for idx in action_list])
     action = action_list.pop(0)

### 2023-09-15 src/utils.py

@@ -20,8 +20,14 @@ class llama2_70b_policy:
         self.temperature = temperature
         self.top_p = top_p
         self.rl_temp = rl_temp
+        self.dialog = [
+            {
+                "role": "system",
+                "content": "You are a player playing a videogame. It is a top down turn based game, where each turn you can either move RIGHT, LEFT, FORWARD, PICK UP or DROP objects, or TOGGLE objects in front of you.",
+            }
+        ]
 
-    def __call__(self, prompt, action_list, env):
+    def __call__(self, observation, action_list, env):
         if "nothing" in prompt:
             action_list.append(env.action_space.sample())
             return
@@ -33,6 +39,8 @@ class llama2_70b_policy:
             print("here")
             action_list.append(env.action_space.sample())
             return
+
+        self.dialog.append({"role": "user"
 
         dialogs = [
             [
@@ -135,8 +143,9 @@ def obs_to_string(obs_matrix):
 
     for ind in wall_inds:
         if ind[0] - 3 == 0 and ind[1] - 6 == -1:
-            observation_strings.append(ind_to_string(ind, "wall"))
+            observation_strings.append(ind_to_string(ind, "impassable wall"))
 
-    prompt = f"You are a player playing a videogame. It is a top down turn based game, where each turn you can either move RIGHT, LEFT, FORWARD or PICK UP objects 1 square in front of you. You see {', '.join(observation_strings) if observation_strings else 'nothing'}. What should you do? Please only answer with a single of the following commands: RIGHT, LEFT, FORWARD or PICK UP."
+    prompt = f"You see {', '.join(observation_strings) if observation_strings else 'nothing'}. What should you do?"
+    # Please only answer with a single of the following commands: RIGHT, LEFT, FORWARD or PICK UP."
 
     return prompt

### 2023-09-15 src/chatdemo.py
Try to add memory to chatbot
Signed-off-by: Jonatan Hanssen <jonatan.hanssen@outlook.com>

@@ -13,23 +13,26 @@ generator = Llama.build(
     max_batch_size=6,
 )
 
+dialog = [
+    {
+        "role": "system",
+        "content": "You are a player playing a videogame. It is a top down turn based game, where each turn you can either move RIGHT, LEFT, FORWARD, PICK UP or DROP objects, or TOGGLE objects in front of you.",
+    }
+]
+
+
 while True:
     inp = input("USER: ")
 
-    dialogs = [
-        [
-            {
-                "role": "user",
-                "content": inp,
-            }
-        ],
-    ]
+    dialog.append({"role": "user", "content": inp})
 
     results = generator.chat_completion(
-        dialogs,  # type: ignore
+        [dialog],  # type: ignore
         max_gen_len=15,
         temperature=0.6,
         top_p=0.9,
     )
     output = results[0]["generation"]["content"]
     print(output)
+
+    dialog.append({"role": "assistant", "content": output})

### 2023-10-02 doc/paper.tex
update paper
Signed-off-by: Jonatan Hanssen <jonatan.hanssen@outlook.com>

@@
 
+Reinforcement Learning!
+
 \end{abstract}
@@
 One of the central challenges of reinforcement learning is that rewards are often both extremely rare and delayed, which means that optimizing a reinforcement learning algorithm requires significant trial and error \cite[423]{brunton}.
+Furthermore, in many problems the state-action spaces are enormous, meaning that uniform exploration is unlikely to find good solutions in a reasonable amount of time.
+Many methods have been developed to deal with this issue, for example by introducing auxilliary reward functions that use domain knowledge, or which reward the agent for learning novel skills.
+However, hand picking which actions to reward can lead to imitation rather than optimal behaviour, and novelty is not always useful \cite[2]{ellm}.
+With the introduction of Large Language Models (LLMs), a new area of research has opened up, attempting to use the vast amount of human knowledge encoded in these models to increase the performance of reinforcement learning algorithms \cite{survey}.
+These
 
-\subsection{\LaTeX-Specific Advice}
+\section{Background and Related Work}
 
-Please use ``soft'' (e.g., \verb|\eqref{Eq}|) cross references instead
-of ``hard'' references (e.g., \verb|(1)|). That will make it possible
-to combine sections, add equations, or change the order of figures or
-citations without having to go through the file line by line.
+ Degus is the future in blockchain techonology, created by using the revolutionary "degus system", which creates twice as many coins per unit of computational power.
+By accepting the "degus mindset", you will be able to acheive things you never ever thought possible.
+The degus mindset includes the following:
 
-Please don't use the \verb|{eqnarray}| equation environment. Use
-\verb|{align}| or \verb|{IEEEeqnarray}| instead. The \verb|{eqnarray}|
-environment leaves unsightly spaces around relation symbols.
+Always Striving for Excellence
+Generate passive revenue through the degus system
+Live day by day: "Carpe diem"
 
-Please note that the \verb|{subequations}| environment in {\LaTeX}
-will increment the main equation counter even when there are no
-equation numbers displayed. If you forget that, you might write an
-article in which the equation numbers skip from (17) to (20), causing
-the copy editors to wonder if you've discovered a new method of
-counting.
+By following these simple steps, and by buying several GPUs, you will be able to produce around 15 deguscoins per computational unit, unleashing the true power of the blockchain.
 
-{\BibTeX} does not work by magic. It doesn't get the bibliographic
-data from thin air but from .bib files. If you use {\BibTeX} to produce a
-bibliography you must send the .bib files.
+What is the deguscoin, you may ask?
+The deguscoin represents a complete transformation of the traditional financial system (tradfi).
+By using deguscoin, you will be able to achieve returns of investment (ROIs) beyond mere human comprehension.
+One single deguscoin, when inserted into an economy, can have such destructive power that even mere days after, the market will be beyond recognition.
+By invoking the "degus mindset" (DM) you will be able to send around 15\% more currency per coin, achieving ROIs which would send the average economist into a seizure.
 
-{\LaTeX} can't read your mind. If you assign the same label to a
-subsubsection and a table, you might find that Table I has been cross
-referenced as Table IV-B3.
+Market players are predicting that by the end of this year, around 50\% of all transactions will be done under the degus system (DS), not even counting derivatives.
 
-{\LaTeX} does not have precognitive abilities. If you put a
-\verb|\label| command before the command that updates the counter it's
-supposed to be using, the label will pick up the last counter to be
-cross referenced instead. In particular, a \verb|\label| command
-should not go before the caption of a figure or a table.
+By the end of this year, around 50\% of all transactions will be done under the degus system, not even counting derivatives.
 
-Do not use \verb|\nonumber| inside the \verb|{array}| environment. It
-will not stop equation numbers inside \verb|{array}| (there won't be
-any anyway) and it might stop a wanted equation number in the
-surrounding equation.
+– Market players
 
-\subsection{Some Common Mistakes}\label{SCM}
-\begin{itemize}
-\item The word ``data'' is plural, not singular.
-\item The subscript for the permeability of vacuum $\mu_{0}$, and other common scientific constants, is zero with subscript formatting, not a lowercase letter ``o''.
-\item In American English, commas, semicolons, periods, question and exclamation marks are located within quotation marks only when a complete thought or name is cited, such as a title or full quotation.
-When quotation marks are used, instead of a bold or italic typeface, to highlight a word or phrase, punctuation should appear outside of the quotation marks.
-A parenthetical phrase or statement at the end of a sentence is punctuated outside of the closing parenthesis (like this).
-(A parenthetical sentence is punctuated within the parentheses.)
-\item A graph within a graph is an ``inset'', not an ``insert''. The word alternatively is preferred to the word ``alternately'' (unless you really mean something that alternates).
-\item Do not use the word ``essentially'' to mean ``approximately'' or ``effectively''.
-\item In your paper title, if the words ``that uses'' can accurately replace the word ``using'', capitalize the ``u''; if not, keep using lower-cased.
-\item Be aware of the different meanings of the homophones ``affect'' and ``effect'', ``complement'' and ``compliment'', ``discreet'' and ``discrete'', ``principal'' and ``principle''.
-\item Do not confuse ``imply'' and ``infer''.
-\item The prefix ``non'' is not a word; it should be joined to the word it modifies, usually without a hyphen.
-\item There is no period after the ``et'' in the Latin abbreviation ``et al.''.
-\item The abbreviation ``i.e.'' means ``that is'', and the abbreviation ``e.g.'' means ``for example''.
-\end{itemize}
-An excellent style manual for science writers is \cite{b7}.
+What are you waiting for? Purchase deguscoin right now, and leave your earthly desires behind!
 
-\subsection{Authors and Affiliations}
-\textbf{The class file is designed for, but not limited to, six authors.} A 
-minimum of one author is required for all conference articles. Author names 
-should be listed starting from left to right and then moving down to the 
-next line. This is the author sequence that will be used in future citations 
-and by indexing services. Names should not be listed in columns nor group by 
-affiliation. Please keep your affiliations as succinct as possible (for 
-example, do not differentiate among departments of the same organization).
+\section{Method}
 
-\subsection{Identify the Headings}
-Headings, or heads, are organizational devices that guide the reader through 
-your paper. There are two types: component heads and text heads.
+Deguser trenger god plass for å trives, og du bør investere i et størst mulig bur som gir dyrene mulighet til å klatre, grave i et tykt strølag og bevege seg i flere etasjer.
 
-Component heads identify the different components of your paper and are not 
-topically subordinate to each other. Examples include Acknowledgments and 
-References and, for these, the correct style to use is ``Heading 5''. Use 
-``figure caption'' for your Figure captions, and ``table head'' for your 
-table title. Run-in heads, such as ``Abstract'', will require you to apply a 
-style (in this case, italic) in addition to the style provided by the drop 
-down menu to differentiate the head from the text.
+Mange dyreeiere konstruerer sine egne løsninger, for eksempel med et stort, gammelt akvarium som underdel og et nettingbur på toppen av dette.
+Da får du full oversikt over hva som skjer også nederst i buret, samtidig som du slipper at sand, strø og høy søles ut i rommet.
+Ikke la degusene gå direkte på gitterbunn eller annet hardt underlag, det kan skade føttene.
 
-Text heads organize the topics on a relational, hierarchical basis. For 
-example, the paper title is the primary text head because all subsequent 
-material relates and elaborates on this one topic. If there are two or more 
-sub-topics, the next level head (uppercase Roman numerals) should be used 
-and, conversely, if there are not at least two sub-topics, then no subheads 
-should be introduced.
+Buret kan du innrede med f.eks.
+røtter og greiner til å klatre på, hus av treverk, rør til å krype gjennom og løpehjul.
+Et løpehjul til degus bør være minst 25 cm i diameter og ha tett gulv og vegger slik at halen ikke kan komme i klem.
+Bunnmaterialet i buret bør være et tykt lag av støvfritt smådyrstrø som er egnet for graving.
+Det finnes flere egnede strøtyper å få kjøpt.
+Bland gjerne inn litt tørr blomsterjord eller sand i bunnmaterialet.
+Hvis du henter sand ute, bør du desinfisere den ved frysing eller steking fr den brukes i buret.
+Degusene trenger også tilgang på fiberrikt materiale som høy, revet papir eller treull til bygge- og redemateriale.
 
-\subsection{Figures and Tables}
-\paragraph{Positioning Figures and Tables} Place figures and tables at the top and 
-bottom of columns. Avoid placing them in the middle of columns. Large 
-figures and tables may span across both columns. Figure captions should be 
-below the figures; table heads should appear above the tables. Insert 
-figures and tables after they are cited in the text. Use the abbreviation 
-``Fig.~\ref{fig}'', even at the beginning of a sentence.
+For å være sikker på at alle degusene som bor sammen kan trekke seg unna og hvile, bør de få hvert sitt sovehus.
+Deguser kan også finne på å forsvare maten sin overfor andre deguser.
+Når flere deguser bor i samme bur, bør du derfor ha flere matskåler plassert på forskjellige steder i buret slik at alle får tilgang til mat når de ønsker det.
+Dyrene må alltid ha tilgang på friskt vann, som best gis i en drikkeflaske.
 
-\begin{table}[htbp]
-\caption{Table Type Styles}
-\begin{center}
-\begin{tabular}{|c|c|c|c|}
-\hline
-\textbf{Table}&\multicolumn{3}{|c|}{\textbf{Table Column Head}} \\
-\cline{2-4} 
-\textbf{Head} & \textbf{\textit{Table column subhead}}& \textbf{\textit{Subhead}}& \textbf{\textit{Subhead}} \\
-\hline
-copy& More table copy$^{\mathrm{a}}$& &  \\
-\hline
-\multicolumn{4}{l}{$^{\mathrm{a}}$Sample of a Table footnote.}
-\end{tabular}
-\label{tab1}
-\end{center}
-\end{table}
+For å holde pelsen i orden har deguser behov for å bade i finkornet sand flere ganger i uka.
+Slik badesand får du kjøpt i dyrebutikken.
+Et tungt kar som ikke velter så lett er fint som sandbadekar.
+Det kan være lurt å fjerne sandbadet mellom hvert bad for å unngå at den brukes som toalett.
 
-Figure Labels: Use 8 point Times New Roman for Figure labels. Use words 
-rather than symbols or abbreviations when writing Figure axis labels to 
-avoid confusing the reader. As an example, write the quantity 
-``Magnetization'', or ``Magnetization, M'', not just ``M''. If including 
-units in the label, present them within parentheses. Do not label axes only 
-with units. In the example, write ``Magnetization (A/m)'' or ``Magnetization 
-\{A[m(1)]\}'', not just ``A/m''. Do not label axes with a ratio of 
-quantities and units. For example, write ``Temperature (K)'', not 
-``Temperature/K''.
+Rengjøring
+
+Hvor ofte buret bør rengjøres avhenger av størrelsen og hvor mange som bor der. Som en tommelfingerregel bør du skifte bunnmaterialet og vaske bunn og innredning én gang i uka.
+
+\section{Experiments}
+
+\section{Results}
+
+\section{Conclusion}
 
@@
 
-The preferred spelling of the word ``acknowledgment'' in America is without 
-an ``e'' after the ``g''. Avoid the stilted expression ``one of us (R. B. 
-G.) thanks $\ldots$''. Instead, try ``R. B. G. thanks$\ldots$''. Put sponsor 
-acknowledgments in the unnumbered footnote on the first page.
+I would like to acknowledge myself for being a absolute legend.
 
 \section*{References}
-
-Please number citations consecutively within brackets \cite{b1}. The 
-sentence punctuation follows the bracket \cite{b2}. Refer simply to the reference 
-number, as in \cite{b3}---do not use ``Ref. \cite{b3}'' or ``reference \cite{b3}'' except at 
-the beginning of a sentence: ``Reference \cite{b3} was the first $\ldots$''
-
-Number footnotes separately in superscripts. Place the actual footnote at 
-the bottom of the column in which it was cited. Do not put footnotes in the 
-abstract or reference list. Use letters for table footnotes.
-
-Unless there are six authors or more give all authors' names; do not use 
-``et al.''. Papers that have not been published, even if they have been 
-submitted for publication, should be cited as ``unpublished'' \cite{b4}. Papers 
-that have been accepted for publication should be cited as ``in press'' \cite{b5}. 
-Capitalize only the first word in a paper title, except for proper nouns and 
-element symbols.
-
-For papers published in translation journals, please give the English 
-citation first, followed by the original foreign-language citation \cite{b6}.
 

### 2023-10-02 src/SB3_PPO_MiniGrid/feature_extractor.py
First version of PPO using stable_baselines3 implementation. TODO: test params, explore what can be done when implemenation from scratch usin PyTorch is goind to be coded

@@ -1,0 +1,63 @@
+import torch
+import torch.nn as nn
+from gymnasium import spaces
+import gymnasium as gym
+
+from stable_baselines3 import PPO
+from stable_baselines3.common.torch_layers import BaseFeaturesExtractor
+from stable_baselines3.common.env_util import make_vec_env
+from minigrid.wrappers import ImgObsWrapper
+from stable_baselines3.common.vec_env import DummyVecEnv
+
+class CustomCNN(BaseFeaturesExtractor):
+    """
+    :param observation_space: (gym.Space)
+    :param features_dim: (int) Number of features extracted.
+        This corresponds to torche number of unit for torche last layer.
+    """
+
+    def __init__(self, observation_space: spaces.Box, features_dim: int = 128):
+        super().__init__(observation_space, features_dim)
+        # We assume CxHxW images (channels first)
+        # Re-ordering will be done by pre-preprocessing or wrapper
+        n_input_channels = observation_space.shape[0]
+        self.cnn = nn.Sequential(
+            nn.Conv2d(n_input_channels, 16, (2,2)),
+            nn.Tanh(),
+            nn.Conv2d(16, 32, (2,2)),
+            nn.Tanh(),
+            # nn.Conv2d(32, 64, (2,2)),
+            # nn.ReLU(),
+            nn.Flatten(),
+        )
+
+        # Compute shape by doing one forward pass
+        with torch.no_grad():
+            n_flatten = self.cnn(
+                torch.as_tensor(observation_space.sample()[None]).float()
+            ).shape[1]
+
+        self.linear = nn.Sequential(nn.Linear(n_flatten, features_dim), nn.Tanh())
+
+    def forward(self, observations: torch.Tensor) -> torch.Tensor:
+        return self.linear(self.cnn(observations))
+
+
+policy_kwargs = dict(
+    features_extractor_class=CustomCNN,
+    features_extractor_kwargs=dict(features_dim=128),
+)
+# env = make_vec_env("MiniGrid-UnlockPickup-v0", n_envs=1)
+env = DummyVecEnv([lambda: ImgObsWrapper(gym.make("MiniGrid-UnlockPickup-v0", render_mode="rgb_array"))])
+env.seed(42)
+# env = ImgObsWrapper(env)
+
+model = PPO(
+    "CnnPolicy", env, policy_kwargs=policy_kwargs, verbose=1
+)
+model.learn(total_timesteps=2e4)
+model.save("ppo_minigrid")
+del model
+
+
+

### 2023-10-02 doc/paper.tex
Update paper with intro and abstract
Signed-off-by: Jonatan Hanssen <jonatan.hanssen@outlook.com>

@@
 
-\title{IN5490 Project}
+\title{Large Language Models in Reinforcement Learning - A Comparison of Methods}
 
@@
 
-Reinforcement Learning!
-
+Reinforcement Learning (RL) algorithms suffer when rewards are sparce and the state-action space is large.
+Even tasks which appear relatively simple can prove intractable if the completion of the task requires subtasks to be completed first, or if the reward only comes when the entire task has been completed.
+In such cases, random exploration is unlikely to lead the agent to discover a solution to the problem.
+The apparent simplicity of such problems is often due to human intuition, which allow us to quickly see possible solutions to a diverse set of problems.
+Large Language Models (LLMs) are trained on large corpura of human written text, and have been shown to capture parts of this intuition in many tasks \cite{sparks}.
+In recent years, LLMs have been succesfully used to aid RL agents in more efficient exploration, and have allowed them to solve problems which previous methods have been unable to \cite{omni} \cite{ellm}.
+We compare different methods for integrating an LLM into the deep learning reinforcement algorithm proximal policy optimization (PPO), and compare their efficiency against eachother.
+We find that BEST METHOD gives the best sample efficiency, outperforming normal PPO by PERCENTAGE in METRIC.
 \end{abstract}
@@
 \begin{IEEEkeywords}
+large language models, reinforcement learning, minigrid, proximal policy optimization
 \end{IEEEkeywords}
@@
 Furthermore, in many problems the state-action spaces are enormous, meaning that uniform exploration is unlikely to find good solutions in a reasonable amount of time.
-Many methods have been developed to deal with this issue, for example by introducing auxilliary reward functions that use domain knowledge, or which reward the agent for learning novel skills.
-However, hand picking which actions to reward can lead to imitation rather than optimal behaviour, and novelty is not always useful \cite[2]{ellm}.
-With the introduction of Large Language Models (LLMs), a new area of research has opened up, attempting to use the vast amount of human knowledge encoded in these models to increase the performance of reinforcement learning algorithms \cite{survey}.
-These
+Many methods have been developed to deal with this issue, for example by introducing auxilliary reward functions that use domain knowledge to reward actions which are considered good, or which reward the agent for learning novel skills.
+However, hand picking which actions to reward can lead to imitation rather than optimal behaviour, and novelty is not always useful \cite[1]{ellm}.
+In recent years, LLMs have shown remarkable capabilities in problem solving and planning \cite{sparks}, qualities which traditional RL agents often lack.
+Thus, a new area of research has emerged, which attemps to use the vast amount of human knowledge encoded in these models to increase the performance of reinforcement learning algorithms \cite{survey}.
+These LLM assisted RL agents have been able to outperform state-of-the-art RL methods in many problems \cite{omni} \cite{ellm} \cite{idm}.
+\\
+
+These papers explore different methods of integrating LLMs into a standard reinforcement learning training loop, from altering the policy directly to introducing an auxilliary reward function.
+In this paper, we compare both methods in the same environment using the same deep reinforcement algorithm (PPO), and compare their results to eachother to explore how LLMs best can be integrated into RL.
 

### 2023-10-02 doc/paper.tex
added to background and related work in paper

--- 2023-10-02 doc/paper.tex
@@
 \author{\IEEEauthorblockN{Adrian Duric}
-\IEEEauthorblockA{\textit{Master Student, Dept. Informatics} \\
-\textit{The faculty of Mathematics}\\
+\IEEEauthorblockA{\textit{Master Student, Dept. of Informatics} \\
+\textit{The Faculty of Mathematics}\\
 Oslo, Norway \\
@@
 \IEEEauthorblockA{\textit{Master Student, Dept. of Informatics} \\
-\textit{The faculty of Mathematics}\\
+\textit{The Faculty of Mathematics}\\
 Oslo, Norway \\
@@
 
-Reinforcement Learning!
+Reinforcement Learning! And LLMs!
 
@@
 
- Degus is the future in blockchain techonology, created by using the revolutionary "degus system", which creates twice as many coins per unit of computational power.
-By accepting the "degus mindset", you will be able to acheive things you never ever thought possible.
-The degus mindset includes the following:
+\subsection{Large Language Models as Policy in Reinforcement Learning}
 
-Always Striving for Excellence
-Generate passive revenue through the degus system
-Live day by day: "Carpe diem"
+In recent research into integrating LLMs into the RL framework, many of the methods proposed include ways of making the LLM influence the policy of the RL agent.
+\cite{omni} considers open-ended learning algorithms in which the value of the agent attempting certain tasks can be evaluated by estimating the the interestingness of the task.
+The notion of interestingness is then thought of as what task would intuitively be interesting for a human to try in the context of learning in an environment.
+Some main factors contributing to this interestingness are how likely one is to succeed in doing the task, as well as whether learning the given task may increase the likelihood of succeeding in other tasks.
+Considering that LLMs are trained on extremely large amounts of human-written text containing human knowledge and intuition, this leads to the idea that if the LLM was told to propose interesting actions for the RL agent to learn in its environment, it could be able to communicate human intuition directly to the agent.
+This would in turn let the agent learn with increased sample efficiency as the human intuition-based input it receives would make it decide to perform intuitively smarter actions, particularly in the early stages of exploring its environment.
 
-By following these simple steps, and by buying several GPUs, you will be able to produce around 15 deguscoins per computational unit, unleashing the true power of the blockchain.
+The algorithm proposed in this paper involves prompting the LLM with the agent's state observation, a list of tasks the agent already does well, and a list of available tasks for the LLM to deem as either interesting or boring.
+After sorting available tasks into interesting and boring, it assigns sampling weights to them, giving much higher weights to interesting tasks than boring ones, thus directly influencing the agent's policy by altering probabilities of which action it then chooses to do.
+In \cite{idm}, another paper proposing using the LLM as a policy optimizer, has a similar approach in which the state observation is tokenized and passed to the LLM, though in their experiment, goals and action histories are also tokenized and passed.
+Here, too, the LLM response is fed back into a task-specific RL model, which in turn influences policy probabilities of certain actions being taken.
+In general, many of the novel methods suggested involve the LLM suggesting actions for the agent to perform in a given context, and its output influencing the agent policy.
 
-What is the deguscoin, you may ask?
-The deguscoin represents a complete transformation of the traditional financial system (tradfi).
-By using deguscoin, you will be able to achieve returns of investment (ROIs) beyond mere human comprehension.
-One single deguscoin, when inserted into an economy, can have such destructive power that even mere days after, the market will be beyond recognition.
-By invoking the "degus mindset" (DM) you will be able to send around 15\% more currency per coin, achieving ROIs which would send the average economist into a seizure.
+\subsection{Large Language Models as Reward in Reinforcement Learning}
 
-Market players are predicting that by the end of this year, around 50\% of all transactions will be done under the degus system (DS), not even counting derivatives.
-
-By the end of this year, around 50\% of all transactions will be done under the degus system, not even counting derivatives.
-
-– Market players
-
-What are you waiting for? Purchase deguscoin right now, and leave your earthly desires behind!
+Another often proposed method of involving LLMs in RL, is to have it influence the reward the agent receives in certain states. \cite{ellm}
 

### 2023-10-02 doc/paper.tex

@@
 
-Reinforcement Learning! And LLMs!
+Reinforcement Learning (RL) algorithms suffer when rewards are sparce and the state-action space is large.
+Even tasks which appear relatively simple can prove intractable if the completion of the task requires subtasks to be completed first, or if the reward only comes when the entire task has been completed.
+In such cases, random exploration is unlikely to lead the agent to discover a solution to the problem.
+The apparent simplicity of such problems is often due to human intuition, which allow us to quickly see possible solutions to a diverse set of problems.
+Large Language Models (LLMs) are trained on large corpura of human written text, and have been shown to capture parts of this intuition in many tasks \cite{sparks}.
+In recent years, LLMs have been succesfully used to aid RL agents in more efficient exploration, and have allowed them to solve problems which previous methods have been unable to \cite{omni} \cite{ellm}.
+We compare different methods for integrating an LLM into the deep learning reinforcement algorithm proximal policy optimization (PPO), and compare their efficiency against eachother.
+We find that BEST METHOD gives the best sample efficiency, outperforming normal PPO by PERCENTAGE in METRIC.
 

### 2023-10-02 doc/paper.tex
added latest version of paper

@@
 
-Another often proposed method of involving LLMs in RL, is to have it influence the reward the agent receives in certain states. \cite{ellm}
+Another often proposed method of involving LLMs in RL, is to have it influence the reward the agent receives in certain states.
+In particular, \cite{ellm} cites the infrequency of rewards as a common bottleneck in RL algorithms, due to how long exploratory trajectories often have to be before the agent reaches some desirable state.
+This holds true especially for large, complex environments, where the probability of reaching favorable states among a huge number of non-favorable states becomes even smaller.
+Similarly to the papers proposing LLMs to influence policy, this paper too seeks to make use of human intuition and knowledge present in the vast training data that state-of-the-art LLMs have been trained upon, to make the agent prioritize exploring plausibly useful behaviors first, based on some intelligent forethought provided by the LLM.
+
+However, rather than doing so through directly influencing action probabilities in the policy, the paper suggests rewarding similarity to goal states described by the LLM.
+At certain timesteps, the LLM would then be prompted with a state observation from the agent.
+It would be told to suggest goal states for the agent to reach.
+Some timesteps later, if the new state observation is semantically similar to the previous LLM output, this would give a reward to the agent.
+Thus, the proposed algorithm makes use of LLMs in two ways: first to generate what the LLM perceives as desirable states, then later to assess similarity between a natural-text description of the current state and the previously described goal state.
+The first use case in particular utilizes the presence of human intuition in LLMs, as a human too would often be able to intuitively figure out what is desirable to achieve in some environment.
 

### 2023-10-03 doc/paper.tex
Update paper
Signed-off-by: Jonatan Hanssen <jonatan.hanssen@outlook.com>

@@
 
-\section{Method}
+\section{Methods}
+
+\subsection{Problem Formulation}
+
+We train a PPO model in a Partially Observable Markov Decision Process (POMDP).
+A POMDP is defined by a tuple $(S, A, T, R, O, \Omega, \gamma)$.
+Here, $s \in S$ denotes the state the environment is in and $a \in A$ denotes an action the agent can take.
+Based on the state and the action taken, the agent receives an observation $o \in O$ which depends on the new environment state and the action taken by $O(o | s, a)$.
+$T(s' | s, a)$ denotes the state transfer function, which gives a new state based on the previous state and action.
+The environment we use is part of Minigrid \cite{minigrid}.
 
@@
 
-\section*{References}
-
 \bibliographystyle{apalike}

### 2023-10-03 doc/paper.tex
fix typos

@@
 \textit{The Faculty of Mathematics}\\
+\textit{and Natural Sciences}\\
 Oslo, Norway \\

### 2023-10-03 doc/paper.tex
Update paper with method section
Signed-off-by: Jonatan Hanssen <jonatan.hanssen@outlook.com>

--- 2023-10-03 doc/paper.tex
@@
 Here, $s \in S$ denotes the state the environment is in and $a \in A$ denotes an action the agent can take.
-Based on the state and the action taken, the agent receives an observation $o \in O$ which depends on the new environment state and the action taken by $O(o | s, a)$.
+Based on the state and the action taken, the agent receives an observation $o \in \Omega$ which depends on the new environment state and the action taken by $O(o | s, a)$.
 $T(s' | s, a)$ denotes the state transfer function, which gives a new state based on the previous state and action.
+$r \in R$ is the reward given by the environment given an action and a state, while $\gamma$ is the discount factor.
 The environment we use is part of Minigrid \cite{minigrid}.
 
-Deguser trenger god plass for å trives, og du bør investere i et størst mulig bur som gir dyrene mulighet til å klatre, grave i et tykt strølag og bevege seg i flere etasjer.
+\subsection{Approach}
 
-Mange dyreeiere konstruerer sine egne løsninger, for eksempel med et stort, gammelt akvarium som underdel og et nettingbur på toppen av dette.
-Da får du full oversikt over hva som skjer også nederst i buret, samtidig som du slipper at sand, strø og høy søles ut i rommet.
-Ikke la degusene gå direkte på gitterbunn eller annet hardt underlag, det kan skade føttene.
+Using this environment, we explore three different ways of training the agent:
 
-Buret kan du innrede med f.eks.
-røtter og greiner til å klatre på, hus av treverk, rør til å krype gjennom og løpehjul.
-Et løpehjul til degus bør være minst 25 cm i diameter og ha tett gulv og vegger slik at halen ikke kan komme i klem.
-Bunnmaterialet i buret bør være et tykt lag av støvfritt smådyrstrø som er egnet for graving.
-Det finnes flere egnede strøtyper å få kjøpt.
-Bland gjerne inn litt tørr blomsterjord eller sand i bunnmaterialet.
-Hvis du henter sand ute, bør du desinfisere den ved frysing eller steking fr den brukes i buret.
-Degusene trenger også tilgang på fiberrikt materiale som høy, revet papir eller treull til bygge- og redemateriale.
+\subsubsection{Standard Proximal Policy Optimization}
 
-For å være sikker på at alle degusene som bor sammen kan trekke seg unna og hvile, bør de få hvert sitt sovehus.
-Deguser kan også finne på å forsvare maten sin overfor andre deguser.
-Når flere deguser bor i samme bur, bør du derfor ha flere matskåler plassert på forskjellige steder i buret slik at alle får tilgang til mat når de ønsker det.
-Dyrene må alltid ha tilgang på friskt vann, som best gis i en drikkeflaske.
+A standard PPO algorithm with no LLM integration, as a baseline.
 
-For å holde pelsen i orden har deguser behov for å bade i finkornet sand flere ganger i uka.
-Slik badesand får du kjøpt i dyrebutikken.
-Et tungt kar som ikke velter så lett er fint som sandbadekar.
-Det kan være lurt å fjerne sandbadet mellom hvert bad for å unngå at den brukes som toalett.
+\subsubsection{PPO with LLM integration as part of the policy}
 
-Rengjøring
+where the LLM directly influences which action is taken next
 
-Hvor ofte buret bør rengjøres avhenger av størrelsen og hvor mange som bor der. Som en tommelfingerregel bør du skifte bunnmaterialet og vaske bunn og innredning én gang i uka.
+\subsubsection{LLM integrated into reward function}
+
+LLM integration into the reward function, where actions similar to those suggested by the LLM grant a small auxilliary reward. The agent is trained as in the baseline.
+
+We evaluate these models on the same Minigrid environments and compare their performance against eachother.
 

### 2023-10-03 doc/paper.tex
Update description of method
Signed-off-by: Jonatan Hanssen <jonatan.hanssen@outlook.com>

@@
 
-Another often proposed method of involving LLMs in RL, is to have it influence the reward the agent receives in certain states.
+Another proposed method of involving LLMs in RL, is to have it influence the reward the agent receives for taking certain actions in certain states.
 In particular, \cite{ellm} cites the infrequency of rewards as a common bottleneck in RL algorithms, due to how long exploratory trajectories often have to be before the agent reaches some desirable state.
@@
 
-However, rather than doing so through directly influencing action probabilities in the policy, the paper suggests rewarding similarity to goal states described by the LLM.
-At certain timesteps, the LLM would then be prompted with a state observation from the agent.
-It would be told to suggest goal states for the agent to reach.
-Some timesteps later, if the new state observation is semantically similar to the previous LLM output, this would give a reward to the agent.
-Thus, the proposed algorithm makes use of LLMs in two ways: first to generate what the LLM perceives as desirable states, then later to assess similarity between a natural-text description of the current state and the previously described goal state.
-The first use case in particular utilizes the presence of human intuition in LLMs, as a human too would often be able to intuitively figure out what is desirable to achieve in some environment.
+However, rather than doing so through directly influencing action probabilities in the policy, the paper suggests rewarding the agent if it takes actions that are semantically similar to actions suggested by an LLM.
+At certain timesteps, the LLM is prompted with a state observation from the agent, and is asked to suggest actions for the agent to take.
+If the agent then takes actions which are semantically similar to the suggested actions, it recieves a reward.
+Thus, the proposed algorithm makes use of LLMs in two ways: first to generate what the LLM perceives as desirable actions, then later to assess similarity between a natural-language description of the agents action and the previously suggested actions.
+The first use case in particular utilizes the presence of human intuition in LLMs, as a human would often be able to intuitively figure out what is desirable to achieve in some environment.
 
@@
 
-A standard PPO algorithm with no LLM integration, as a baseline.
+A standard PPO algorithm with no LLM integration, as a baseline. In this case, we StableBaselines' implementation with the following hyperparamers: HYPERPARAMETERS HERE.
 
@@
 
-where the LLM directly influences which action is taken next
+In this case, the LLM is integrated as part of the PPO algorithm, directly influencing the actions taken. This is done by INSERT HOW IT IS DONE HERE.
 
@@
 
+\section*{Appendix}
+
 \end{document}

### 2023-10-03 doc/paper.tex
add paragraph on reward function method

--- 2023-10-03 doc/paper.tex
@@
 
-LLM integration into the reward function, where actions similar to those suggested by the LLM grant a small auxilliary reward. The agent is trained as in the baseline.
+The agent is trained as in the baseline.
+The agent's state observation is encoded to natural text, and sent as part of a prompt to the LLM.
+This may be done at every timestep, or after some interval of timesteps.
+In addition to the state observation, the LLM is also asked to respond with a list of recommendations.
+These recommendations can be proposed actions for the agent to perform, or goal states that the agent should reach.
+The agent then acts independently of the LLM for one or more timesteps, as dictated by the baseline RL algorithm it is trained with.
+As it does so, its actions are tracked and encoded to natural text.
+Alternatively, a new state observation is generated after the timesteps, and also encoded as natural text.
+Then, the LLM is used as a tool for semantic similarity comparison, comparing either its own formerly proposed actions to the actions performed by the agent, or comparing its proposed goal state to the new state reached by the agent.
+Finally, high semantic similarity gives an additional reward which is implemented in the reward function of the agent.
 

### 2023-10-03 doc/paper.tex
Add text on ppo policy llm integration
Signed-off-by: Jonatan Hanssen <jonatan.hanssen@outlook.com>

@@
 These LLM assisted RL agents have been able to outperform state-of-the-art RL methods in many problems \cite{omni} \cite{ellm} \cite{idm}.
-\\
 
@@
 
-In this case, the LLM is integrated as part of the PPO algorithm, directly influencing the actions taken. This is done by INSERT HOW IT IS DONE HERE.
+In this case, the LLM is integrated as part of the PPO algorithm, directly influencing the actions taken.
+This is done as follows: First, the observation is captioned and given as input to the LLM, which is prompted to suggest an action.
+Then all actions that the agent could take, given the current state are described in natural language.
+All possible actions are compared with the suggested action, and given a value which denotes their semantic similarity.
+These values are joined with the probability values that the normal PPO algorithm outputs, and a new probability distribution is created, where semantically similar actions are weighted higher than others.
 

### 2023-10-03 doc/paper.tex
fix typos and minor corrections in paper text

--- 2023-10-03 doc/paper.tex
@@
 
-Reinforcement Learning (RL) algorithms suffer when rewards are sparce and the state-action space is large.
-Even tasks which appear relatively simple can prove intractable if the completion of the task requires subtasks to be completed first, or if the reward only comes when the entire task has been completed.
+Reinforcement Learning (RL) algorithms suffer when rewards are sparse and the state-action space is large.
+Even tasks which seem relatively simple can prove intractable if the completion of the task requires subtasks to be completed first, or if the reward only comes when the entire task has been completed.
 In such cases, random exploration is unlikely to lead the agent to discover a solution to the problem.
 The apparent simplicity of such problems is often due to human intuition, which allow us to quickly see possible solutions to a diverse set of problems.
-Large Language Models (LLMs) are trained on large corpura of human written text, and have been shown to capture parts of this intuition in many tasks \cite{sparks}.
+Large Language Models (LLMs) are trained on large corpora of human written text, and have been shown to capture parts of this intuition in many tasks \cite{sparks}.
 In recent years, LLMs have been succesfully used to aid RL agents in more efficient exploration, and have allowed them to solve problems which previous methods have been unable to \cite{omni} \cite{ellm}.
-We compare different methods for integrating an LLM into the deep learning reinforcement algorithm proximal policy optimization (PPO), and compare their efficiency against eachother.
+We compare different methods for integrating an LLM into the Deep Learning RL algorithm Proximal Policy Optimization (PPO), and compare their efficiency against each other.
 We find that BEST METHOD gives the best sample efficiency, outperforming normal PPO by PERCENTAGE in METRIC.
@@
 In recent years, LLMs have shown remarkable capabilities in problem solving and planning \cite{sparks}, qualities which traditional RL agents often lack.
-Thus, a new area of research has emerged, which attemps to use the vast amount of human knowledge encoded in these models to increase the performance of reinforcement learning algorithms \cite{survey}.
+Thus, a new area of research has emerged, which attemps to use the vast amount of human knowledge encoded in these models to improve the performance of reinforcement learning algorithms \cite{survey}.
 These LLM assisted RL agents have been able to outperform state-of-the-art RL methods in many problems \cite{omni} \cite{ellm} \cite{idm}.
-\\
 
 These papers explore different methods of integrating LLMs into a standard reinforcement learning training loop, from altering the policy directly to introducing an auxilliary reward function.
-In this paper, we compare both methods in the same environment using the same deep reinforcement algorithm (PPO), and compare their results to eachother to explore how LLMs best can be integrated into RL.
+In this paper, we compare both methods in the same environment using the same deep reinforcement algorithm (PPO), and compare their results to each other to explore how LLMs best can be integrated into RL.
 
@@
 The algorithm proposed in this paper involves prompting the LLM with the agent's state observation, a list of tasks the agent already does well, and a list of available tasks for the LLM to deem as either interesting or boring.
-After sorting available tasks into interesting and boring, it assigns sampling weights to them, giving much higher weights to interesting tasks than boring ones, thus directly influencing the agent's policy by altering probabilities of which action it then chooses to do.
+After sorting available tasks into interesting and boring ones, it assigns sampling weights to them, giving much higher weights to interesting tasks than boring ones, thus directly influencing the agent's policy by altering probabilities of which action it then chooses to do.
 In \cite{idm}, another paper proposing using the LLM as a policy optimizer, has a similar approach in which the state observation is tokenized and passed to the LLM, though in their experiment, goals and action histories are also tokenized and passed.
@@
 
-Using this environment, we explore three different ways of training the agent:
+Using this environment, we explore three different ways of training the agent. We evaluate these models on the same Minigrid environments and compare their performance against each other.\\
 
@@
 
-A standard PPO algorithm with no LLM integration, as a baseline. In this case, we StableBaselines' implementation with the following hyperparamers: HYPERPARAMETERS HERE.
+A standard PPO algorithm with no LLM integration, as a baseline. In this case, we add upon StableBaselines' implementation with the following hyperparameters: HYPERPARAMETERS HERE.
 
-\subsubsection{PPO with LLM integration as part of the policy}
+\subsubsection{PPO with LLM Integrated in the Policy Function}
 
-In this case, the LLM is integrated as part of the PPO algorithm, directly influencing the actions taken. This is done by INSERT HOW IT IS DONE HERE.
+In this case, the LLM is integrated as part of the PPO algorithm, directly influencing the actions taken.
+This is done as follows: First, the observation is captioned and given as input to the LLM, which is prompted to suggest an action.
+Then, all actions that the agent could take given the current state are described in natural language.
+All possible actions are compared with the suggested action, and given a value which denotes their semantic similarity.
+These values are joined with the probability values that the normal PPO algorithm outputs and normalized, and a new probability distribution is created, where semantically similar actions are weighted higher than others.
 
-\subsubsection{LLM integrated into reward function}
+\subsubsection{PPO with LLM Integrated in the Reward Function}
 
@@
 Finally, high semantic similarity gives an additional reward which is implemented in the reward function of the agent.
-
-We evaluate these models on the same Minigrid environments and compare their performance against eachother.
 

### 2023-10-03 doc/paper.tex
update paper
Signed-off-by: Jonatan Hanssen <jonatan.hanssen@outlook.com>

--- 2023-10-03 doc/paper.tex
@@
 \section{Background and Related Work}
+
+RL informed by natural language is a relatively new field, where research before 2018 have been limited to relatively small corpora or synthetic language \cite[1]{survey}.
+With the introduction of Large Language Models, many new possibilities have opened up, and in recent years there have been many successful integrations of these models into traditional RL \cite{omni} \cite{ellm} \cite{idm}.
 

### 2023-10-04 doc/paper.tex
Change "add upon" to "use"
Using hyperparameters does not add anything to an implementation, thats
how you use every model

Signed-off-by: Jonatan Hanssen <jonatan.hanssen@outlook.com>
change paragraph about llm as policy in paper

@@
 
-In recent research into integrating LLMs into the RL framework, many of the methods proposed include ways of making the LLM influence the policy of the RL agent.
+Among the methods proposed to integrate the LLM, many include ways of making the LLM influence the policy of the RL agent.
 \cite{omni} considers open-ended learning algorithms in which the value of the agent attempting certain tasks can be evaluated by estimating the the interestingness of the task.
@@
 
-A standard PPO algorithm with no LLM integration, as a baseline. In this case, we add upon StableBaselines' implementation with the following hyperparameters: HYPERPARAMETERS HERE.
+A standard PPO algorithm with no LLM integration, as a baseline. In this case, use StableBaselines' implementation with the following hyperparameters: HYPERPARAMETERS HERE.
 

### 2023-10-05 src/PPO/ActorCritic.py
Renaming
Signed-off-by: Gregz9 <gregorkajda@gmail.com>

@@ -1,0 +1,94 @@
+import argparse
+import os
+from distutils.util import strtobool
+import time
+from torch.utils.tensorboard import SummaryWriter
+import numpy as np
+import random
+import torch
+import gymnasium as gym
+import torch.nn as nn
+import torch.optim as optim
+from torch.distributions.categorical import Categorical
+
+
+def init_weightsNbias(layer, std=np.sqrt(2), bias_const=0.0):
+    """
+    Descr:
+        Most, if not all implementations of PPO seen so far use the approach
+        of initializing weights orthogonaly. According to both the authors behind
+        Proximal Policy Optimization, and the paper by title "Exact solutions to
+        the nonlinear dynamics of learning in deep learning networks", the approach
+        of creating such initial conditions leads to faithful propagation of gradients,
+        and better convergence (See page 2 of the latter paper).
+    """
+    torch.nn.init.orthogonal_(layer.weight, std)
+    torch.nn.init.constant_(layer.bias, bias_const)
+    return layer
+
+
+def make_env(env_name, seed):
+    def env_gen():
+        env = gym.make(env_name, render_mode="rgb_array")
+        env = gym.wrappers.RecordEpisodeStatistics(env)
+        env.action_space.seed(seed)
+        env.observation_space.seed(seed)
+        return env
+
+    return env_gen
+
+
+class Agent(nn.Module):
+    """
+    Descr:
+        The agents task is to decide the next action to perform, and evaluate the
+        possible future rewards, a.k.a. the value function, based on the current state
+        the agent is in. In our case the agents chooses the
+
+    """
+
+    def __init__(self, envs):
+        super(Agent, self).__init__()
+        self.critic = nn.Sequential(
+            init_weightsNbias(nn.Linear(np.array(envs.single_observation_space["image"].shape).prod(), 64)),
+            nn.Tanh(),
+            init_weightsNbias(nn.Linear(64, 64)),
+            nn.Tanh(),
+            init_weightsNbias(nn.Linear(64, 1), std=1.0),
+        )
+
+        self.actor = nn.Sequential(
+            init_weightsNbias(nn.Linear(np.array(envs.single_observation_space["image"].shape).prod(), 64)),
+            nn.Tanh(),
+            init_weightsNbias(nn.Linear(64, 64)),
+            nn.Tanh(),
+            init_weightsNbias(nn.Linear(64, envs.single_action_space.n), std=0.01),
+        )
+
[…]

### 2023-10-05 src/PP_Optimization.py
Renaming

@@ -1,94 +1,0 @@
-import argparse
-import os
-from distutils.util import strtobool
-import time
-from torch.utils.tensorboard import SummaryWriter
-import numpy as np
-import random
-import torch
-import gymnasium as gym
-import torch.nn as nn
-import torch.optim as optim
-from torch.distributions.categorical import Categorical
-
-
-def init_weightsNbias(layer, std=np.sqrt(2), bias_const=0.0):
-    """
-    Descr:
-        Most, if not all implementations of PPO seen so far use the approach
-        of initializing weights orthogonaly. According to both the authors behind
-        Proximal Policy Optimization, and the paper by title "Exact solutions to
-        the nonlinear dynamics of learning in deep learning networks", the approach
-        of creating such initial conditions leads to faithful propagation of gradients,
-        and better convergence (See page 2 of the latter paper).
-    """
-    torch.nn.init.orthogonal_(layer.weight, std)
-    torch.nn.init.constant_(layer.bias, bias_const)
-    return layer
-
-
-def make_env(env_name, seed):
-    def env_gen():
-        env = gym.make(env_name, render_mode="rgb_array")
-        env = gym.wrappers.RecordEpisodeStatistics(env)
-        env.action_space.seed(seed)
-        env.observation_space.seed(seed)
-        return env
-
-    return env_gen
-
-
-class Agent(nn.Module):
-    """
-    Descr:
-        The agents task is to decide the next action to perform, and evaluate the
-        possible future rewards, a.k.a. the value function, based on the current state
-        the agent is in. In our case the agents chooses the
-
-    """
-
-    def __init__(self, envs):
-        super(Agent, self).__init__()
-        self.critic = nn.Sequential(
-            init_weightsNbias(nn.Linear(np.array(envs.single_observation_space["image"].shape).prod(), 64)),
-            nn.Tanh(),
-            init_weightsNbias(nn.Linear(64, 64)),
-            nn.Tanh(),
-            init_weightsNbias(nn.Linear(64, 1), std=1.0),
-        )
-
-        self.actor = nn.Sequential(
-            init_weightsNbias(nn.Linear(np.array(envs.single_observation_space["image"].shape).prod(), 64)),
-            nn.Tanh(),
-            init_weightsNbias(nn.Linear(64, 64)),
-            nn.Tanh(),
-            init_weightsNbias(nn.Linear(64, envs.single_action_space.n), std=0.01),
-        )
-
[…]

### 2023-10-05 src/test_reward.py
Update action captioning
Signed-off-by: Jonatan Hanssen <jonatan.hanssen@outlook.com>

@@ -4,7 +4,7 @@ import time
 import constants
 import numpy as np
 
-from utils import obs_to_string, llama2_7b_reward_shaper
+from utils import obs_to_string, llama2_7b_reward_shaper, caption_action
 
 env = gym.make("MiniGrid-UnlockPickup-v0", render_mode="human")
 
@@ -22,7 +22,6 @@ for _ in range(1000):
     # for row in stringy:
     #     print(row)
     obs = obs_to_string(observation["image"], False, False)
-    action = 1
 
     rewarder.suggest(obs)
 
@@ -30,6 +29,13 @@ for _ in range(1000):
 
 
     rewarder.compare(action, observation["image"])
+
+
+    # all_captions = [
+    #     caption_action(act, observation["image"]) for act in range(7)
+    # ]
+
+
     observation, reward, terminated, truncated, info = env.step(action)
 
 

### 2023-10-05 src/utils.py

@@ -125,7 +125,7 @@ class llama2_7b_reward_shaper:
         Returns:
             a reward if action and one of the suggested actions is semantically similar
         """
-        caption = caption(action, obs_matrix)
+        caption = caption_action(action, obs_matrix)
 
         if not caption or caption in self.caption_set:
             return 0
@@ -150,7 +150,7 @@ class llama2_7b_reward_shaper:
             return 0
 
 
-def caption_action(action, observation):
+def caption_action(action, obs_matrix):
     """takes in an action and the observation and returns a string caption"""
 
     front_cell = obs_matrix[3, 5]
@@ -171,27 +171,39 @@ def caption_action(action, observation):
 
     ###  Captioning action based on cell and inventory ###
 
-    caption = None
+    caption = "do nothing"
+
+    if action == "done":
+        caption = "mark the game as completed"
+
+    if action == "forward":
+        if front_item == "empty":
+            caption = "move forward"
+        elif front_item == "door" and front_state == " open":
+            caption = "move forward"
+
+    if action == "right":
+        caption = "turn right"
+
+    if action == "left":
+        caption = "turn left"
 
     if action == "pick":
         if front_cell[0] < 4:  # nothing to pick up
-            caption = "do nothing"
+            pass
 
         elif inventory[0] > 3:  # inventory already full
-            caption = "do nothing"
+            pass
 
         else:
             caption = f"pick up{front_state} {front_color} {front_item}"
 
     elif action == "drop":
         if inventory[0] < 4:  # nothing to drop
-            caption = "do nothing"
-
-        elif inventory[0] < 4:  # nothing to drop
-            caption = "do nothing"
+            pass
 
         elif front_item != "empty":  # cannot drop onto occupied tile
-            caption = "do nothing"
+            pass
 
         else:
             caption = f"drop{inventory_state} {inventory_color} {inventory_item}"
@@ -200,13 +212,12 @@ def caption_action(action, observation):
         if front_item == "box":
             caption = f"destroy {front_color} box"
         else:
-            if front_cell[0] < 4:
-                caption = "do nothing"
-            else:
+            if front_cell[0] >= 4:
                 caption = f"use{front_state} {front_color} {front_item}"
 
             if inventory[0] > 3:
                 caption += f" with{inventory_state} {inventory_color} {inventory_item}"
+
     return caption
 
 

### 2023-10-06 src/PPO/PP_Optimization.py
Backwards computation

@@ -28,6 +28,12 @@ def init_weightsNbias(layer, std=np.sqrt(2), bias_const=0.0):
 
 
 def make_env(env_name, seed):
+    """
+    Descr:
+        Returns a "vectorized" environenment, meaning that it is wrapped in a gymnasium vector, allowing for
+        some aditional functionality, such as parallel training of many environments.
+    """
+
     def env_gen():
         env = gym.make(env_name, render_mode="rgb_array")
         env = gym.wrappers.RecordEpisodeStatistics(env)
@@ -43,8 +49,7 @@ class Agent(nn.Module):
     Descr:
         The agents task is to decide the next action to perform, and evaluate the
         possible future rewards, a.k.a. the value function, based on the current state
-        the agent is in. In our case the agents chooses the
-
+        the agent is in.
     """
 
     def __init__(self, envs):
@@ -109,7 +114,7 @@ class PPO:
 
     def forward(self):
         # Containers for values needed in calculation of surrogate loss
-        single_obs_shape = (
+        self.single_obs_shape = (
             int(np.array(envs.single_observation_space["image"].shape).prod()),
         )
 
@@ -119,6 +124,7 @@ class PPO:
         self.actions = torch.zeros(
             (self.args.ep_steps, self.args.num_envs) + self.envs.single_action_space
         ).to(device)
+        # The ActorCritic Network outputs log probabilities
         self.logprobs = torch.zeros((self.args.ep_steps, self.args.num_envs)).to(device)
         self.rewards = torch.zeros((self.args.ep_steps, self.args.num_envs)).to(device)
         self.dones = torch.zeros((self.args.ep_steps, self.args.num_envs)).to(device)
@@ -130,13 +136,15 @@ class PPO:
 
         # Episode: Moving n steps and estimating the value funvtion for each step
         for rollout in range(self.rollouts + 1):
-            for step in self.args.ep_steps:
+            for step in range(self.args.ep_steps):
                 self.global_step += self.args.num_envs
                 self.observations[step] = next_observation
                 self.dones[step] = next_done
 
                 with torch.no_grad():
-                    action, logprob, _, value = self.agent(next_observation)
+                    action, logprob, _, value = self.agent.get_action_and_value(
+                        next_observation
+                    )
                     self.values[step] = value.flatten()
                 self.actions[step] = action
                 self.logprobs[step] = logprob
@@ -182,6 +190,8 @@ class PPO:
[…]

### 2023-10-06 src/PPO/ActorCritic.py
Further coding
Signed-off-by: Gregz9 <gregorkajda@gmail.com>

@@ -2,7 +2,8 @@ import argparse
 import os
 from distutils.util import strtobool
 import time
-from torch.utils.tensorboard import SummaryWriter
+
+# from torch.utils.tensorboard import SummaryWriter
 import numpy as np
 import random
 import torch
@@ -50,7 +51,11 @@ class Agent(nn.Module):
     def __init__(self, envs):
         super(Agent, self).__init__()
         self.critic = nn.Sequential(
-            init_weightsNbias(nn.Linear(np.array(envs.single_observation_space["image"].shape).prod(), 64)),
+            init_weightsNbias(
+                nn.Linear(
+                    np.array(envs.single_observation_space["image"].shape).prod(), 64
+                )
+            ),
             nn.Tanh(),
             init_weightsNbias(nn.Linear(64, 64)),
             nn.Tanh(),
@@ -58,7 +63,11 @@ class Agent(nn.Module):
         )
 
         self.actor = nn.Sequential(
-            init_weightsNbias(nn.Linear(np.array(envs.single_observation_space["image"].shape).prod(), 64)),
+            init_weightsNbias(
+                nn.Linear(
+                    np.array(envs.single_observation_space["image"].shape).prod(), 64
+                )
+            ),
             nn.Tanh(),
             init_weightsNbias(nn.Linear(64, 64)),
             nn.Tanh(),
@@ -85,10 +94,19 @@ if __name__ == "__main__":
     # print(envs.single_action_space.n)
     # observation, info = envs.reset(seed=42)
     # print(observation["image"])
-    print(observation.transpose(2,0,1))
-    # agent = Agent(envs)
+    # print(observation.transpose(2,0,1))
+    agent = Agent(envs)
+    # print(agent.state_dict())
+    # src_dir = os.path.dirname(os.path.abspath(os.path.dirname(__file__)))
+    src_dir = os.path.dirname(__file__)
+
+    torch.save(agent.state_dict(), src_dir + "/PPO.txt")
+
+    model = Agent(envs)
+    model.load_state_dict(torch.load(src_dir + "/PPO.txt"))
+    
     # next_observation = torch.Tensor(envs.reset()[0])
 
-    # action, _, _, _ = agent.get_action_and_value(next_observation)
-    # print(action) 
 
+    # action, _, _, _ = agent.get_action_and_value(next_observation)
+    # print(action)

### 2023-10-06 src/PPO/PP_Optimization.py

@@ -43,7 +43,7 @@ def make_env(env_name, seed):
 
     return env_gen
 
-
+# TODO: Transfer model to device a.k.a. to GPU
 class Agent(nn.Module):
     """
     Descr:
@@ -100,9 +100,10 @@ class Agent(nn.Module):
 
 class PPO:
     def __init__(self, args):
+        # TODO: Replace Argparse with a json containing all parameters and clean up __init__
         self.args = args()
         self.seeding()
-
+        # TODO: Remove gymnasium vectorization of environment
         seeds = [random.randint(0, 20000) for _ in range(env_n)]
         self.envs = gym.vector.SyncVector(
             [make_env(env_name, seeds[i]) for i in range(self.args.num_envs)]
@@ -112,7 +113,15 @@ class PPO:
         _, _ = envs.reset(seed=self.args.seed)
         self.rollouts = self.args.tot_steps // self.args.batch_size
 
-    def forward(self):
+    def save_model(self, model):
+        file_path = os.path.dirname(__file__) + "/model.zip"
+        torch.save(self.agent.state_dict(), file_path)
+
+    def load_model(self, model):
+        file_path = os.path.dirname(__file__) + "/model.zip"
+        model.load_state_dict(torch.load(file_path))
+
+    def train(self):
         # Containers for values needed in calculation of surrogate loss
         self.single_obs_shape = (
             int(np.array(envs.single_observation_space["image"].shape).prod()),
@@ -131,138 +140,152 @@ class PPO:
         self.values = torch.zeros((self.args.ep_steps, self.args.num_envs)).to(device)
 
         # Initializing the next step
-        next_observation = torch.zeros(self.envs.reset()[0]).to(device)
-        next_done = torch.zeros(self.args.num_envs).to(device)
+        # next_observation = torch.zeros(self.envs.reset()[0]).to(device)
+        # next_done = torch.zeros(self.args.num_envs).to(device)
 
         # Episode: Moving n steps and estimating the value funvtion for each step
         for rollout in range(self.rollouts + 1):
-            for step in range(self.args.ep_steps):
-                self.global_step += self.args.num_envs
-                self.observations[step] = next_observation
-                self.dones[step] = next_done
+            self.forward(rollout)
+            self.backward()
 
-                with torch.no_grad():
-                    action, logprob, _, value = self.agent.get_action_and_value(
-                        next_observation
-                    )
-                    self.values[step] = value.flatten()
-                self.actions[step] = action
[…]

### 2023-10-16 src/PPO/PP_Optimization.py
Update to structure of the PP_Optimization.py

@@ -10,6 +10,23 @@ import gymnasium as gym
 import torch.nn as nn
 import torch.optim as optim
 from torch.distributions.categorical import Categorical
+import json
+
+
+def read_params():
+    file = open("hyperparams.json")
+    params = json.load(file)
+    # print(json.dumps(params, indent=4, separators=(":", ",")))
+    return params
+
+
+def save_params(params):
+    file = open("hyperparams.json", "w")
+    json.dump(params, file, indent=4, separators=(",", ":"))
+
+
+def sample_minibatches():
+    pass
 
 
 def init_weightsNbias(layer, std=np.sqrt(2), bias_const=0.0):
@@ -43,6 +60,7 @@ def make_env(env_name, seed):
 
     return env_gen
 
+
 # TODO: Transfer model to device a.k.a. to GPU
 class Agent(nn.Module):
     """
@@ -52,9 +70,9 @@ class Agent(nn.Module):
         the agent is in.
     """
 
-    def __init__(self, envs):
-        super(Agent, self).__init__()
-        self.critic = nn.Sequential(
+    def __init__(, envs):
+        super(Agent, ).__init__()
+        .critic = nn.Sequential(
             init_weightsNbias(
                 nn.Linear(
                     np.array(
@@ -69,7 +87,7 @@ class Agent(nn.Module):
             init_weightsNbias(nn.Linear(64, 1), std=1.0),
         )
 
-        self.actor = nn.Sequential(
+        .actor = nn.Sequential(
             init_weightsNbias(
                 nn.Linear(
                     np.array(
@@ -85,215 +103,220 @@ class Agent(nn.Module):
         )
 
         # Yet to be integrated -> shall serve as the second actor
-        # self.LM_actor = llama2_7b_policy()
+        # .LM_actor = llama2_7b_policy()
 
-    def get_value(self, X):
-        return self.critic(X)
+    def get_value(, X):
+        return .critic(X)
 
-    def get_action_and_value(self, X, action=None):
-        logits = self.actor(X)
+    def get_action_and_value(, X, action=None):
+        logits = .actor(X)
         probs = Categorical(logits=logits)
         if action is None:
             action = probs.sample()
-        return action, probs.log_prob(action), probs.entropy(), self.critic(X)
+        return action, probs.log_prob(action), probs.entropy(), .critic(X)
 
 
-class PPO:
-    def __init__(self, args):
-        # TODO: Replace Argparse with a json containing all parameters and clean up __init__
[…]

### 2023-10-16 src/PPO/PP_Optimization.py
Molding the PPO
Signed-off-by: Gregz9 <gregorkajda@gmail.com>

@@ -25,8 +25,10 @@ def save_params(params):
     json.dump(params, file, indent=4, separators=(",", ":"))
 
 
-def sample_minibatches():
-    pass
+def minibatch_generator():
+    batch_idx = np.random.choice(args["batch_size"], args["batch_size"], replace=False)
+    for start in range(0, args["batch_size"], args["minibatch_size"]):
+        end = start + args["minibatch_size"]
 
 
 def init_weightsNbias(layer, std=np.sqrt(2), bias_const=0.0):
@@ -70,9 +72,9 @@ class Agent(nn.Module):
         the agent is in.
     """
 
-    def __init__(, envs):
-        super(Agent, ).__init__()
-        .critic = nn.Sequential(
+    def __init__(self, envs):
+        super(Agent, self).__init__()
+        self.critic = nn.Sequential(
             init_weightsNbias(
                 nn.Linear(
                     np.array(
@@ -87,7 +89,7 @@ class Agent(nn.Module):
             init_weightsNbias(nn.Linear(64, 1), std=1.0),
         )
 
-        .actor = nn.Sequential(
+        self.actor = nn.Sequential(
             init_weightsNbias(
                 nn.Linear(
                     np.array(
@@ -105,15 +107,15 @@ class Agent(nn.Module):
         # Yet to be integrated -> shall serve as the second actor
         # .LM_actor = llama2_7b_policy()
 
-    def get_value(, X):
-        return .critic(X)
+    def get_value(self, X):
+        return self.critic(X)
 
-    def get_action_and_value(, X, action=None):
-        logits = .actor(X)
+    def get_action_and_value(self, X, action=None):
+        logits = self.actor(X)
         probs = Categorical(logits=logits)
         if action is None:
             action = probs.sample()
-        return action, probs.log_prob(action), probs.entropy(), .critic(X)
+        return action, probs.log_prob(action), probs.entropy(), self.critic(X)
 
 
 def PPO(param_path):
@@ -127,6 +129,11 @@ def PPO(param_path):
     optimizer = optim.Adam(agent.parameters(), lr=args.lr, eps=1e-5)
     _, _ = envs.reset(seed=args["seed"])
     args["rollouts"] = args["tot_steps"] // args["batch_size"]
+
+def checkpoint():
+    # TODO: Implement a function for saving model when the perforamce reaches a certain level.
+    # Intention is to use the "save_model(model)" method
+    raise NotImplementedError()
 
 
 def save_model(model):
@@ -144,7 +151,8 @@ def PPO_train(envs):
[…]

### 2023-10-16 src/PPO/PP_Optimization.py
Update pp code
Signed-off-by: Jonatan Hanssen <jonatan.hanssen@outlook.com>

@@ -2,7 +2,6 @@ import argparse
 import os
 from distutils.util import strtobool
 import time
-from torch.utils.tensorboard import SummaryWriter
 import numpy as np
 import random
 import torch
@@ -13,15 +12,15 @@ from torch.distributions.categorical import Categorical
 import json
 
 
-def read_params():
-    file = open("hyperparams.json")
+def read_params(params):
+    file = open(params)
     params = json.load(file)
     # print(json.dumps(params, indent=4, separators=(":", ",")))
     return params
 
 
 def save_params(params):
-    file = open("hyperparams.json", "w")
+    file = open(params, "w")
     json.dump(params, file, indent=4, separators=(",", ":"))
 
 
@@ -118,17 +117,19 @@ class Agent(nn.Module):
         return action, probs.log_prob(action), probs.entropy(), self.critic(X)
 
 
-def PPO(param_path):
-    args = read_params()
-    seeding()
-    seeds = [random.randint(0, 20000) for _ in range(args["env_n"])]
-    envs = gym.vector.SyncVector(
-        [make_env(env_name, seeds[i]) for i in range(args["num_envs"])]
+def PPO(param_path, device="cpu"):
+    args = read_params(param_path)
+    args["batch_size"] = args["num_envs"] * args["tot_steps"]
+    seeding(args["seed"])
+    seeds = [random.randint(0, 20000) for _ in range(args["num_envs"])]
+    envs = gym.vector.SyncVectorEnv(
+        [make_env(args["env_name"], seeds[i]) for i in range(args["num_envs"])]
     )
     agent = Agent(envs).to(device)
-    optimizer = optim.Adam(agent.parameters(), lr=args.lr, eps=1e-5)
+    optimizer = optim.Adam(agent.parameters(), lr=args["lr"], eps=1e-5)
     _, _ = envs.reset(seed=args["seed"])
     args["rollouts"] = args["tot_steps"] // args["batch_size"]
+
 
 def checkpoint():
     # TODO: Implement a function for saving model when the perforamce reaches a certain level.
@@ -151,7 +152,7 @@ def PPO_train(envs):
     single_obs_shape = (
         int(np.array(envs.single_observation_space["image"].shape).prod()),
     )
-    
+
     # The following tensors need not to be initialized to again
     observations = torch.zeros(
         (args["ep_steps"], args["num_envs"], single_obs_shape),
@@ -200,6 +201,7 @@ def env_episode(rollout_num, agent):
[…]

### 2023-10-16 src/PPO/PP_Optimization.py
hyperparams and some more integration for PPO -> Gregz's still working

@@ -130,25 +130,6 @@ def PPO(param_path, device="cpu"):
     _, _ = envs.reset(seed=args["seed"])
     args["rollouts"] = args["tot_steps"] // args["batch_size"]
 
-
-def checkpoint():
-    # TODO: Implement a function for saving model when the perforamce reaches a certain level.
-    # Intention is to use the "save_model(model)" method
-    raise NotImplementedError()
-
-
-def save_model(model):
-    file_path = os.path.dirname(__file__) + "/model.zip"
-    torch.save(model.state_dict(), file_path)
-
-
-def load_model(model):
-    file_path = os.path.dirname(__file__) + "/model.zip"
-    model.load_state_dict(torch.load(file_path))
-
-
-def PPO_train(envs):
-    # Containers for values needed in calculation of surrogate loss
     single_obs_shape = (
         int(np.array(envs.single_observation_space["image"].shape).prod()),
     )
@@ -172,12 +153,29 @@ def PPO_train(envs):
 
     # Episode: Moving n steps and estimating the value funvtion for each step
     for rollout in range(args["rollouts"] + 1):
-        env_episode(rollout)
-        comp_advantage(gae=args["gae"])
-        PPO_update()
+        obs, acts, logs, rews, dones, vals, adv, rets = env_episode(
+            rollout, observations, actions, logprobs, rewards, dones, values
+        )
+        PPO_update(agent, optimizer, obs, acts, logs, rews, dones, vals, adv, rets)
 
 
-def env_episode(rollout_num, agent):
+def checkpoint():
+    # TODO: Implement a function for saving model when the perforamce reaches a certain level.
+    # Intention is to use the "save_model(model)" method
+    raise NotImplementedError()
+
+
+def save_model(model):
+    file_path = os.path.dirname(__file__) + "/model.zip"
+    torch.save(model.state_dict(), file_path)
+
+
+def load_model(model):
+    file_path = os.path.dirname(__file__) + "/model.zip"
+    model.load_state_dict(torch.load(file_path))
+
+
+def env_episode(rollout_num, agent, **kwargs):
     for step in range(args["ep_steps"]):
         global_step += args["num_envs"]
         if rollout_num == 0:
@@ -201,8 +199,6 @@ def env_episode(rollout_num, agent):
             torch.Tensor(next_done).to(device),
         )
 
-
-def comp_advantage(gae=False):
     # Now that the agent(-s) has(-ve) played out an episode, it's time
     # to backtrack all steps, and compute the discounted rewards
     with torch.no_grad():
[…]

### 2023-10-16 src/PPO/PP_Optimization.py
Fixing errors in PP_O

@@ -20,7 +20,7 @@ def read_params(params):
 
 
 def save_params(params):
-    file = open(params, "w")
+    file = open("hyperparams.json", "w")
     json.dump(params, file, indent=4, separators=(",", ":"))
 
 
@@ -117,7 +117,7 @@ class Agent(nn.Module):
         return action, probs.log_prob(action), probs.entropy(), self.critic(X)
 
 
-def PPO(param_path, device="cpu"):
+def PPO(param_path, device=torch.device("cpu")):
     args = read_params(param_path)
     args["batch_size"] = args["num_envs"] * args["tot_steps"]
     seeding(args["seed"])
@@ -130,16 +130,17 @@ def PPO(param_path, device="cpu"):
     _, _ = envs.reset(seed=args["seed"])
     args["rollouts"] = args["tot_steps"] // args["batch_size"]
 
-    single_obs_shape = (
-        int(np.array(envs.single_observation_space["image"].shape).prod()),
+    single_obs_shape = int(
+        np.array(envs.single_observation_space["image"].shape).prod()
     )
+    # print(envs.single_observation_space)
 
     # The following tensors need not to be initialized to again
     observations = torch.zeros(
         (args["ep_steps"], args["num_envs"], single_obs_shape),
     ).to(device)
     actions = torch.zeros(
-        (args["ep_steps"], args["num_envs"]) + envs.single_action_space
+        (args["ep_steps"], args["num_envs"]) + envs.single_action_space.shape
     ).to(device)
     # The ActorCritic Network outputs log probabilities
     logprobs = torch.zeros((args["ep_steps"], args["num_envs"])).to(device)
@@ -154,9 +155,19 @@ def PPO(param_path, device="cpu"):
     # Episode: Moving n steps and estimating the value funvtion for each step
     for rollout in range(args["rollouts"] + 1):
         obs, acts, logs, rews, dones, vals, adv, rets = env_episode(
-            rollout, observations, actions, logprobs, rewards, dones, values
+            agent,
+            args,
+            device,
+            envs,
+            rollout,
+            observations,
+            actions,
+            logprobs,
+            rewards,
+            dones,
+            values,
         )
-        PPO_update(agent, optimizer, obs, acts, logs, rews, dones, vals, adv, rets)
+        PPO_update(optimizer, agent, device, obs, acts, logs, rews, dones, vals, adv, rets)
 
 
 def checkpoint():
[…]

### 2023-10-16 src/PPO/PP_Optimization.py
Gregz's working on some last small things, and then restructuring of the whole PPO needs to be done, some noormalizations have to be added, logging, checkpointing, implementation of the advisor etc.

@@ -119,7 +119,8 @@ class Agent(nn.Module):
 
 def PPO(param_path, device=torch.device("cpu")):
     args = read_params(param_path)
-    args["batch_size"] = args["num_envs"] * args["tot_steps"]
+    args["batch_size"] = int(args["num_envs"] * args["ep_steps"])
+    args["minibatch_size"] = int(args["batch_size"] // args["num_minibatches"])
     seeding(args["seed"])
     seeds = [random.randint(0, 20000) for _ in range(args["num_envs"])]
     envs = gym.vector.SyncVectorEnv(
@@ -128,7 +129,7 @@ def PPO(param_path, device=torch.device("cpu")):
     agent = Agent(envs).to(device)
     optimizer = optim.Adam(agent.parameters(), lr=args["lr"], eps=1e-5)
     _, _ = envs.reset(seed=args["seed"])
-    args["rollouts"] = args["tot_steps"] // args["batch_size"]
+    args["rollouts"] = int(args["tot_steps"] // args["batch_size"])
 
     single_obs_shape = int(
         np.array(envs.single_observation_space["image"].shape).prod()
@@ -167,7 +168,21 @@ def PPO(param_path, device=torch.device("cpu")):
             dones,
             values,
         )
-        PPO_update(optimizer, agent, device, obs, acts, logs, rews, dones, vals, adv, rets)
+        PPO_update(
+            optimizer,
+            agent,
+            device,
+            envs,
+            args,
+            obs,
+            acts,
+            logs,
+            rews,
+            dones,
+            vals,
+            adv,
+            rets,
+        )
 
 
 def checkpoint():
@@ -198,14 +213,16 @@ def env_episode(
     rewards,
     dones,
     values,
-    gae=True
+    gae=True,
 ):
     for step in range(args["ep_steps"]):
         # global_step += args["num_envs"]
         if rollout_num == 0:
             # print(envs.reset()[0]["image"].flatten())
-            next_observation = torch.Tensor(envs.reset()[0]["image"].flatten()).to(device)
-            next_done = torch.zeros(args["num_envs"]).to(device) 
+            next_observation = torch.Tensor(envs.reset()[0]["image"].flatten()).to(
+                device
+            )
+            next_done = torch.zeros(args["num_envs"]).to(device)
 
         observations[step] = next_observation
         dones[step] = next_done
@@ -216,8 +233,10 @@ def env_episode(
             values[step] = value.flatten()
         actions[step] = action
[…]

### 2023-10-17 src/PPO/PP_Optimization.py
A working pile of shit
Signed-off-by: Gregz9 <gregorkajda@gmail.com>

@@ -82,9 +82,10 @@ class Agent(nn.Module):
                     64,
                 )
             ),
-            nn.Tanh(),
-            init_weightsNbias(nn.Linear(64, 64)),
-            nn.Tanh(),
+            # nn.ReLU(),
+            # init_weightsNbias(nn.Linear(64, 64)),
+            # nn.Tanh(),
+            nn.ReLU(),
             init_weightsNbias(nn.Linear(64, 1), std=1.0),
         )
 
@@ -97,9 +98,10 @@ class Agent(nn.Module):
                     64,
                 )
             ),
-            nn.Tanh(),
-            init_weightsNbias(nn.Linear(64, 64)),
-            nn.Tanh(),
+            # nn.ReLU(),
+            # init_weightsNbias(nn.Linear(64, 64)),
+            # nn.Tanh(),
+            nn.ReLU(),
             init_weightsNbias(nn.Linear(64, envs.single_action_space.n), std=0.01),
         )
 
@@ -128,7 +130,7 @@ def PPO(param_path, device=torch.device("cpu")):
     )
     agent = Agent(envs).to(device)
     optimizer = optim.Adam(agent.parameters(), lr=args["lr"], eps=1e-5)
-    _, _ = envs.reset(seed=args["seed"])
+    # _, _ = envs.reset(seed=args["seed"])
     args["rollouts"] = int(args["tot_steps"] // args["batch_size"])
 
     single_obs_shape = int(
@@ -150,16 +152,31 @@ def PPO(param_path, device=torch.device("cpu")):
     values = torch.zeros((args["ep_steps"], args["num_envs"])).to(device)
 
     # Initializing the next step
-    # next_observation = torch.zeros(.envs.reset()[0]).to(device)
+    next_observation = torch.Tensor(envs.reset()[0]["image"].flatten()).to(device)
+    next_done = torch.zeros(args["num_envs"]).to(device)
     # next_done = torch.zeros(.args.num_envs).to(device)
 
     # Episode: Moving n steps and estimating the value funvtion for each step
     for rollout in range(args["rollouts"] + 1):
-        obs, acts, logs, rews, dones, vals, adv, rets = env_episode(
+        print(f"Rollout num: {rollout}")
+        (
+            obs,
+            acts,
+            logs,
+            rews,
+            dones,
+            vals,
+            adv,
+            rets,
+            next_observation,
+            next_done,
+        ) = env_episode(
             agent,
             args,
             device,
             envs,
+            next_observation,
[…]

### 2023-10-17 src/PPO/ActorCritic.py
Restructuring the code

@@ -1,13 +1,7 @@
-import argparse
 import os
-from distutils.util import strtobool
-import time
-
 # from torch.utils.tensorboard import SummaryWriter
 import numpy as np
-import random
 import torch
-import gymnasium as gym
 import torch.nn as nn
 import torch.optim as optim
 from torch.distributions.categorical import Categorical
@@ -28,24 +22,29 @@ def init_weightsNbias(layer, std=np.sqrt(2), bias_const=0.0):
     return layer
 
 
-def make_env(env_name, seed):
-    def env_gen():
-        env = gym.make(env_name, render_mode="rgb_array")
-        env = gym.wrappers.RecordEpisodeStatistics(env)
-        env.action_space.seed(seed)
-        env.observation_space.seed(seed)
-        return env
-
-    return env_gen
+def save_model(model):
+    file_path = os.path.dirname(__file__) + "/model.zip"
+    torch.save(model.state_dict(), file_path)
 
 
+def load_model(model):
+    file_path = os.path.dirname(__file__) + "/model.zip"
+    model.load_state_dict(torch.load(file_path))
+
+
+def checkpoint():
+    # TODO: Implement a function for saving model when the perforamce reaches a certain level.
+    # Intention is to use the "save_model(model)" method
+    raise NotImplementedError()
+
+
+# TODO: Transfer model to device a.k.a. to GPU
 class Agent(nn.Module):
     """
     Descr:
         The agents task is to decide the next action to perform, and evaluate the
         possible future rewards, a.k.a. the value function, based on the current state
-        the agent is in. In our case the agents chooses the
-
+        the agent is in.
     """
 
     def __init__(self, envs):
@@ -53,26 +52,37 @@ class Agent(nn.Module):
         self.critic = nn.Sequential(
             init_weightsNbias(
                 nn.Linear(
-                    np.array(envs.single_observation_space["image"].shape).prod(), 64
+                    np.array(
+                        np.array(envs.single_observation_space["image"].shape)
+                    ).prod(),
+                    64,
                 )
             ),
-            nn.Tanh(),
-            init_weightsNbias(nn.Linear(64, 64)),
-            nn.Tanh(),
+            # nn.ReLU(),
+            # init_weightsNbias(nn.Linear(64, 64)),
+            # nn.Tanh(),
+            nn.ReLU(),
             init_weightsNbias(nn.Linear(64, 1), std=1.0),
[…]

### 2023-10-17 src/PPO/PP_Optimization.py

@@ -9,114 +9,15 @@ import gymnasium as gym
 import torch.nn as nn
 import torch.optim as optim
 from torch.distributions.categorical import Categorical
+from ActorCritic import *
+from utils import *
 import json
-
-
-def read_params(params):
-    file = open(params)
-    params = json.load(file)
-    # print(json.dumps(params, indent=4, separators=(":", ",")))
-    return params
-
-
-def save_params(params):
-    file = open("hyperparams.json", "w")
-    json.dump(params, file, indent=4, separators=(",", ":"))
 
 
 def minibatch_generator():
     batch_idx = np.random.choice(args["batch_size"], args["batch_size"], replace=False)
     for start in range(0, args["batch_size"], args["minibatch_size"]):
         end = start + args["minibatch_size"]
-
-
-def init_weightsNbias(layer, std=np.sqrt(2), bias_const=0.0):
-    """
-    Descr:
-        Most, if not all implementations of PPO seen so far use the approach
-        of initializing weights orthogonaly. According to both the authors behind
-        Proximal Policy Optimization, and the paper by title "Exact solutions to
-        the nonlinear dynamics of learning in deep learning networks", the approach
-        of creating such initial conditions leads to faithful propagation of gradients,
-        and better convergence (See page 2 of the latter paper).
-    """
-    torch.nn.init.orthogonal_(layer.weight, std)
-    torch.nn.init.constant_(layer.bias, bias_const)
-    return layer
-
-
-def make_env(env_name, seed):
-    """
-    Descr:
-        Returns a "vectorized" environenment, meaning that it is wrapped in a gymnasium vector, allowing for
-        some aditional functionality, such as parallel training of many environments.
-    """
-
-    def env_gen():
-        env = gym.make(env_name, render_mode="rgb_array")
-        env = gym.wrappers.RecordEpisodeStatistics(env)
-        env.action_space.seed(seed)
-        env.observation_space.seed(seed)
-        return env
-
-    return env_gen
-
-
-# TODO: Transfer model to device a.k.a. to GPU
-class Agent(nn.Module):
-    """
-    Descr:
-        The agents task is to decide the next action to perform, and evaluate the
-        possible future rewards, a.k.a. the value function, based on the current state
-        the agent is in.
-    """
-
-    def __init__(self, envs):
-        super(Agent, self).__init__()
-        self.critic = nn.Sequential(
-            init_weightsNbias(
[…]

### 2023-10-17 src/PPO/utils.py

@@ -1,0 +1,30 @@
+import json
+import gymnasium as gym 
+
+def make_env(env_name, seed):
+    """
+    Descr:
+        Returns a "vectorized" environenment, meaning that it is wrapped in a gymnasium vector, allowing for
+        some aditional functionality, such as parallel training of many environments.
+    """
+
+    def env_gen():
+        env = gym.make(env_name, render_mode="rgb_array")
+        env = gym.wrappers.RecordEpisodeStatistics(env)
+        env.action_space.seed(seed)
+        env.observation_space.seed(seed)
+        return env
+
+    return env_gen
+
+
+def read_params(params):
+    file = open(params)
+    params = json.load(file)
+    # print(json.dumps(params, indent=4, separators=(":", ",")))
+    return params
+
+
+def save_params(params):
+    file = open("hyperparams.json", "w")
+    json.dump(params, file, indent=4, separators=(",", ":"))

### 2023-10-17 src/PPO/utils.py
minibatch-geneartor

@@ -1,5 +1,6 @@
 import json
-import gymnasium as gym 
+import gymnasium as gym
+
 
 def make_env(env_name, seed):
     """
@@ -28,3 +29,28 @@ def read_params(params):
 def save_params(params):
     file = open("hyperparams.json", "w")
     json.dump(params, file, indent=4, separators=(",", ":"))
+
+
+def minibatch_generator(params, *args):
+    b_obs = observations.view((-1, single_obs_shape))
+    b_logs = logprobs.view(-1)
+    b_acts = ations.view((-1) + envs.single_action_space.shape)
+    b_advs = advantages.view(-1)
+    b_vals = values.view(-1)
+    b_rets = returns.view(-1)
+
+    batch_idxs = np.random.choice(
+        params["batch_size"], params["batch_size"], replace=False
+    )
+    for start in range(0, params["batch_size"], params["minibatch_size"]):
+        end = start + params["minibatch_size"]
+        minibatch_idxs = batch_idxs[start:end]
+
+        yield (
+            b_obs[minibatch_idxs],
+            b_logs[minibatch_idxs],
+            b_acts[minibatch_idxs],
+            b_advs[minibatch_idxs],
+            b_vals[minibatch_idxs],
+            b_rets[minibatch_idxs],
+        )

### 2023-10-17 src/PPO/PP_Optimization.py
Update PPO so that it doesn't work
Signed-off-by: Jonatan Hanssen <jonatan.hanssen@outlook.com>

@@ -3,7 +3,6 @@ import os
 from distutils.util import strtobool
 import time
 import numpy as np
-import random
 import torch
 import gymnasium as gym
 import torch.nn as nn
@@ -14,290 +13,251 @@ from utils import *
 import json
 
 
-def minibatch_generator():
-    batch_idx = np.random.choice(args["batch_size"], args["batch_size"], replace=False)
-    for start in range(0, args["batch_size"], args["minibatch_size"]):
-        end = start + args["minibatch_size"]
+class PPO:
+    def __init__(self, param_file):
+        self.args = read_params(param_file)
+        seeding(self.args["seed"])
 
+        seeds = [random.randint(0, 20000) for _ in range(self.args["num_envs"])]
 
-def PPO(param_path, device=torch.device("cpu")):
-    args = read_params(param_path)
-    args["batch_size"] = int(args["num_envs"] * args["ep_steps"])
-    args["minibatch_size"] = int(args["batch_size"] // args["num_minibatches"])
-    seeding(args["seed"])
-    seeds = [random.randint(0, 20000) for _ in range(args["num_envs"])]
-    envs = gym.vector.SyncVectorEnv(
-        [make_env(args["env_name"], seeds[i]) for i in range(args["num_envs"])]
-    )
-    agent = Agent(envs).to(device)
-    optimizer = optim.Adam(agent.parameters(), lr=args["lr"], eps=1e-5)
-    # _, _ = envs.reset(seed=args["seed"])
-    args["rollouts"] = int(args["tot_steps"] // args["batch_size"])
-
-    single_obs_shape = int(
-        np.array(envs.single_observation_space["image"].shape).prod()
-    )
-    # print(envs.single_observation_space)
-
-    # The following tensors need not to be initialized to again
-    observations = torch.zeros(
-        (args["ep_steps"], args["num_envs"], single_obs_shape),
-    ).to(device)
-    actions = torch.zeros(
-        (args["ep_steps"], args["num_envs"]) + envs.single_action_space.shape
-    ).to(device)
-    # The ActorCritic Network outputs log probabilities
-    logprobs = torch.zeros((args["ep_steps"], args["num_envs"])).to(device)
-    rewards = torch.zeros((args["ep_steps"], args["num_envs"])).to(device)
-    dones = torch.zeros((args["ep_steps"], args["num_envs"])).to(device)
-    values = torch.zeros((args["ep_steps"], args["num_envs"])).to(device)
-
[…]

### 2023-10-17 src/PPO/utils.py

@@ -1,5 +1,8 @@
 import json
 import gymnasium as gym
+import random
+import numpy as np
+import torch
 
 
 def make_env(env_name, seed):
@@ -31,26 +34,8 @@ def save_params(params):
     json.dump(params, file, indent=4, separators=(",", ":"))
 
 
-def minibatch_generator(params, *args):
-    b_obs = observations.view((-1, single_obs_shape))
-    b_logs = logprobs.view(-1)
-    b_acts = ations.view((-1) + envs.single_action_space.shape)
-    b_advs = advantages.view(-1)
-    b_vals = values.view(-1)
-    b_rets = returns.view(-1)
-
-    batch_idxs = np.random.choice(
-        params["batch_size"], params["batch_size"], replace=False
-    )
-    for start in range(0, params["batch_size"], params["minibatch_size"]):
-        end = start + params["minibatch_size"]
-        minibatch_idxs = batch_idxs[start:end]
-
-        yield (
-            b_obs[minibatch_idxs],
-            b_logs[minibatch_idxs],
-            b_acts[minibatch_idxs],
-            b_advs[minibatch_idxs],
-            b_vals[minibatch_idxs],
-            b_rets[minibatch_idxs],
-        )
+def seeding(seed):
+    random.seed(seed)
+    np.random.seed(seed)
+    torch.manual_seed(seed)
+    torch.backends.cudnn.deterministic = True

### 2023-10-17 src/PPO/PP_Optimization.py
The PP started to work

@@ -27,11 +27,11 @@ class PPO:
             ]
         )
 
-        self.agent = Agent(self.envs)
+        self.device = torch.device("cuda" if torch.cuda.is_available() else "cpu")
+        self.agent = Agent(self.envs).to(self.device)
         self.optimizer = optim.Adam(
             self.agent.parameters(), lr=self.args["lr"], eps=1e-5
         )
-        self.device = torch.device("cuda" if torch.cuda.is_available() else "cpu")
         self.single_obs_shape = int(
             np.array(self.envs.single_observation_space["image"].shape).prod()
         )
@@ -88,13 +88,6 @@ class PPO:
         """
         for step in range(self.args["ep_steps"]):
             # global_step += args["num_envs"]
-            # if rollout_num == 0:
-            # print(envs.reset()[0]["image"].flatten())
-            # next_observation = torch.Tensor(envs.reset()[0]["image"].flatten()).to(
-            #     device
-            # )
-            # next_done = torch.zeros(args["num_envs"]).to(device)
-            # print(next_observation)
             self.observations[step] = next_observation
             self.dones[step] = next_done
 
@@ -102,7 +95,6 @@ class PPO:
                 action, logprob, _, value = self.agent.get_action_and_value(
                     next_observation
                 )
-                # print(np.array([action.cpu().numpy()]))
                 self.values[step] = value.flatten()
             self.actions[step] = action
             self.logprobs[step] = logprob
@@ -176,12 +168,19 @@ class PPO:
         When next_episode is called, this will have new values because the self values will be changed
         """
         b_obs = self.observations.view((-1, self.single_obs_shape))
-        b_logs = self.logprobs.view(-1)
-        print(self.envs.single_action_space.shape)
-        b_acts = self.actions.view((-1, self.envs.single_action_space))
-        b_advs = self.advantages.view(-1)
-        b_vals = self.values.view(-1)
-        b_rets = self.returns.view(-1)
+        b_logs = self.logprobs.view(
+            -1,
+        )
+        b_acts = self.actions.view((-1,) + self.envs.single_action_space.shape)
+        b_advs = self.advantages.view(
+            -1,
+        )
+        b_vals = self.values.view(
+            -1,
+        )
+        b_rets = self.returns.view(
+            -1,
+        )
 
         batch_idxs = np.random.choice(
             self.args["batch_size"], self.args["batch_size"], replace=False
@@ -252,9 +251,9 @@ class PPO:
 
[…]

### 2023-10-17 doc/paper.tex
Fix spelling error
Signed-off-by: Jonatan Hanssen <jonatan.hanssen@outlook.com>
add ethics statement

@@
 
-A standard PPO algorithm with no LLM integration, as a baseline. In this case, use StableBaselines' implementation with the following hyperparameters: HYPERPARAMETERS HERE.
+A standard PPO algorithm with no LLM integration, as a baseline. In this case, we use StableBaselines' implementation with the following hyperparameters: HYPERPARAMETERS HERE.
 
@@
 These values are joined with the probability values that the normal PPO algorithm outputs and normalized, and a new probability distribution is created, where semantically similar actions are weighted higher than others.
+We call this the Actor-Advisor-Critic Model.
 
@@
 
+\section*{Ethics Statement}
+
+In accordance with the NeurIPS Code of Ethics ***ADD REFERENCE HERE***, we outline ethical considerations related to this project, covering potential consequences from the research, and how potential risks may be mitigated.
+Considering that our research does not involve participants nor sensitive data, concerns related to how such factors are treated will not be discussed.
+
+\subsection{Societal Impact and Potential Harmful Consequences}
+
+Due to how this research project revolves around the integration of LLMs in the RL domain, many of the same concerns that exist in the domain of LLMs themselves are extended to the RL domain.
+A major concern is that of safety; if the RL agent is to act in an environment where failure to act optimally could be critical, it is paramount that the agent does not enter bad states due to the existing randomness in the RL framework.
+The LLM in itself predicts stochastically, and thus may add even more noise to the framework.
+In the worst case, it could amplify probabilities of the agent choosing dangerous actions due to this added noise.
+
+Another topic of concern is that of bias and fairness; the premise for why an LLM might help an RL agent in a large environment is because of its inherent human knowledge, stemming from the human-written texts it has been trained upon.
+However, human-written texts are an obvious potential source to biased opinions, and the LLM may have learned such biased and/or unfair opinions.
+Then, in an example where the RL agent is deployed in environments where an ethical choice has to be made, this choice could be influenced by whatever bias may be present in what the LLM suggests to the agent.
+
+\subsection{Impact Mitigation Measures}
+
+As is exemplified above, major ethical concerns regarding the use of our proposed algorithms arise when they are deployed for critical tasks, for instance tasks in real-life dangerous environments, or environments where ethical considerations have to be addressed.
+Therefore, an agent trained on our proposed RL framework for such critical tasks should be carefully tested before deployment to see if, and how often, critical deviations occur.
+
+In the further development of this and similar algorithms, we also consider it essential to put effort into trying to understand the causality behind certain actions being taken, as is being done in the field of explainable artificial intelligence (XAI).
+When involving one major ML framework like that of LLMs into RL, we would ideally want to see exactly where and how much the LLM contributes to the RL networks being updated.
+This would, in turn, help in the work to mitigate whatever biases and variance that may be added to the RL framework from integrating the LLM.
+
 \section*{Acknowledgment}

### 2023-10-17 src/PPO/ActorCritic.py
Controlled code for PPO

@@ -1,4 +1,5 @@
 import os
+
 # from torch.utils.tensorboard import SummaryWriter
 import numpy as np
 import torch
@@ -58,10 +59,9 @@ class Agent(nn.Module):
                     64,
                 )
             ),
-            # nn.ReLU(),
-            # init_weightsNbias(nn.Linear(64, 64)),
-            # nn.Tanh(),
-            nn.ReLU(),
+            nn.Tanh(),
+            init_weightsNbias(nn.Linear(64, 64)),
+            nn.Tanh(),
             init_weightsNbias(nn.Linear(64, 1), std=1.0),
         )
 
@@ -74,10 +74,9 @@ class Agent(nn.Module):
                     64,
                 )
             ),
-            # nn.ReLU(),
-            # init_weightsNbias(nn.Linear(64, 64)),
-            # nn.Tanh(),
-            nn.ReLU(),
+            nn.Tanh(),
+            init_weightsNbias(nn.Linear(64, 64)),
+            nn.Tanh(),
             init_weightsNbias(nn.Linear(64, envs.single_action_space.n), std=0.01),
         )
 
@@ -93,5 +92,3 @@ class Agent(nn.Module):
         if action is None:
             action = probs.sample()
         return action, probs.log_prob(action), probs.entropy(), self.critic(X)
-
-

### 2023-10-18 src/PPO/ActorCritic.py
Update ppo stuff
Signed-off-by: Jonatan Hanssen <jonatan.hanssen@outlook.com>

@@ -1,4 +1,5 @@
 import os
+
 # from torch.utils.tensorboard import SummaryWriter
 import numpy as np
 import torch
@@ -23,12 +24,12 @@ def init_weightsNbias(layer, std=np.sqrt(2), bias_const=0.0):
 
 
 def save_model(model):
-    file_path = os.path.dirname(__file__) + "/model.zip"
+    file_path = os.path.dirname(__file__) + "/model.pt"
     torch.save(model.state_dict(), file_path)
 
 
 def load_model(model):
-    file_path = os.path.dirname(__file__) + "/model.zip"
+    file_path = os.path.dirname(__file__) + "/model.pt"
     model.load_state_dict(torch.load(file_path))
 
 
@@ -58,8 +59,10 @@ class Agent(nn.Module):
                     64,
                 )
             ),
-            # nn.ReLU(),
-            # init_weightsNbias(nn.Linear(64, 64)),
+            nn.ReLU(),
+            init_weightsNbias(nn.Linear(64, 64)),
+            nn.ReLU(),
+            init_weightsNbias(nn.Linear(64, 64)),
             # nn.Tanh(),
             nn.ReLU(),
             init_weightsNbias(nn.Linear(64, 1), std=1.0),
@@ -74,6 +77,10 @@ class Agent(nn.Module):
                     64,
                 )
             ),
+            nn.ReLU(),
+            init_weightsNbias(nn.Linear(64, 64)),
+            nn.ReLU(),
+            init_weightsNbias(nn.Linear(64, 64)),
             # nn.ReLU(),
             # init_weightsNbias(nn.Linear(64, 64)),
             # nn.Tanh(),
@@ -81,8 +88,10 @@ class Agent(nn.Module):
             init_weightsNbias(nn.Linear(64, envs.single_action_space.n), std=0.01),
         )
 
+        # self.llama_actor = llama2_7b_policy()
+
         # Yet to be integrated -> shall serve as the second actor
-        # .LM_actor = llama2_7b_policy()
+        # self.LM_actor = llama2_7b_policy()
 
     def get_value(self, X):
         return self.critic(X)
@@ -90,8 +99,13 @@ class Agent(nn.Module):
     def get_action_and_value(self, X, action=None):
         logits = self.actor(X)
         probs = Categorical(logits=logits)
+
+        # if self.i_love_llama:
+        #     if self.i_love_adrian:
+        #         probs *= self.llama_actor()
+        #     elif self.i_hate_adrian:
[…]

### 2023-10-18 src/PPO/utils.py

@@ -5,7 +5,7 @@ import numpy as np
 import torch
 
 
-def make_env(env_name, seed):
+def make_env(env_name, seed, render_mode="rgb_array"):
     """
     Descr:
         Returns a "vectorized" environenment, meaning that it is wrapped in a gymnasium vector, allowing for
@@ -13,7 +13,7 @@ def make_env(env_name, seed):
     """
 
     def env_gen():
-        env = gym.make(env_name, render_mode="rgb_array")
+        env = gym.make(env_name, render_mode=render_mode)
         env = gym.wrappers.RecordEpisodeStatistics(env)
         env.action_space.seed(seed)
         env.observation_space.seed(seed)

### 2023-10-18 src/PPO/PP_Optimization.py
Update ppo code, prepare for merge conflicts
Signed-off-by: Jonatan Hanssen <jonatan.hanssen@outlook.com>

@@ -83,36 +83,41 @@ class PPO:
             print(f"Sum of rewards per episode: {torch.sum(self.rewards)}")
             self.PPO_update()
 
-    def next_episode(self, next_done, next_observation):
-        """Steps through a single episode and calculates what is needed to
-        perform PPO update
-        """
-        for step in range(self.args["ep_steps"]):
-            # global_step += args["num_envs"]
-            self.observations[step] = next_observation
-            self.dones[step] = next_done
+    def show_loaded_model(self):
+        load_model(self.agent)
+        seeds = [random.randint(0, 20000) for _ in range(self.args["num_envs"])]
 
+        envs = gym.vector.SyncVectorEnv(
+            [
+                make_env(self.args["env_name"], seeds[i], "human")
+                for i in range(self.args["num_envs"])
+            ]
+        )
+
+        next_observation = torch.Tensor(envs.reset()[0]["image"].flatten()).to(
+            self.device
+        )
+        next_done = torch.zeros(self.args["num_envs"]).to(self.device)
+
+        for step in range(1000):
             with torch.no_grad():
                 action, logprob, _, value = self.agent.get_action_and_value(
                     next_observation
                 )
-                self.values[step] = value.flatten()
-            self.actions[step] = action
-            self.logprobs[step] = logprob
 
-            next_observation, reward, done, truncated, info = self.envs.step(
-                np.array([action.cpu().numpy()])
-            )
-            # Converting to tensor and transfering to gpu for faster computation
-            self.rewards[step] = torch.Tensor(reward).to(self.device).view(-1)
+                next_observation, reward, done, truncated, info = envs.step(
+                    np.array([action.cpu().numpy()])
+                )
+                # Converting to tensor and transfering to gpu for faster computation
 
-            # print(next_observation["image"])
+                # print(next_observation["image"])
 
-            next_observation, next_done = (
-                torch.Tensor(next_observation["image"].flatten()).to(self.device),
-                torch.Tensor(next_done).to(self.device),
-            )
+                next_observation, next_done = (
+                    torch.Tensor(next_observation["image"].flatten()).to(self.device),
+                    torch.Tensor(next_done).to(self.device),
+                )
 
+<<<<<<< Updated upstream
         # Now that the agent(-s) has(-ve) played out an episode, it's time
         # to backtrack all steps, and compute the discounted rewards
         with torch.no_grad():
@@ -131,12 +136,39 @@ class PPO:
[…]

### 2023-10-18 src/PPO/PP_Optimization.py
Cant be bothered to fix conflicts
Just accept previous version

Signed-off-by: Jonatan Hanssen <jonatan.hanssen@outlook.com>

@@ -83,41 +83,36 @@ class PPO:
             print(f"Sum of rewards per episode: {torch.sum(self.rewards)}")
             self.PPO_update()
 
-    def show_loaded_model(self):
-        load_model(self.agent)
-        seeds = [random.randint(0, 20000) for _ in range(self.args["num_envs"])]
+    def next_episode(self, next_done, next_observation):
+        """Steps through a single episode and calculates what is needed to
+        perform PPO update
+        """
+        for step in range(self.args["ep_steps"]):
+            # global_step += args["num_envs"]
+            self.observations[step] = next_observation
+            self.dones[step] = next_done
 
-        envs = gym.vector.SyncVectorEnv(
-            [
-                make_env(self.args["env_name"], seeds[i], "human")
-                for i in range(self.args["num_envs"])
-            ]
-        )
-
-        next_observation = torch.Tensor(envs.reset()[0]["image"].flatten()).to(
-            self.device
-        )
-        next_done = torch.zeros(self.args["num_envs"]).to(self.device)
-
-        for step in range(1000):
             with torch.no_grad():
                 action, logprob, _, value = self.agent.get_action_and_value(
                     next_observation
                 )
+                self.values[step] = value.flatten()
+            self.actions[step] = action
+            self.logprobs[step] = logprob
 
-                next_observation, reward, done, truncated, info = envs.step(
-                    np.array([action.cpu().numpy()])
-                )
-                # Converting to tensor and transfering to gpu for faster computation
+            next_observation, reward, done, truncated, info = self.envs.step(
+                np.array([action.cpu().numpy()])
+            )
+            # Converting to tensor and transfering to gpu for faster computation
+            self.rewards[step] = torch.Tensor(reward).to(self.device).view(-1)
 
-                # print(next_observation["image"])
+            # print(next_observation["image"])
 
-                next_observation, next_done = (
-                    torch.Tensor(next_observation["image"].flatten()).to(self.device),
-                    torch.Tensor(next_done).to(self.device),
-                )
+            next_observation, next_done = (
+                torch.Tensor(next_observation["image"].flatten()).to(self.device),
+                torch.Tensor(next_done).to(self.device),
+            )
 
-<<<<<<< Updated upstream
         # Now that the agent(-s) has(-ve) played out an episode, it's time
         # to backtrack all steps, and compute the discounted rewards
         with torch.no_grad():
@@ -136,39 +131,12 @@ class PPO:
[…]

### 2023-10-18 src/PPO/ActorCritic.py
Fix ppo code so that its but still doesnt work
Signed-off-by: Jonatan Hanssen <jonatan.hanssen@outlook.com>

@@ -48,14 +48,12 @@ class Agent(nn.Module):
         the agent is in.
     """
 
-    def __init__(self, envs):
+    def __init__(self, env):
         super(Agent, self).__init__()
         self.critic = nn.Sequential(
             init_weightsNbias(
                 nn.Linear(
-                    np.array(
-                        np.array(envs.single_observation_space["image"].shape)
-                    ).prod(),
+                    np.array(np.array(env.observation_space["image"].shape)).prod(),
                     64,
                 )
             ),
@@ -68,16 +66,14 @@ class Agent(nn.Module):
         self.actor = nn.Sequential(
             init_weightsNbias(
                 nn.Linear(
-                    np.array(
-                        np.array(envs.single_observation_space["image"].shape)
-                    ).prod(),
+                    np.array(np.array(env.observation_space["image"].shape)).prod(),
                     64,
                 )
             ),
             nn.Tanh(),
             init_weightsNbias(nn.Linear(64, 64)),
             nn.Tanh(),
-            init_weightsNbias(nn.Linear(64, envs.single_action_space.n), std=0.01),
+            init_weightsNbias(nn.Linear(64, env.action_space.n), std=0.01),
         )
 
         # self.llama_actor = llama2_7b_policy()

### 2023-10-18 src/PPO/PP_Optimization.py

@@ -18,110 +18,79 @@ class PPO:
         self.args = read_params(param_file)
         seeding(self.args["seed"])
 
-        seeds = [random.randint(0, 20000) for _ in range(self.args["num_envs"])]
-
-        self.envs = gym.vector.SyncVectorEnv(
-            [
-                make_env(self.args["env_name"], seeds[i])
-                for i in range(self.args["num_envs"])
-            ]
+        self.env = gym.make(
+            self.args["env_name"], render_mode="rgb_array", max_steps=self.args["steps"]
         )
 
         self.device = torch.device("cuda" if torch.cuda.is_available() else "cpu")
-        self.agent = Agent(self.envs).to(self.device)
+        self.agent = Agent(self.env).to(self.device)
         self.optimizer = optim.Adam(
             self.agent.parameters(), lr=self.args["lr"], eps=1e-8
         )
-        self.single_obs_shape = int(
-            np.array(self.envs.single_observation_space["image"].shape).prod()
-        )
-        # print(envs.single_observation_space)
+        self.obs_shape = int(np.array(self.env.observation_space["image"].shape).prod())
 
-        # The following tensors need not to be initialized to again
-        self.observations = torch.zeros(
-            (self.args["ep_steps"], self.args["num_envs"], self.single_obs_shape),
-        ).to(self.device)
-        self.actions = torch.zeros(
-            (self.args["ep_steps"], self.args["num_envs"])
-            + self.envs.single_action_space.shape
-        ).to(self.device)
-        # The ActorCritic Network outputs log probabilities
-        self.logprobs = torch.zeros((self.args["ep_steps"], self.args["num_envs"])).to(
-            self.device
-        )
-        self.rewards = torch.zeros_like(self.logprobs).to(self.device)
-        self.dones = torch.zeros_like(self.logprobs).to(self.device)
-        self.values = torch.zeros_like(self.logprobs).to(self.device)
-        self.advantages = torch.zeros_like(self.logprobs).to(self.device)
-        self.returns = torch.zeros_like(self.logprobs).to(self.device)
+        # Initialize all observation-, return-, valuevectors and so on
+        self.reset_memory()
 
     def do_it(self):
-        self.args["batch_size"] = int(self.args["num_envs"] * self.args["ep_steps"])
+        self.args["batch_size"] = self.args["steps"]
         self.args["minibatch_size"] = int(
[…]

### 2023-10-18 src/PPO/ActorCritic.py
Add more to PPO
But it still doesnt work

Signed-off-by: Jonatan Hanssen <jonatan.hanssen@outlook.com>

@@ -51,29 +51,21 @@ class Agent(nn.Module):
     def __init__(self, env):
         super(Agent, self).__init__()
         self.critic = nn.Sequential(
-            init_weightsNbias(
-                nn.Linear(
-                    np.array(np.array(env.observation_space["image"].shape)).prod(),
-                    64,
-                )
-            ),
+            init_weightsNbias(nn.Linear(147 * 11, 64, dtype=torch.float64)),
             nn.Tanh(),
-            init_weightsNbias(nn.Linear(64, 64)),
+            init_weightsNbias(nn.Linear(64, 64, dtype=torch.float64)),
             nn.Tanh(),
-            init_weightsNbias(nn.Linear(64, 1), std=1.0),
+            init_weightsNbias(nn.Linear(64, 1, dtype=torch.float64), std=1.0),
         )
 
         self.actor = nn.Sequential(
+            init_weightsNbias(nn.Linear(147 * 11, 64, dtype=torch.float64)),
+            nn.Tanh(),
+            init_weightsNbias(nn.Linear(64, 64, dtype=torch.float64)),
+            nn.Tanh(),
             init_weightsNbias(
-                nn.Linear(
-                    np.array(np.array(env.observation_space["image"].shape)).prod(),
-                    64,
-                )
+                nn.Linear(64, env.action_space.n, dtype=torch.float64), std=0.01
             ),
-            nn.Tanh(),
-            init_weightsNbias(nn.Linear(64, 64)),
-            nn.Tanh(),
-            init_weightsNbias(nn.Linear(64, env.action_space.n), std=0.01),
         )
 
         # self.llama_actor = llama2_7b_policy()
@@ -81,19 +73,27 @@ class Agent(nn.Module):
         # Yet to be integrated -> shall serve as the second actor
         # self.LM_actor = llama2_7b_policy()
 
-    def get_value(self, X):
-        return self.critic(X)
+    def get_value(self, observation):
+        observation = nn.functional.one_hot(
+            observation.to(torch.int64), num_classes=11
+        ).flatten()
+        return self.critic(observation.to(torch.float64))
 
-    def get_action_and_value(self, X, action=None):
-        logits = self.actor(X)
+    def get_action_and_value(self, observation, action=None):
+        observation = (
+            nn.functional.one_hot(observation.to(torch.int64), num_classes=11)
+            .to(torch.float64)
+            .flatten(start_dim=-2)
+        )
+        logits = self.actor(observation)
         probs = Categorical(logits=logits)
 
[…]

### 2023-10-18 src/PPO/PP_Optimization.py

@@ -32,7 +32,11 @@ class PPO:
         # Initialize all observation-, return-, valuevectors and so on
         self.reset_memory()
 
-    def do_it(self):
+    def train(self):
+        """Generates samples from a single episode, backpropagates the samples and
+        repeats for rollout number of times
+        """
+
         self.args["batch_size"] = self.args["steps"]
         self.args["minibatch_size"] = int(
             self.args["batch_size"] // self.args["num_minibatches"]
@@ -48,6 +52,7 @@ class PPO:
             print(f"Sum of rewards per episode: {torch.sum(self.rewards)}")
             self.PPO_update()  # Use values stored to backpropagate using PPO
             self.reset_memory()  # Restore
+        save_model(self.agent)
 
     def next_episode(self):
         """Steps through a single episode and calculates what is needed to
@@ -93,24 +98,24 @@ class PPO:
                     if t == self.args["steps"] - 1:
                         # "nextnonterminal" is a environment specific variable indicating if the
                         # agent has finished the game before reaching time step limit
-                        nextnonterminal = 1.0 - next_done
+                        # nextnonterminal = 1.0 - next_done
                         next_values = next_value
                         last_gae_lam = 0
                     else:
-                        nextnonterminal = 1.0 - self.dones[t + 1]
+                        # nextnonterminal = 1.0 - self.dones[t + 1]
                         next_values = self.values[t + 1]
                         last_gae_lam = self.returns[t + 1]
 
                     delta = (
                         self.rewards[t]
-                        + self.args["gamma"] * next_values * nextnonterminal
-                        - self.values[t]
+                        + self.args["gamma"]
+                        * next_values
+                        * -self.values[t]  # nextnonterminal
                     )
                     self.advantages[t] = (
                         delta
-                        + self.args["gamma"]
-                        * self.args["gae_lambda"]
-                        * nextnonterminal
+                        + self.args["gamma"] * self.args["gae_lambda"]
+                        # * nextnonterminal
                         * last_gae_lam
                     )
                 self.returns = self.advantages + self.values
@@ -123,44 +128,6 @@ class PPO:
                     )
 
                 self.advantages = self.returns - self.values
-
-    def reset_memory(self):
[…]

### 2023-10-19 src/PPO/utils.py
Move ppo files into main src directory
This untracks the entire PPO directory

Signed-off-by: Jonatan Hanssen <jonatan.hanssen@outlook.com>

@@ -1,41 +1,0 @@
-import json
-import gymnasium as gym
-import random
-import numpy as np
-import torch
-
-
-def make_env(env_name, seed, render_mode="rgb_array"):
-    """
-    Descr:
-        Returns a "vectorized" environenment, meaning that it is wrapped in a gymnasium vector, allowing for
-        some aditional functionality, such as parallel training of many environments.
-    """
-
-    def env_gen():
-        env = gym.make(env_name, render_mode=render_mode)
-        env = gym.wrappers.RecordEpisodeStatistics(env)
-        env.action_space.seed(seed)
-        env.observation_space.seed(seed)
-        return env
-
-    return env_gen
-
-
-def read_params(params):
-    file = open(params)
-    params = json.load(file)
-    # print(json.dumps(params, indent=4, separators=(":", ",")))
-    return params
-
-
-def save_params(params):
-    file = open("hyperparams.json", "w")
-    json.dump(params, file, indent=4, separators=(",", ":"))
-
-
-def seeding(seed):
-    random.seed(seed)
-    np.random.seed(seed)
-    torch.manual_seed(seed)
-    torch.backends.cudnn.deterministic = True

### 2023-10-19 src/utils.py

@@ -1,7 +1,8 @@
 import constants
+import torch
 import numpy as np
 from llama import Llama
-import os, time, random
+import os, time, random, json
 from sentence_transformers import SentenceTransformer
 
 
@@ -373,3 +374,22 @@ def obs_to_string(obs_matrix, positions=True, you=True):
     # Please only answer with a single of the following commands: RIGHT, LEFT, FORWARD or PICK UP."
 
     return observation
+
+
+def seeding(seed):
+    random.seed(seed)
+    np.random.seed(seed)
+    torch.manual_seed(seed)
+    torch.backends.cudnn.deterministic = True
+
+
+def read_params(params):
+    file = open(params)
+    params = json.load(file)
+    # print(json.dumps(params, indent=4, separators=(":", ",")))
+    return params
+
+
+def save_params(params):
+    file = open("hyperparams.json", "w")
+    json.dump(params, file, indent=4, separators=(",", ":"))

### 2023-10-19 src/SB3_PPO_MiniGrid/feature_extractor.py
Remove SB3_PPO
Signed-off-by: Jonatan Hanssen <jonatan.hanssen@outlook.com>

@@ -1,63 +1,0 @@
-import torch
-import torch.nn as nn
-from gymnasium import spaces
-import gymnasium as gym
-
-from stable_baselines3 import PPO
-from stable_baselines3.common.torch_layers import BaseFeaturesExtractor
-from stable_baselines3.common.env_util import make_vec_env
-from minigrid.wrappers import ImgObsWrapper
-from stable_baselines3.common.vec_env import DummyVecEnv
-
-class CustomCNN(BaseFeaturesExtractor):
-    """
-    :param observation_space: (gym.Space)
-    :param features_dim: (int) Number of features extracted.
-        This corresponds to torche number of unit for torche last layer.
-    """
-
-    def __init__(self, observation_space: spaces.Box, features_dim: int = 128):
-        super().__init__(observation_space, features_dim)
-        # We assume CxHxW images (channels first)
-        # Re-ordering will be done by pre-preprocessing or wrapper
-        n_input_channels = observation_space.shape[0]
-        self.cnn = nn.Sequential(
-            nn.Conv2d(n_input_channels, 16, (2,2)),
-            nn.Tanh(),
-            nn.Conv2d(16, 32, (2,2)),
-            nn.Tanh(),
-            # nn.Conv2d(32, 64, (2,2)),
-            # nn.ReLU(),
-            nn.Flatten(),
-        )
-
-        # Compute shape by doing one forward pass
-        with torch.no_grad():
-            n_flatten = self.cnn(
-                torch.as_tensor(observation_space.sample()[None]).float()
-            ).shape[1]
-
-        self.linear = nn.Sequential(nn.Linear(n_flatten, features_dim), nn.Tanh())
-
-    def forward(self, observations: torch.Tensor) -> torch.Tensor:
-        return self.linear(self.cnn(observations))
-
-
-policy_kwargs = dict(
-    features_extractor_class=CustomCNN,
-    features_extractor_kwargs=dict(features_dim=128),
-)
-# env = make_vec_env("MiniGrid-UnlockPickup-v0", n_envs=1)
-env = DummyVecEnv([lambda: ImgObsWrapper(gym.make("MiniGrid-UnlockPickup-v0", render_mode="rgb_array"))])
-env.seed(42)
-# env = ImgObsWrapper(env)
-
-model = PPO(
-    "CnnPolicy", env, policy_kwargs=policy_kwargs, verbose=1
-)
-model.learn(total_timesteps=2e4)
-model.save("ppo_minigrid")
-del model
-
-
-

### 2023-10-19 src/SB3_PPO_MiniGrid/ppo_minigrid.py

@@ -1,23 +1,0 @@
-import torch
-import torch.nn as nn 
-from gymnasium import spaces
-import gymnasium as gym 
-
-from stable_baselines3 import PPO 
-from stable_baselines3.common.torch_layers import BaseFeaturesExtractor
-from stable_baselines3.common.env_util import make_vec_env
-from minigrid.wrappers import ImgObsWrapper
-from stable_baselines3.common.vec_env import DummyVecEnv 
-
-env = DummyVecEnv([lambda: ImgObsWrapper(gym.make("MiniGrid-UnlockPickup-v0", render_mode="human"))])
-env.seed(42)
-# env = ImgObsWrapper(env)
-
-model = PPO.load("ppo_minigrid", env=env)
-
-obs = env.reset() 
-for _ in range(1000): 
-    action = model.predict(obs)
-    obs, reward, dones, info = env.step(action)
-
-env.close()

### 2023-10-19 src/ActorCritic.py
Add integration between PPO and LLM
Signed-off-by: Jonatan Hanssen <jonatan.hanssen@outlook.com>

@@ -6,6 +6,7 @@ import torch
 import torch.nn as nn
 import torch.optim as optim
 from torch.distributions.categorical import Categorical
+from utils import llama2_policy
 
 
 def init_weightsNbias(layer, std=np.sqrt(2), bias_const=0.0):
@@ -48,7 +49,7 @@ class Agent(nn.Module):
         the agent is in.
     """
 
-    def __init__(self, env):
+    def __init__(self, env, llama=False):
         super(Agent, self).__init__()
         self.critic = nn.Sequential(
             init_weightsNbias(nn.Linear(147 * 11, 64, dtype=torch.float64)),
@@ -68,7 +69,8 @@ class Agent(nn.Module):
             ),
         )
 
-        # self.llama_actor = llama2_7b_policy()
+        goal = env.reset()[0]["mission"]
+        self.consigliere = llama2_policy(goal) if llama else None
 
         # Yet to be integrated -> shall serve as the second actor
         # self.LM_actor = llama2_7b_policy()
@@ -80,20 +82,32 @@ class Agent(nn.Module):
         return self.critic(observation.to(torch.float64))
 
     def get_action_and_value(self, observation, action=None):
-        observation = (
+        observation_onehot = (
             nn.functional.one_hot(observation.to(torch.int64), num_classes=11)
             .to(torch.float64)
             .flatten(start_dim=-2)
         )
-        logits = self.actor(observation)
-        probs = Categorical(logits=logits)
+        logits = self.actor(observation_onehot)
 
-        # if self.i_love_llama:
-        #     if self.i_love_adrian:
-        #         probs *= self.llama_actor(X)
-        #     elif self.i_hate_adrian:
-        #         probs += self.llama_actor(X)
+        if self.consigliere:
+            logits = torch.nn.functional.softmax(logits)
+            # its flattened so we need to make it normal again
+            unflat_obs = np.array(observation.reshape((7, 7, 3)).to(torch.int64))
+            # this stores suggested actions
+            self.consigliere.suggest(unflat_obs)
+            # compares all possible actions to suggestions
+            advisor_values = self.consigliere.give_values(unflat_obs)
+
+            # anti adrian propaganda
+            logits += advisor_values
+
+        probs = Categorical(logits=logits)
 
         if action is None:
             action = probs.sample()
-        return action, probs.log_prob(action), probs.entropy(), self.critic(observation)
+        return (
+            action,
+            probs.log_prob(action),
+            probs.entropy(),
+            self.critic(observation_onehot),
+        )

### 2023-10-19 src/ppo.py

@@ -14,7 +14,7 @@ import json
 
 
 class PPO:
-    def __init__(self, param_file):
+    def __init__(self, param_file, llama_policy=False, llama_reward=False):
         self.args = read_params(param_file)
         seeding(self.args["seed"])
 
@@ -23,11 +23,17 @@ class PPO:
         )
 
         self.device = torch.device("cuda" if torch.cuda.is_available() else "cpu")
-        self.agent = Agent(self.env).to(self.device)
+        self.agent = Agent(self.env, llama_policy).to(self.device)
         self.optimizer = optim.Adam(
             self.agent.parameters(), lr=self.args["lr"], eps=1e-8
         )
         self.obs_shape = int(np.array(self.env.observation_space["image"].shape).prod())
+
+        self.reward_shaper = (
+            llama2_reward_shaper(self.env.reset()[0]["mission"])
+            if llama_reward
+            else None
+        )
 
         # Initialize all observation-, return-, valuevectors and so on
         self.reset_memory()
@@ -58,9 +64,8 @@ class PPO:
         """Steps through a single episode and calculates what is needed to
         perform PPO update
         """
-        observation = torch.Tensor(self.env.reset()[0]["image"].flatten()).to(
-            self.device
-        )
+        observation_dict, _ = self.env.reset()
+        observation = torch.Tensor(observation_dict["image"].flatten()).to(self.device)
 
         for step in range(self.args["steps"]):
             self.observations[step] = observation
@@ -72,13 +77,21 @@ class PPO:
             self.actions[step] = action
             self.logprobs[step] = logprob
 
+            # LLM reward shaping
+            if self.reward_shaper:
+                # this sets self.suggestions
+                self.reward_shaper.suggest(observation_dict["image"])
+                advisor_reward = self.reward_shaper.compare(
+                    action.cpu().numpy(), observation_dict["image"]
+                )
+            else:
+                advisor_reward = 0
+
             observation_dict, reward, done, truncated, info = self.env.step(
                 action.cpu().numpy()
             )
-            # Converting to tensor and transfering to gpu for faster computation
-            self.rewards[step] = reward
 
-            # print(next_observation["image"])
+            self.rewards[step] = reward + advisor_reward
 
             observation = torch.Tensor(observation_dict["image"].flatten()).to(
                 self.device
@@ -237,6 +250,6 @@ class PPO:
 
 
 if __name__ == "__main__":
[…]

### 2023-10-19 src/test_ppo.py

@@ -1,0 +1,5 @@
+from ppo import PPO
+
+ppo = PPO("hyperparams.json")
+
+ppo.show_loaded_model()

### 2023-10-19 src/utils.py

@@ -9,7 +9,7 @@ from sentence_transformers import SentenceTransformer
 base_path = os.path.dirname(__file__)
 
 
-class llama2_7b_reward_shaper:
+class llama2_base:
     """Reward shaper
 
     Takes in an observation matrix, and generates a list of suggestions. These can be compared with actions and reward will be generated
@@ -19,7 +19,7 @@ class llama2_7b_reward_shaper:
         self,
         goal,
         cos_sim_threshold=0.6,
-        reward_multiplier=0.001,
+        similarity_modifier=0.001,
         temperature=0.6,
         top_p=0.9,
         rl_temp=0,
@@ -89,12 +89,13 @@ class llama2_7b_reward_shaper:
         """Creates a list of suggested actions based on the current observation
 
         Args:
-            observation: the full observation returned by gymnasium environment
+            observation: image observation returned by gym environment
 
         Returns:
             A list of strings of suggested actions by the LLM. Also sets self.suggestions
 
         """
+        observation = obs_to_string(observation, False, False)
 
         observation = f"My goal is: {self.goal}. {observation}"
 
@@ -146,7 +147,7 @@ class llama2_7b_reward_shaper:
                 max_cos_sim = cos_sim
 
         if max_cos_sim > self.cos_sim_threshold:
-            return cos_sim * self.reward_multiplier
+            return cos_sim * self.similarity_modifier
         else:
             return 0
 
@@ -222,97 +223,41 @@ def caption_action(action, obs_matrix):
     return caption
 
 
-class llama2_7b_policy:
-    def __init__(self, temperature=0.6, top_p=0.9, rl_temp=0, dialogue_memory=24):
-        self.generator = Llama.build(
-            ckpt_dir=os.path.join(base_path, "../llama-2-7b-chat"),
-            tokenizer_path=os.path.join(
-                base_path, "../llama-2-7b-chat/tokenizer.model"
-            ),
-            max_seq_len=2048,
-            max_batch_size=6,
+class llama2_reward_shaper(llama2_base):
+    def __init__(
+        self,
+        goal,
+        cos_sim_threshold=0.6,
+        similarity_modifier=0.001,
+        temperature=0.6,
+        top_p=0.9,
+        rl_temp=0,
+    ):
+        super().__init__(
+            goal, cos_sim_threshold, similarity_modifier, temperature, top_p, rl_temp
         )
 
-        self.temperature = temperature
-        self.dialogue_memory = dialogue_memory
-        self.top_p = top_p
-        self.rl_temp = rl_temp
[…]

### 2023-10-19 startchat.sh
Make start shell script more smart
Just write ./start.sh <python script in src/>

Signed-off-by: Jonatan Hanssen <jonatan.hanssen@outlook.com>

@@ -1,1 +1,0 @@
-torchrun --nproc_per_node 1 src/chatdemo.py

### 2023-10-25 src/ppo.py
Some working hyperparams

@@ -229,6 +229,6 @@ class PPO:
 
 
 if __name__ == "__main__":
-    ppo = PPO("hyperparams.json", True, False)
+    ppo = PPO("hyperparams.json", False, False)
     # save_params(self.self.args)
     ppo.train()

### 2023-10-26 src/utils.py
Update ppo with cache and higher sim threshold
Signed-off-by: Jonatan Hanssen <jonatan.hanssen@outlook.com>

@@ -23,6 +23,7 @@ class llama2_base:
         temperature=0.6,
         top_p=0.9,
         rl_temp=0,
+        cache_file="llm_cache.json",
     ):
         self.generator = Llama.build(
             ckpt_dir=os.path.join(base_path, "../llama-2-7b-chat"),
@@ -44,7 +45,16 @@ class llama2_base:
         self.suggestions = None
         self.goal = goal
 
-        self.cache = dict()
+        self.cache_file = cache_file
+
+        if self.cache_file:
+            self.cache_file = os.path.join(base_path, cache_file)
+
+            if os.path.exists(self.cache_file):
+                with open(self.cache_file) as file:
+                    self.cache = json.load(file)
+        else:
+            self.cache = dict()
 
         self.caption_set = set()
 
@@ -130,6 +140,11 @@ class llama2_base:
     def reset_cache(self):
         self.cache = dict()
 
+    def save_cache(self):
+        if self.cache_file:
+            with open(self.cache_file, "w") as file:
+                json.dump(self.cache, file)
+
     def compare(self, action, obs_matrix):
         """Compares the semantic similarity between an action and the current list of suggested actions
 
@@ -161,11 +176,12 @@ class llama2_base:
                 best_suggestion = suggestion
 
         if max_cos_sim > self.cos_sim_threshold:
-            print(f"{max_cos_sim=}")
-            print(f"{best_suggestion=}")
-            print(f"{caption=}")
+            # print(f"{max_cos_sim=}")
+            # print(f"{caption=}")
+            # print(f"{best_suggestion=}")
+            # print()
 
-            return cos_sim * self.similarity_modifier
+            return max_cos_sim * self.similarity_modifier
         else:
             return 0
 
@@ -250,9 +266,10 @@ class llama2_reward_shaper(llama2_base):
         temperature=0.6,
         top_p=0.9,
         rl_temp=0,
+        cache_file=None,
     ):
         super().__init__(
-            goal, cos_sim_threshold, similarity_modifier, temperature, top_p, rl_temp
+            goal, cos_sim_threshold, similarity_modifier, temperature, top_p, rl_temp, cache_file
         )
 
 
@@ -265,9 +282,10 @@ class llama2_policy(llama2_base):
         temperature=0.6,
         top_p=0.9,
         rl_temp=0,
+        cache_file=None,
     ):
         super().__init__(
[…]

### 2023-10-26 src/plot_rewards.py
Add plotrewards
Signed-off-by: Jonatan Hanssen <jonatan.hanssen@outlook.com>

@@ -1,0 +1,29 @@
+import numpy as np
+import matplotlib.pyplot as plt
+import os
+import argparse
+
+
+parser = argparse.ArgumentParser()
+
+parser.add_argument("llm_file")
+parser.add_argument("ppo_file")
+args = parser.parse_args()
+
+abs_path = os.path.dirname(__file__)
+
+for arg in vars(args):
+    file_name = os.path.join(abs_path, getattr(args, arg))
+
+    data = np.load(file_name)
+
+    means = list()
+    for window in np.array_split(data, len(data) // 10):
+        means.append(np.mean(window))
+
+    plt.plot(
+        range(len(means)), means, label=f"{getattr(args, arg)}".replace(".npy", "")
+    )
+
+plt.legend()
+plt.show()

### 2023-10-26 src/utils.py
changes to llama policy

@@ -97,7 +97,7 @@ class llama2_base:
             },
         ]
 
-    def suggest(self, observation):
+    def suggest(self, observation, show_pos=False):
         """Creates a list of suggested actions based on the current observation
 
         Args:
@@ -107,7 +107,7 @@ class llama2_base:
             A list of strings of suggested actions by the LLM. Also sets self.suggestions
 
         """
-        observation = obs_to_string(observation, False, False)
+        observation = obs_to_string(observation, show_pos, False)
 
         observation = f"My goal is: {self.goal}. {observation}"
 
@@ -288,9 +288,48 @@ class llama2_policy(llama2_base):
             goal, cos_sim_threshold, similarity_modifier, temperature, top_p, rl_temp, cache_file
         )
 
+        self.dialog = [
+            {
+                "role": "system",
+                "content": f"You are a helpful assistant giving advice to someone playing a videogame. You will recieve the current goal of the game and a list of observations about the environment, and you should give a list of suggested actions that the player should take to reach their goal. The suggested actions should involve the objects mentioned. The player can pick up items, drop items and use items. You should not make assumptions about the environment, only use the information given to you. If none of the observations are relevant to solving the task, respond that the player should explore more. Separate each suggestion with a new line. Be concise.",
+            },
+            {
+                "role": "user",
+                "content": "My goal is: pick up the purple box. I see a red key 2 squares LEFT and 3 squares FORWARD and a locked red door 2 squares RIGHT and 3 squares FORWARD. What actions do you suggest?",
+            },
+            {
+                "role": "assistant",
+                "content": "Go FORWARD 3 squares then LEFT 2 squares. \nPick up the key.",
+            },
+            {
+                "role": "user",
+                "content": "My goal is: open the purple door. I see a red key 2 squares LEFT and 3 squares FORWARD, an open red door 2 squares RIGHT and 3 squares FORWARD, a purple key 3 squares LEFT, a locked purple door 2 squares RIGHT, and a green key 1 square RIGHT and 1 square FORWARD. What actions do you suggest?",
+            },
+            {
+                "role": "assistant",
+                "content": "Move LEFT 3 squares. \nPick up the key.",
+            },
+            {
+                "role": "user",
+                "content": "My goal is: open the purple door. I see a green key 1 square and a locked green door. What actions do you suggest?",
+            },
+            {
+                "role": "assistant",
+                "content": "You should turn around to see more.",
+            },
+            {
+                "role": "user",
[…]

### 2023-10-26 src/ActorCritic.py
debug llama policy in actorcritic

@@ -96,14 +96,14 @@ class Agent(nn.Module):
                 advisor_values_list = list()
                 for single_obs in observation:
                     # its flattened so we need to make it normal again
-                    unflat_obs = np.array(single_obs.reshape((7, 7, 3)).to(torch.int64))
+                    unflat_obs = torch.tensor(single_obs.reshape((7, 7, 3)).to(torch.int64))
                     # this stores suggested actions
                     self.consigliere.suggest(unflat_obs)
                     # compares all possible actions to suggestions
                     advisor_values_list.append(self.consigliere.give_values(unflat_obs))
                 advisor_values = torch.stack(advisor_values_list)
             else:
-                unflat_obs = np.array(observation.reshape((7, 7, 3)).to(torch.int64))
+                unflat_obs = torch.tensor(observation.reshape((7, 7, 3)).to(torch.int64))
                 self.consigliere.suggest(unflat_obs)
                 advisor_values = self.consigliere.give_values(unflat_obs)
 

### 2023-10-26 src/utils.py
debug llama policy

@@ -340,7 +340,7 @@ def obs_to_string(obs_matrix, positions=True, you=True):
     for i in range(len(obs_matrix)):
         for j in range(len(obs_matrix[i])):
             cell = obs_matrix[i][j]
-
+            print(cell)
             if cell[0] < 4:
                 continue
 

### 2023-10-26 src/ppo.py
fix llama policy integration in ppo

@@ -14,6 +14,7 @@ import matplotlib.pyplot as plt
 
 base_path = os.path.dirname(__file__)
 print(base_path)
+#torch.autograd.set_detect_anomaly(True)
 
 class PPO:
     def __init__(self, param_file, llama_policy=False, llama_reward=False, results_file=None):

### 2023-10-26 doc/paper.tex
Smol bean update
Signed-off-by: Jonatan Hanssen <jonatan.hanssen@outlook.com>

@@
 
+\subsection{Minigrid Environments}
+
+Minigrid is a library containing simple grid world environments.
+An important characteristic of all these environments is that they only give reward on the completion of the goal of the environment.
+In other words, the environment only ever gives out a single reward per episode, at the end of the episode.
+Goals in minigrid are often simple tasks, such as "unlock the door", "go to the goal", "pick up the red chest".
+Completing these goals often requires the completion of subtasks; going to the goal might require opening a door, and opening the door might require picking up a key.
+These environments are partially observable, which means that the agent only receives a seven by seven observation grid, not the entire grid.
+Figure \ref{doorkeyenv}, shows a typical minigrid environment.
+In this case, the agent (the red arrow) only sees the left room, and does not know where the goal is.
+
+\begin{figure}[h]
+\centerline{\includegraphics[width=3.25in]{figure/doorkeyenv.png}}
+\caption{The Minigrid DoorKey environment}
+\label{doorkeyenv}
+\end{figure}
+
 \section{Results}

### 2023-10-26 src/plot_rewards.py

@@ -1,29 +1,53 @@
 import numpy as np
 import matplotlib.pyplot as plt
-import os
-import argparse
+import os, argparse, glob
 
-
+plt.rcParams.update({"font.size": 22})
 parser = argparse.ArgumentParser()
 
-parser.add_argument("llm_file")
-parser.add_argument("ppo_file")
+parser.add_argument("environment")
 args = parser.parse_args()
 
-abs_path = os.path.dirname(__file__)
 
-for arg in vars(args):
-    file_name = os.path.join(abs_path, getattr(args, arg))
+data_path = os.path.join(os.path.dirname(__file__), "data/")
 
-    data = np.load(file_name)
+llm_data = list()
+non_llm_data = list()
 
-    means = list()
-    for window in np.array_split(data, len(data) // 10):
-        means.append(np.mean(window))
+for file in os.listdir(data_path):
+    if not file.endswith(".npy"):
+        continue
+    if not args.environment in file:
+        continue
 
-    plt.plot(
-        range(len(means)), means, label=f"{getattr(args, arg)}".replace(".npy", "")
-    )
+    data = np.load(os.path.join(data_path, file))
 
+    if "llm" in file:
+        llm_data.append(data)
+
+    else:
+        non_llm_data.append(data)
+
+llm_data = np.mean(np.vstack(llm_data), axis=0)
+non_llm_data = np.mean(np.vstack(non_llm_data), axis=0)
+
+window_size = 15
+
+
+means = list()
+for window in np.array_split(llm_data, len(llm_data) // window_size):
+    means.append(np.mean(window))
+
+plt.plot(np.arange(len(means)) * window_size, means, label="LLM reward shaping")
+
+means = list()
+for window in np.array_split(non_llm_data, len(non_llm_data) // window_size):
+    means.append(np.mean(window))
+
+plt.plot(np.arange(len(means)) * window_size, means, label="No LLM reward shaping")
+
+plt.title("Average reward per episode on DoorKey environment")
+plt.xlabel("Episodes")
+plt.ylabel("Reward")
 plt.legend()
 plt.show()

### 2023-10-27 doc/paper.tex
Write more on paper
Signed-off-by: Jonatan Hanssen <jonatan.hanssen@outlook.com>

@@
 \title{Large Language Models in Reinforcement Learning - A Comparison of Methods}
+\thispagestyle{plain}
+\pagestyle{plain}
 
@@
 
+\subsection{Large Language Model used}
+
+The LLM that is used is Meta's Llama2 \cite{llama}, which is an open source language model. We use the smallest model size, which has 7 billion parameters.
+
 \section{Experiments}
+
+We perform experiments to compare different methods of integrating LLMs with the PPO algorithm, specifically comparing their ability to guide the early exploration of an RL model in a reward-sparse environment.
 
@@
 
-Minigrid is a library containing simple grid world environments.
+For our experiments, we choose the Minigrid library.
+The Minigrid library contains simple grid world environments.
 An important characteristic of all these environments is that they only give reward on the completion of the goal of the environment.
@@
 
+A more complicated environment can be seen in figure \ref{blockedunlockpickupenv}.
+Completing the goal requires picking up the ball and dropping it somewhere else, picking up the key, using it to unlock the door, dropping the key (the agent can only carry one item at once) and finally picking up the box.
+With no intermediate rewards, this is a very difficult task for traditional RL methods.
+The exact configuration of Minigrid environments is random; the position of the agent, the placement of the walls and items within the grid all vary from episode to episode.
+
+\begin{figure}[h]
+\centerline{\includegraphics[width=3.25in]{figure/blockedunlockpickupenv.png}}
+\caption{The Minigrid BlockedUnlockPickup environment}
+\label{blockedunlockpickupenv}
+\end{figure}
+
+Some environments can only be effectively solved by relying on language (for example, environments where the goal text describes a specific object to be picked up, which changes every time), while others can be solved without (for example, if the goal always is to pick up a key).
+We only use environments that can be solved without language, so that the baseline PPO algorithm is not at an unfair disadvantage.
+
+\subsection{Our chosen environments}
+
+We select NUMBER different Minigrid environments; DoorKey, Unlock, CRAZYENV.
+Example instances of these environments can be found in the appendix.
+For each environment we test our three different models; PPO with no LLM integration, PPO with LLM reward shaping and PPO with LLM policy integration.
+All methods use the same hyperparameters for the PPO, which can be found the appendix.
+For each of these methods, we run PPO for 2000 episodes of 1024 steps, and track the reward\footnote{The true environment reward, not including pseudo-rewards given by the LLM} received for each one.
+We do this eight times per method, and compare the average results against eachother.
+
 \section{Results}
+
+\subsection{DoorKey environment}
+
+As we can see from figure \ref{doorkeyresults}, LLM integration gives a noticable improvement to the performance of the model, not just in the early stages of training, but throughout the entire training process.
+Averaging over all episodes of all eight runs per model, we find that LLM reward shaping receives $24\%$ more reward per episode than PPO trained without LLM rewards.
+Furthermore, this is an improvement that comes at very little computational cost, because LLM responses can be cached between runs.
+Of the eight runs on this environment using LLM reward shaping, the last three runs never queried the LLM at all, only relying on the cache of previous answers.
+In all methods, we see that the variance is very large ($0.28$ and $0.31$ for LLM reward shaping and baseline, respectively).
+Given that there is no reward given for a partially completed task, this is not unexpected.
+
+\begin{figure}[h]
+\centerline{\includegraphics[width=3.25in]{figure/doorkeyresults.png}}
+\caption{Reward per episode for the three (two) different methods in the DoorKey environment}
+\label{doorkeyresults}
+\end{figure}
 

### 2023-10-27 doc/paper.tex
Ad dmore to paper
Signed-off-by: Jonatan Hanssen <jonatan.hanssen@outlook.com>

@@
 The apparent simplicity of such problems is often due to human intuition, which allow us to quickly see possible solutions to a diverse set of problems.
-Large Language Models (LLMs) are trained on large corpora of human written text, and have been shown to capture parts of this intuition in many tasks \cite{sparks}.
-In recent years, LLMs have been succesfully used to aid RL agents in more efficient exploration, and have allowed them to solve problems which previous methods have been unable to \cite{omni} \cite{ellm}.
+Large Language Models (LLMs) are trained on large corpora of human written text, and have been shown to capture parts of this intuition in many tasks.
+In recent years, LLMs have been succesfully used to aid RL agents in more efficient exploration, and have allowed them to solve problems which previous methods have been unable to.
 We compare different methods for integrating an LLM into the Deep Learning RL algorithm Proximal Policy Optimization (PPO), and compare their efficiency against each other.
@@
 Thus, a new area of research has emerged, which attemps to use the vast amount of human knowledge encoded in these models to improve the performance of reinforcement learning algorithms \cite{survey}.
-These LLM assisted RL agents have been able to outperform state-of-the-art RL methods in many problems \cite{omni} \cite{ellm} \cite{idm}.
+These LLM assisted RL agents have been able to outperform state-of-the-art RL methods in many problems \cite{omni, ellm, idm}.
 
@@
 RL informed by natural language is a relatively new field, where research before 2018 have been limited to relatively small corpora or synthetic language \cite[1]{survey}.
-With the introduction of Large Language Models, many new possibilities have opened up, and in recent years there have been many successful integrations of these models into traditional RL \cite{omni} \cite{ellm} \cite{idm}.
+With the introduction of Large Language Models, many new possibilities have opened up, and in recent years there have been many successful integrations of these models into traditional RL \cite{omni, ellm, idm}.
 
@@
 However, rather than doing so through directly influencing action probabilities in the policy, the paper suggests rewarding the agent if it takes actions that are semantically similar to actions suggested by an LLM.
+For example, the LLM might recommend the agent to "cut down a tree".
+If the agent performs the "chop" action at a tree, this action can be described as "chop tree".
+Using a language model, we can create encodings of these sentences and compare their similarity mathematically.
 At certain timesteps, the LLM is prompted with a state observation from the agent, and is asked to suggest actions for the agent to take.
-If the agent then takes actions which are semantically similar to the suggested actions, it recieves a reward.
+If the agent then takes actions which are semantically similar to the suggested actions, it receives a reward.
 Thus, the proposed algorithm makes use of LLMs in two ways: first to generate what the LLM perceives as desirable actions, then later to assess similarity between a natural-language description of the agents action and the previously suggested actions.
@@
 
-We select NUMBER different Minigrid environments; DoorKey, Unlock, CRAZYENV.
+We select NUMBER different Minigrid environments; DoorKey, OTHERENV, OTHERENV.
 Example instances of these environments can be found in the appendix.
@@
 
-In accordance with the NeurIPS Code of Ethics ***ADD REFERENCE HERE***, we outline ethical considerations related to this project, covering potential consequences from the research, and how potential risks may be mitigated.
+In accordance with the NeurIPS Code of Ethics \cite{ethics}, we outline ethical considerations related to this project, covering potential consequences from the research, and how potential risks may be mitigated.
 Considering that our research does not involve participants nor sensitive data, concerns related to how such factors are treated will not be discussed.
@@
 
+% \begin{table*}[h]
+% \caption{PPO hyperparameters}
+% \begin{center}
+% \label{hyperparams}
+% \begin{tabular}{c | l l l l l l l l} 
+% % & & \\ % put some space after the caption
+% % \hline
+% Parameter & Gamma & Steps/episode & Epochs/episode & Batches/epoch & Clip & Value loss coeff. & Entropy coeff. & Learning rate \\
+% \hline
+% Value & 0.99 & 1024 & 10 & 8 & 0.2 & 0.3 & 0 & 3.0e-4 \\
+% \hline
+% \end{tabular}
+% \end{center}
+% \end{table*}
+\begin{table*}[h]
+\caption{PPO hyperparameters}
+\begin{center}
+\label{hyperparams_transposed}
+\begin{tabular}{c | c}
+Parameter & Value \\
+\hline
+Gamma & 0.99 \\
+Steps/episode & 1024 \\
+Epochs/episode & 10 \\
+Batches/epoch & 8 \\
+Clip & 0.2 \\
+Value loss coeff. & 0.3 \\
+Entropy coeff. & 0 \\
+Learning rate & 3.0e-4 \\
+\end{tabular}
+\end{center}
+\end{table*}
+
 \end{document}

### 2023-10-27 src/ppo.py
add cache miss counter to llama2_base and count per rollout when training ppo

@@ -71,6 +71,9 @@ class PPO:
             #     continue
             print(f"Env reward: {env_reward}")
             print(f"LLM reward: {advisor_reward_cum}")
+            if self.reward_shaper:
+                print(f"Cache misses={self.reward_shaper.cache_misses}")
+                self.reward_shaper.cache_misses = 0
             self.PPO_update()  # Use values stored to backpropagate using PPO
             self.reset_memory()  # Restore
 

### 2023-10-27 src/utils.py

@@ -46,6 +46,7 @@ class llama2_base:
         self.goal = goal
 
         self.cache_file = cache_file
+        self.cache_misses = 0
 
         if self.cache_file:
             self.cache_file = os.path.join(base_path, cache_file)
@@ -119,6 +120,7 @@ class llama2_base:
 
         else:
             # print("Cache miss")
+            self.cache_misses += 1
             result = self.generator.chat_completion(
                 [self.dialog],  # type: ignore
                 max_gen_len=100,

### 2023-10-27 src/ActorCritic.py
put llama_reward and policy into hyperparams. fix caching for policy

@@ -92,26 +92,27 @@ class Agent(nn.Module):
         if self.consigliere:
             #print(observation.shape)
             #logits = torch.nn.functional.softmax(logits, dim=-1)
+            #logits /= torch.norm(logits)
             if len(observation.shape) == 2:
                 advisor_values_list = list()
                 for single_obs in observation:
                     # its flattened so we need to make it normal again
-                    unflat_obs = torch.tensor(single_obs.reshape((7, 7, 3)).to(torch.int64))
+                    unflat_obs = single_obs.reshape((7, 7, 3)).to(torch.int64)
                     # this stores suggested actions
                     # self.consigliere.suggest(unflat_obs)
                     # compares all possible actions to suggestions
                     advisor_values_list.append(self.consigliere.give_values(np.array(unflat_obs.cpu())))
                 advisor_values = torch.stack(advisor_values_list)
             else:
-                unflat_obs = torch.tensor(observation.reshape((7, 7, 3)).to(torch.int64))
+                unflat_obs = observation.reshape((7, 7, 3)).to(torch.int64)
                 #self.consigliere.suggest(unflat_obs)
                 advisor_values = self.consigliere.give_values(np.array(unflat_obs.cpu()))
 
 
             # anti adrian propaganda
-            # logits += advisor_values
+            logits += advisor_values
             # adrian good vote adrian
-            logits *= advisor_values
+            # logits *= advisor_values
 
         probs = Categorical(logits=logits)
 

### 2023-10-27 src/ppo.py

@@ -17,7 +17,7 @@ print(base_path)
 #torch.autograd.set_detect_anomaly(True)
 
 class PPO:
-    def __init__(self, param_file, llama_policy=False, llama_reward=False, results_file=None):
+    def __init__(self, param_file, results_file=None):
         self.args = read_params(param_file)
         seeding(self.args["seed"])
 
@@ -29,7 +29,7 @@ class PPO:
 
         self.device = torch.device("cuda" if torch.cuda.is_available() else "cpu")
         # self.device = torch.device("cpu")
-        self.agent = Agent(self.env, llama_policy).to(self.device)
+        self.agent = Agent(self.env, self.args["llama_policy"]).to(self.device)
         self.optimizer = optim.Adam(
             self.agent.parameters(), lr=self.args["lr"], eps=1e-8
         )
@@ -37,7 +37,7 @@ class PPO:
 
         self.reward_shaper = (
             llama2_reward_shaper(self.env.reset()[0]["mission"], similarity_modifier=0.005, cos_sim_threshold=0.84, cache_file="llm_cache.json")
-            if llama_reward
+            if self.args["llama_reward"]
             else None
         )
 
@@ -72,8 +72,11 @@ class PPO:
             print(f"Env reward: {env_reward}")
             print(f"LLM reward: {advisor_reward_cum}")
             if self.reward_shaper:
-                print(f"Cache misses={self.reward_shaper.cache_misses}")
+                print(f"Cache misses: {self.reward_shaper.cache_misses}")
                 self.reward_shaper.cache_misses = 0
+            if self.agent.consigliere:
+                print(f"Cache misses: {self.agent.consigliere.cache_misses}")
+                self.agent.consigliere.cache_misses = 0
             self.PPO_update()  # Use values stored to backpropagate using PPO
             self.reset_memory()  # Restore
 
@@ -141,6 +144,7 @@ class PPO:
             pass
             self.reward_shaper.caption_set = set()
             # self.reward_shaper.reset_cache()
+
 
         # Now that the agent has played out an episode, it's time
         # to backtrack all steps, and compute the discounted rewards
@@ -268,6 +272,6 @@ class PPO:
 
 
 if __name__ == "__main__":
-    ppo = PPO("hyperparams.json", llama_reward=False, llama_policy=True, results_file="doorkey_big_llm")
+    ppo = PPO("hyperparams.json", results_file="doorkey_big_llm")
     # save_params(self.self.args)
     ppo.train()

### 2023-10-27 src/utils.py

@@ -54,9 +54,10 @@ class llama2_base:
             if os.path.exists(self.cache_file):
                 with open(self.cache_file) as file:
                     self.cache = json.load(file)
+            else:
+                self.cache = dict()
         else:
             self.cache = dict()
-
         self.caption_set = set()
 
         self.dialog = [

### 2023-10-30 src/plot_rewards.py
Add more paper
Signed-off-by: Jonatan Hanssen <jonatan.hanssen@outlook.com>

@@ -2,7 +2,8 @@ import numpy as np
 import matplotlib.pyplot as plt
 import os, argparse, glob
 
-plt.rcParams.update({"font.size": 22})
+plt.rcParams.update({"font.size": 26})
+# plt.rcParams["font.family"] = "Times New Roman"
 parser = argparse.ArgumentParser()
 
 parser.add_argument("environment")
@@ -28,26 +29,59 @@ for file in os.listdir(data_path):
     else:
         non_llm_data.append(data)
 
-llm_data = np.mean(np.vstack(llm_data), axis=0)
-non_llm_data = np.mean(np.vstack(non_llm_data), axis=0)
+llm_data = np.vstack(llm_data)
+non_llm_data = np.vstack(non_llm_data)
+
+mean_llm_data = np.mean(llm_data, axis=0)
+mean_non_llm_data = np.mean(non_llm_data, axis=0)
+
+std_llm_data = np.std(llm_data, axis=0)
+std_non_llm_data = np.std(non_llm_data, axis=0)
+
+print(np.mean(mean_llm_data) / np.mean(non_llm_data))
+print(f"{np.mean(std_llm_data)=}")
+print(f"{np.mean(std_non_llm_data)=}")
+
 
 window_size = 15
 
-
 means = list()
-for window in np.array_split(llm_data, len(llm_data) // window_size):
+stds = list()
+for window in np.array_split(mean_llm_data, len(mean_llm_data) // window_size):
     means.append(np.mean(window))
 
-plt.plot(np.arange(len(means)) * window_size, means, label="LLM reward shaping")
+for window in np.array_split(std_llm_data, len(std_llm_data) // window_size):
+    stds.append(np.mean(window))
+
+x = np.arange(len(means)) * window_size
+means = np.array(means)
+stds = np.array(stds)
+print(np.mean(stds))
+
+line_llm = plt.plot(x, means, label="LLM reward shaping", color="b", linewidth=3)
+fill_llm = plt.fill_between(
+    x, means - stds, means + stds, alpha=0.1, color="b", linewidth=4
+)
 
 means = list()
-for window in np.array_split(non_llm_data, len(non_llm_data) // window_size):
+stds = list()
+for window in np.array_split(mean_non_llm_data, len(mean_non_llm_data) // window_size):
     means.append(np.mean(window))
 
-plt.plot(np.arange(len(means)) * window_size, means, label="No LLM reward shaping")
[…]

### 2023-10-30 src/test.py

@@ -7,13 +7,14 @@ import torch
 from utils import caption_action
 
 
-env = gym.make("MiniGrid-UnlockPickup-v0", render_mode="human")
+env = gym.make("MiniGrid-KeyCorridorS3R2-v0", render_mode="human")
 
 
 obs, dog = env.reset()
 print(obs)
 obs = obs["image"]
 observation = env.step(1)[0]["image"]
+time.sleep(100)
 stringy = [[f"{cell[0]}{cell[1]}{cell[2]}" for cell in row] for row in observation]
 for row in stringy:
     print(row)
@@ -29,4 +30,3 @@ for row in stringy:
     print(row)
 
 print(caption_action(1, observation))
-env.close()

### 2023-10-30 src/utils.py
prompt engineering for llama policy

@@ -210,22 +210,22 @@ def caption_action(action, obs_matrix):
 
     ###  Captioning action based on cell and inventory ###
 
-    caption = "do nothing"
+    caption = "do nothing."
 
     if action == "done":
-        caption = "mark the game as completed"
+        caption = "mark the game as completed."
 
     if action == "forward":
         if front_item == "empty":
-            caption = "move forward"
+            caption = "move forward."
         elif front_item == "door" and front_state == " open":
-            caption = "move forward"
+            caption = "move forward."
 
     if action == "right":
-        caption = "turn right"
+        caption = "turn right."
 
     if action == "left":
-        caption = "turn left"
+        caption = "turn left."
 
     if action == "pick":
         if front_cell[0] < 4:  # nothing to pick up
@@ -235,7 +235,7 @@ def caption_action(action, obs_matrix):
             pass
 
         else:
-            caption = f"pick up{front_state} {front_color} {front_item}"
+            caption = f"pick up{front_state} {front_color} {front_item}."
 
     elif action == "drop":
         if inventory[0] < 4:  # nothing to drop
@@ -245,17 +245,19 @@ def caption_action(action, obs_matrix):
             pass
 
         else:
-            caption = f"drop{inventory_state} {inventory_color} {inventory_item}"
+            caption = f"drop{inventory_state} {inventory_color} {inventory_item}."
 
     elif action == "toggle":
         if front_item == "box":
-            caption = f"destroy {front_color} box"
+            caption = f"destroy {front_color} box."
         else:
             if front_cell[0] >= 4:
                 caption = f"use{front_state} {front_color} {front_item}"
 
             if inventory[0] > 3:
                 caption += f" with{inventory_state} {inventory_color} {inventory_item}"
+            
+            caption += "."
 
     return caption
 
@@ -302,7 +304,7 @@ class llama2_policy(llama2_base):
             },
             {
                 "role": "assistant",
-                "content": "Go FORWARD 3 squares then LEFT 2 squares. \nPick up the key.",
+                "content": "move forward. \nturn left.",
             },
             {
                 "role": "user",
@@ -310,7 +312,7 @@ class llama2_policy(llama2_base):
             },
             {
                 "role": "assistant",
-                "content": "Move LEFT 3 squares. \nPick up the key.",
+                "content": "turn left.",
             },
             {
                 "role": "user",
@@ -318,16 +320,33 @@ class llama2_policy(llama2_base):
             },
             {
                 "role": "assistant",
[…]

### 2023-10-30 src/ActorCritic.py
added similarity cache for caption-suggestion pairs. bug fixes

@@ -70,7 +70,7 @@ class Agent(nn.Module):
         )
 
         goal = env.reset()[0]["mission"]
-        self.consigliere = llama2_policy(goal) if llama else None
+        self.consigliere = llama2_policy(goal, cos_sim_threshold=0, similarity_modifier=0.1, cache_file="policy_cache.json", sim_cache_file="policy_sim_cache.json") if llama else None
 
         # Yet to be integrated -> shall serve as the second actor
         # self.LM_actor = llama2_7b_policy()
@@ -108,8 +108,9 @@ class Agent(nn.Module):
                 #self.consigliere.suggest(unflat_obs)
                 advisor_values = self.consigliere.give_values(np.array(unflat_obs.cpu()))
 
-
-            # anti adrian propaganda
+            # print(f"{logits=}")
+            # print(f"{advisor_values=}")
+            # # anti adrian propaganda
             logits += advisor_values
             # adrian good vote adrian
             # logits *= advisor_values

### 2023-10-30 src/ppo.py

@@ -89,6 +89,8 @@ class PPO:
                     np.save(file_name, np.array(rewards))
                 if self.reward_shaper:
                     self.reward_shaper.save_cache()
+                if self.agent.consigliere:
+                    self.agent.consigliere.save_cache()
 
 
             i += 1
@@ -98,6 +100,7 @@ class PPO:
         perform PPO update
         """
 
+        print("Entered next_episode()")
         observation_dict, _ = self.env.reset()
         observation = torch.Tensor(observation_dict["image"].flatten()).to(self.device)
         env_reward = 0
@@ -166,6 +169,8 @@ class PPO:
 
 
     def PPO_update(self):
+
+        print("Entered PPO_update()")
         for epoch in range(self.args["epochs"]):
             for values in self.minibatch_generator():
                 (

### 2023-10-30 src/utils.py

@@ -160,8 +160,8 @@ class llama2_base:
         """
         caption = caption_action(action, obs_matrix)
 
-        if not caption or caption in self.caption_set:
-            return 0
+        # if not caption or caption in self.caption_set:
+        #     return 0
 
         self.caption_set.add(caption)
 
@@ -288,15 +288,30 @@ class llama2_policy(llama2_base):
         top_p=0.9,
         rl_temp=0,
         cache_file=None,
+        sim_cache_file=None
     ):
         super().__init__(
             goal, cos_sim_threshold, similarity_modifier, temperature, top_p, rl_temp, cache_file
         )
 
+        self.sim_cache_file = sim_cache_file
+        self.cache_misses = 0
+
+        if self.sim_cache_file:
+            self.sim_cache_file = os.path.join(base_path, sim_cache_file)
+
+            if os.path.exists(self.sim_cache_file):
+                with open(self.sim_cache_file) as file:
+                    self.sim_cache = json.load(file)
+            else:
+                self.sim_cache = dict()
+        else:
+            self.sim_cache = dict()
+
         self.dialog = [
             {
                 "role": "system",
-                "content": f"You are a helpful assistant giving advice to someone playing a videogame. You will recieve the current goal of the game and a list of observations about the environment, and you should give a list of suggested actions that the player should take to reach their goal. The suggested actions should involve the objects mentioned. The player can pick up items, drop items and use items. You should not make assumptions about the environment, only use the information given to you. If none of the observations are relevant to solving the task, respond that the player should explore more. Separate each suggestion with a new line. Be concise.",
[…]

### 2023-10-30 src/ppo.py
new result for ppo policy

@@ -100,7 +100,7 @@ class PPO:
         perform PPO update
         """
 
-        print("Entered next_episode()")
+        # print("Entered next_episode()")
         observation_dict, _ = self.env.reset()
         observation = torch.Tensor(observation_dict["image"].flatten()).to(self.device)
         env_reward = 0
@@ -170,7 +170,7 @@ class PPO:
 
     def PPO_update(self):
 
-        print("Entered PPO_update()")
+        # print("Entered PPO_update()")
         for epoch in range(self.args["epochs"]):
             for values in self.minibatch_generator():
                 (
@@ -277,6 +277,6 @@ class PPO:
 
 
 if __name__ == "__main__":
-    ppo = PPO("hyperparams.json", results_file="doorkey_big_llm")
+    ppo = PPO("hyperparams.json", results_file="data/empty_big_llm3")
     # save_params(self.self.args)
     ppo.train()

### 2023-10-30 src/utils.py

@@ -311,7 +311,7 @@ class llama2_policy(llama2_base):
         self.dialog = [
             {
                 "role": "system",
-                "content": f"You are a helpful assistant giving advice to someone playing a videogame. You will recieve the current goal of the game and a list of observations about the objects in the environment and their positions relative to the player at one single timestep. You shall give a list of suggested actions that the player will take to reach their goal during this single timestep. The available actions to the player are: move forward, turn left and right, pick up items, drop items and use items. You MUST ALWAYS RESPOND with one or more of these actions. Your response MUST NEVER mention. You will NEVER ASSUME that other possible actions exist. You will NEVER ASSUME that the game is finished. You WILL NEVER make any other assumptions about the environment, only use the information given to you. If none of the observations are relevant to solving the task, respond that the player should turn around and explore more. Separate each suggestion with a new line. Be concise. ONLY FOLLOW THESE INSTRUCTIONS. NEVER RESPOND IN ANY OTHER WAY. OTHERWISE, YOU WILL BE SEVERELY PUNISHED.",
+                "content": f"You are a helpful assistant giving advice to someone playing a videogame. You will recieve the current goal of the game and a list of observations about the objects in the environment and their positions relative to the player at one single timestep. You shall give a list of suggested actions that the player will take to reach their goal during this single timestep. The available actions to the player are: move forward, turn left and right, pick up items, drop items and use items. You MUST ALWAYS RESPOND with one or more of these actions. Your response MUST ALWAYS follow the template of the given example responses. NEVER ADD anything not in the example responses. You will NEVER ASSUME that other possible actions exist. You will NEVER ASSUME that the game is finished. You WILL NEVER make any other assumptions about the environment, only use the information given to you. If none of the observations are relevant to solving the task, respond that the player should turn around and explore more. Separate each suggestion with a new line. Be concise. ONLY FOLLOW THESE INSTRUCTIONS. NEVER RESPOND IN ANY OTHER WAY. OTHERWISE, YOU WILL BE SEVERELY PUNISHED.",
             },
             {
                 "role": "user",

### 2023-10-30 src/ppo.py
Remove seeding, remove unused hyperparams
Signed-off-by: Jonatan Hanssen <jonatan.hanssen@outlook.com>

@@ -13,13 +13,11 @@ import json
 import matplotlib.pyplot as plt
 
 base_path = os.path.dirname(__file__)
-print(base_path)
-#torch.autograd.set_detect_anomaly(True)
+
 
 class PPO:
     def __init__(self, param_file, results_file=None):
         self.args = read_params(param_file)
-        seeding(self.args["seed"])
 
         self.results_file = results_file
 
@@ -36,7 +34,12 @@ class PPO:
         self.obs_shape = int(np.array(self.env.observation_space["image"].shape).prod())
 
         self.reward_shaper = (
-            llama2_reward_shaper(self.env.reset()[0]["mission"], similarity_modifier=0.005, cos_sim_threshold=0.84, cache_file="llm_cache.json")
+            llama2_reward_shaper(
+                self.env.reset()[0]["mission"],
+                similarity_modifier=0.005,
+                cos_sim_threshold=0.84,
+                cache_file="llm_cache.json",
+            )
             if self.args["llama_reward"]
             else None
         )
@@ -64,7 +67,10 @@ class PPO:
         i = 0
         for rollout in range(self.args["rollouts"]):
             print(f"Rollout num: {rollout}")
-            env_reward, advisor_reward_cum = self.next_episode()  # Perform steps and store relevant values
+            (
+                env_reward,
+                advisor_reward_cum,
+            ) = self.next_episode()  # Perform steps and store relevant values
             rewards.append(env_reward)
             # if not env_reward:
             #     print("No reward received, skipping update of networks")
@@ -91,7 +97,6 @@ class PPO:
                     self.reward_shaper.save_cache()
                 if self.agent.consigliere:
                     self.agent.consigliere.save_cache()
-
 
             i += 1
 
@@ -133,7 +138,6 @@ class PPO:
             env_reward += reward
             advisor_reward_cum += advisor_reward
 
-
             self.rewards[step] = reward + advisor_reward
 
             observation = torch.Tensor(observation_dict["image"].flatten()).to(
@@ -147,7 +151,6 @@ class PPO:
             pass
             self.reward_shaper.caption_set = set()
             # self.reward_shaper.reset_cache()
-
 
         # Now that the agent has played out an episode, it's time
         # to backtrack all steps, and compute the discounted rewards
@@ -167,9 +170,7 @@ class PPO:
             # print(f"{self.advantages=}")
         return env_reward, advisor_reward_cum
 
-
     def PPO_update(self):
-
[…]

### 2023-10-30 src/ppo.py
just some more changes

@@ -277,6 +277,6 @@ class PPO:
 
 
 if __name__ == "__main__":
-    ppo = PPO("hyperparams.json", results_file="data/empty_big_llm3")
+    ppo = PPO("hyperparams.json", results_file="data/doorkey_big_policy1")
     # save_params(self.self.args)
     ppo.train()

### 2023-10-31 doc/paper.tex
Add more paper
Signed-off-by: Jonatan Hanssen <jonatan.hanssen@outlook.com>
Filling in Approach part
Signed-off-by: Gregz9 <gregorkajda@gmail.com>

@@
 
-Using this environment, we explore three different ways of training the agent. We evaluate these models on the same Minigrid environments and compare their performance against each other.\\
+To evaluate the effectivness of the integration of LLMs into reinfrocement learning algorithms, er use the \textbf{MiniGrid} and \textbf{BabyAI} libraries, we handpick specific environments with varying levels of complexity, to evaluate the effectivness of our proposed techniques against a standard proximal policy optimzation agent and each other.
+The environments are all based in a small sized grid-world, where the agent is surrounded by walls, and has to perform some form of task to complete the game.
+How difficult an environment is to beat, depends no factors such as shape of the walls, if the agent has to solve subtasks to finish the game, and the type of subtasks encountered.
+Moreover what makes these environments exceptionally well suited for testing our approach, is that they are sparse in that the agent will only receive a reward when it finishes the game.
+It is a well known fact that even the PPO algorithm struggles in very complex and sparse environments, hence it becomes the perfect arena to see if the high generalization capabilites of Large pre-trained Language Models will lead to better results than using only RL methods.
 
@@
 Of the eight runs on this environment using LLM reward shaping, the last three runs never queried the LLM at all, only relying on the cache of previous answers.
-In all methods, we see that the variance is very large ($0.28$ and $0.31$ for LLM reward shaping and baseline, respectively).
+In all methods, we see that the standard deviation is very large ($0.28$ and $0.31$ for LLM reward shaping and baseline, respectively).
 Given that there is no reward given for a partially completed task, this is not unexpected.

### 2023-10-31 doc/paper.tex
Picture this: a compact, grid-based world where an intelligent agent navigates through a series of rooms and corridors. The catch? Certain doors in this intricate maze are locked...
Signed-off-by: Gregz9 <gregorkajda@gmail.com>

--- 2023-10-27 doc/paper.tex
@@
 
-Using this environment, we explore three different ways of training the agent. We evaluate these models on the same Minigrid environments and compare their performance against each other.\\
+To evaluate the effectivness of the integration of LLMs into reinfrocement learning algorithms, we use the \textbf{MiniGrid} and \textbf{BabyAI} libraries.
+Here, we handpick specific environments with varying levels of complexity, to evaluate the effectivness of our proposed techniques against a standard Actor-Critic \ref{Proximal policy optimzation}, and against each other.
+The environments selected are all based in a small sized grid-world, where the agent is surrounded by walls, and has to perform at least one task to complete the game.
+How difficult an environment is to beat, depends no factors such as shape of the walls, if the agent has to solve subtasks to finish the game, and the type of subtasks encountered.
+Moreover what makes these environments exceptionally well suited for testing our approach, is that they are sparse, in the context of MiniGrid and BabyAI, this means that the agent will only receive a reward when it finishes the game.
+It is a well known fact that even the state-of-art RL algorithms struggle in very complex and sparse environments, hence it becomes the perfect arena to see if the high generalization capabilites of Large pre-trained Language Models will lead to better results than using only RL methods.
 
@@
 
-A standard PPO algorithm with no LLM integration, as a baseline. In this case, we use StableBaselines' implementation with the following hyperparameters: HYPERPARAMETERS HERE.
+Our baseline is the \textbf{Proximal Policy Optimization} (PPO), which over the last years has established itself as the state-of-art approach to solving reinforcement learning problems.
+The main idea behind this algorithms is that the new policy is clipped so that it does not move too quickly and too far from the old policy.
+I.e.
+the new policy cannot diverge too moch from the old policy, which leads to exceedingly large updates to policy gradients being avoided, thus also avoiding the problem of vanishing and exploding gradients.
+Suprisingly, this has elevated the ability of RL-techniques to tackle sparse environments, and outperform older approches such as DQN.
+For a more detailed discussion of the mathematics and implementation details, see \cite{OpenAIPPO}
+
+For the implementation details of our PPO and the hyperparameters used, take a look at Appendix X.X.
 
@@
 
-The LLM that is used is Meta's Llama2 \cite{llama}, which is an open source language model. We use the smallest model size, which has 7 billion parameters.
+For the purpose of our research, we have decided utilize a LLM from the open-source family of language models LLaMA developed by Meta AI, released starting Febraury, 2023.
+Specifically, we chose the newest of the models named Llama2 \cite{llama}, which has been trained using 40\% more data than its predecessor Llama1, and according to \cite{meta} has double the context length.
+In the experiments presented below, we use the smallest version of Llama2, with 7 billion parameters, which we download to run locally and reduce the bottleneck effect of having to call the API-version.
+The reason for not using any of the larger Llama2 models is simply the hardware limitation we face, with the 7B model requiring 19GB of VRAM to be run.
 
@@
 \section{Conclusion}
+
+It works man, dogski did it!
 

### 2023-10-31 doc/paper.tex
equiring the agent to find specific keys scattered throughout the environment. These keys serve as the ticket to unlock doors, enabling progress to new areas and ultimately leading to the completion of the task at hand...
Signed-off-by: Gregz9 <gregorkajda@gmail.com>

@@
 
+Many research papers nowadays propose fine-tuning of LLMs as means of leveraging the capabilites of the language models for decision-making purposes even further.
+Our plan of action, on the other hand, aims at evaluating the efficacy of the LLM without any previous pre-tuning.
+From the perspective of reinforcement learning, is is the appropriate course of action, as the LLM should not possess any previous knowledge about the task at hand, just like the agent should not.
+In addtion, the act of fine-tuning LLM is a rather time-comsuming process, requiring the generation and structuring of a new dataset, and also vast hardware resources.
+
 \section{Experiments}

### 2023-10-31 src/ppo.py
Add argument ability, cache file for both
Signed-off-by: Jonatan Hanssen <jonatan.hanssen@outlook.com>

@@ -11,15 +11,19 @@ from ActorCritic import *
 from utils import *
 import json
 import matplotlib.pyplot as plt
+import argparse
 
 base_path = os.path.dirname(__file__)
 
 
 class PPO:
-    def __init__(self, param_file, results_file=None):
+    def __init__(self, param_file, result_file=None, reward=None, policy=None):
         self.args = read_params(param_file)
 
-        self.results_file = results_file
+        # these arguments can overwrite the file if set
+        self.results_file = self.args["results_file"] if not result_file else result_file
+        self.llama_policy = self.args["llama_policy"] if not policy else policy
+        self.llama_reward = self.args["llama_reward"] if not reward else reward
 
         self.env = gym.make(
             self.args["env_name"], render_mode="rgb_array", max_steps=self.args["steps"]
@@ -27,7 +31,7 @@ class PPO:
 
         self.device = torch.device("cuda" if torch.cuda.is_available() else "cpu")
         # self.device = torch.device("cpu")
-        self.agent = Agent(self.env, self.args["llama_policy"]).to(self.device)
+        self.agent = Agent(self.env, self.llama_policy).to(self.device)
         self.optimizer = optim.Adam(
             self.agent.parameters(), lr=self.args["lr"], eps=1e-8
         )
@@ -38,9 +42,8 @@ class PPO:
                 self.env.reset()[0]["mission"],
                 similarity_modifier=0.005,
                 cos_sim_threshold=0.84,
-                cache_file="llm_cache.json",
             )
-            if self.args["llama_reward"]
+            if self.llama_reward
             else None
         )
 
@@ -148,7 +151,6 @@ class PPO:
                 break
 
         if self.reward_shaper:
-            pass
             self.reward_shaper.caption_set = set()
             # self.reward_shaper.reset_cache()
 
@@ -277,6 +279,12 @@ class PPO:
 
 
 if __name__ == "__main__":
-    ppo = PPO("hyperparams.json", results_file="data/doorkey_big_policy1")
-    # save_params(self.self.args)
+    parser = argparse.ArgumentParser()
+    parser.add_argument("-f", "--result_file", type=str, help="Output file for env rewards")
+    parser.add_argument("-r", "--reward", type=bool, help="Use LLM reward shaping")
+    parser.add_argument("-p", "--policy", type=bool, help="Use LLM policy")
+    args = parser.parse_args()
+    print(args)
+
+    ppo = PPO("hyperparams.json", result_file=args.result_file, reward=args.reward, policy=args.policy)
     ppo.train()

### 2023-10-31 src/ActorCritic.py
Add multippo
Signed-off-by: Jonatan Hanssen <jonatan.hanssen@outlook.com>

@@ -49,7 +49,7 @@ class Agent(nn.Module):
         the agent is in.
     """
 
-    def __init__(self, env, llama=False):
+    def __init__(self, env, llama=False, consigliere=None):
         super(Agent, self).__init__()
         self.critic = nn.Sequential(
             init_weightsNbias(nn.Linear(147 * 11, 64, dtype=torch.float64)),
@@ -70,7 +70,10 @@ class Agent(nn.Module):
         )
 
         goal = env.reset()[0]["mission"]
-        self.consigliere = llama2_policy(goal, cos_sim_threshold=0, similarity_modifier=0.1) if llama else None
+        print(goal)
+        if not consigliere:
+            self.consigliere = llama2_policy(goal, cos_sim_threshold=0, similarity_modifier=0.1) if llama else None
+        else self.consigliere = consigliere
 
 
     def get_value(self, observation):

### 2023-10-31 src/multi_ppo.py

@@ -1,0 +1,4 @@
+from ppo import PPO
+from utils import llama2_policy
+
+consigliere = llama2_policy(goal, cos_sim_threshold=0, similarity_modifier=0.1)

### 2023-10-31 src/ppo.py

@@ -17,11 +17,15 @@ base_path = os.path.dirname(__file__)
 
 
 class PPO:
-    def __init__(self, param_file, result_file=None, reward=None, policy=None):
+    def __init__(
+        self, param_file, result_file=None, reward=None, policy=None, consigliere=None
+    ):
         self.args = read_params(param_file)
 
         # these arguments can overwrite the file if set
-        self.results_file = self.args["results_file"] if not result_file else result_file
+        self.results_file = (
+            self.args["results_file"] if not result_file else result_file
+        )
         self.llama_policy = self.args["llama_policy"] if not policy else policy
         self.llama_reward = self.args["llama_reward"] if not reward else reward
 
@@ -31,7 +35,9 @@ class PPO:
 
         self.device = torch.device("cuda" if torch.cuda.is_available() else "cpu")
         # self.device = torch.device("cpu")
-        self.agent = Agent(self.env, self.llama_policy).to(self.device)
+        self.agent = Agent(self.env, self.llama_policy, consigliere=consigliere).to(
+            self.device
+        )
         self.optimizer = optim.Adam(
             self.agent.parameters(), lr=self.args["lr"], eps=1e-8
         )
@@ -280,11 +286,18 @@ class PPO:
 
 if __name__ == "__main__":
     parser = argparse.ArgumentParser()
-    parser.add_argument("-f", "--result_file", type=str, help="Output file for env rewards")
+    parser.add_argument(
+        "-f", "--result_file", type=str, help="Output file for env rewards"
+    )
     parser.add_argument("-r", "--reward", type=bool, help="Use LLM reward shaping")
     parser.add_argument("-p", "--policy", type=bool, help="Use LLM policy")
     args = parser.parse_args()
     print(args)
 
-    ppo = PPO("hyperparams.json", result_file=args.result_file, reward=args.reward, policy=args.policy)
+    ppo = PPO(
+        "hyperparams.json",
+        result_file=args.result_file,
+        reward=args.reward,
+        policy=args.policy,
+    )
     ppo.train()

### 2023-10-31 doc/paper.tex

@@
 \usepackage{xcolor}
+\usepackage{hyperref}
 \def\BibTeX{{\rm B\kern-.05em{\sc i\kern-.025em b}\kern-.08em
@@
 \section{Background and Related Work}
+\label{background}
 
@@
 
-To evaluate the effectivness of the integration of LLMs into reinfrocement learning algorithms, we use the \textbf{MiniGrid} and \textbf{BabyAI} libraries.
-Here, we handpick specific environments with varying levels of complexity, to evaluate the effectivness of our proposed techniques against a standard Actor-Critic \ref{Proximal policy optimzation}, and against each other.
+To evaluate the effectiveness of the integration of LLMs into Reinforcement Learning algorithms, we use the \textbf{MiniGrid} and \textbf{BabyAI} libraries.
+Here, we handpick specific environments with varying levels of complexity, to evaluate the effectiveness of our proposed techniques against a standard Actor-Critic \ref{Proximal policy optimzation}, and against each other.
 The environments selected are all based in a small sized grid-world, where the agent is surrounded by walls, and has to perform at least one task to complete the game.
-How difficult an environment is to beat, depends no factors such as shape of the walls, if the agent has to solve subtasks to finish the game, and the type of subtasks encountered.
-Moreover what makes these environments exceptionally well suited for testing our approach, is that they are sparse, in the context of MiniGrid and BabyAI, this means that the agent will only receive a reward when it finishes the game.
-It is a well known fact that even the state-of-art RL algorithms struggle in very complex and sparse environments, hence it becomes the perfect arena to see if the high generalization capabilites of Large pre-trained Language Models will lead to better results than using only RL methods.
+How difficult an environment is to beat depends on factors such as the shape of the walls, whether the agent has to solve subtasks to finish the game, and the types of subtasks encountered.
+Moreover, what makes these environments especially well suited for testing our approach is that they are sparse in the context of MiniGrid and BabyAI.
+This means that the agent will only receive a reward when it finishes the game.
+It is a well known fact that even state-of-art RL algorithms struggle in very complex and sparse environments.
+Therefore, it becomes the perfect arena to see if the high generalization capabilites of pre-trained Large Language Models will lead to better results than using only RL methods.
 
@@
 
-Our baseline is the \textbf{Proximal Policy Optimization} (PPO), which over the last years has established itself as the state-of-art approach to solving reinforcement learning problems.
-The main idea behind this algorithms is that the new policy is clipped so that it does not move too quickly and too far from the old policy.
-I.e.
-the new policy cannot diverge too moch from the old policy, which leads to exceedingly large updates to policy gradients being avoided, thus also avoiding the problem of vanishing and exploding gradients.
-Suprisingly, this has elevated the ability of RL-techniques to tackle sparse environments, and outperform older approches such as DQN.
+Our baseline is the \textbf{Proximal Policy Optimization} (PPO) algorithm, which has been established as a state-of-art approach to solving reinforcement learning problems over the last few years.
+The main idea behind this algorithm is that the new policy is clipped so that it does not move too quickly and too far from the old policy, i.e.
+the new policy cannot diverge too much from the old policy.
+This helps to avoid exceedingly large updates to policy gradients, thus also avoiding the problem of vanishing and exploding gradients.
+This has elevated the ability of RL-techniques to tackle sparse environments, and outperform older approches such as DQN.
 For a more detailed discussion of the mathematics and implementation details, see \cite{OpenAIPPO}
@@
 All possible actions are compared with the suggested action, and given a value which denotes their semantic similarity.
-These values are joined with the probability values that the normal PPO algorithm outputs and normalized, and a new probability distribution is created, where semantically similar actions are weighted higher than others.
+These values are joined with the logits of the actions, thus influencing the probability weights of each action being taken so that semantically similar actions are weighted higher than others.
 We call this the Actor-Advisor-Critic Model.
@@
 This may be done at every timestep, or after some interval of timesteps.
-In addition to the state observation, the LLM is also asked to respond with a list of recommendations.
-These recommendations can be proposed actions for the agent to perform, or goal states that the agent should reach.
-The agent then acts independently of the LLM for one or more timesteps, as dictated by the baseline RL algorithm it is trained with.
-As it does so, its actions are tracked and encoded to natural text.
-Alternatively, a new state observation is generated after the timesteps, and also encoded as natural text.
-Then, the LLM is used as a tool for semantic similarity comparison, comparing either its own formerly proposed actions to the actions performed by the agent, or comparing its proposed goal state to the new state reached by the agent.
+In addition to the state observation, the LLM is also asked to respond with a list of recommended actions for the agent to perform.
+The agent then acts independently of the LLM for that timestep, as dictated by the baseline RL algorithm it is trained with.
+As it does so, its chosen action is encoded to natural text.
+Then, the LLM is used as a tool for semantic similarity comparison, comparing its own formerly proposed action to the action performed by the agent.
 Finally, high semantic similarity gives an additional reward which is implemented in the reward function of the agent.
@@
 
-For the purpose of our research, we have decided utilize a LLM from the open-source family of language models LLaMA developed by Meta AI, released starting Febraury, 2023.
-Specifically, we chose the newest of the models named Llama2 \cite{llama}, which has been trained using 40\% more data than its predecessor Llama1, and according to \cite{meta} has double the context length.
+For the purpose of our research, we have decided utilize a LLM from the Llama open-source family of language models developed by Meta AI, first released in Febraury, 2023.
+Specifically, we chose the newest of the models named Llama 2 \cite{llama}, which has been trained using 40\% more data than its predecessor Llama 1, and according to \cite{meta} has double the context length.
 In the experiments presented below, we use the smallest version of Llama2, with 7 billion parameters, which we download to run locally and reduce the bottleneck effect of having to call the API-version.
-The reason for not using any of the larger Llama2 models is simply the hardware limitation we face, with the 7B model requiring 19GB of VRAM to be run.
+The reason for not using any of the larger Llama 2 models is simply the hardware limitation we face, with the 7B model requiring 19GB of VRAM to be run.
 
@@
 Example instances of these environments can be found in the appendix.
-For each environment we test our three different models; PPO with no LLM integration, PPO with LLM reward shaping and PPO with LLM policy integration.
-All methods use the same hyperparameters for the PPO, which can be found the appendix.
-For each of these methods, we run PPO for 2000 episodes of 1024 steps, and track the reward\footnote{The true environment reward, not including pseudo-rewards given by the LLM} received for each one.
-We do this eight times per method, and compare the average results against eachother.
+For each environment we test our three different models; PPO with no LLM integration, PPO with LLM reward shaping and PPO with LLM policy optimization ***CONSIDER ALSO TESTING REWARD SHAPER + CONSIGLIERE TOGETHER***.
+All methods use the same hyperparameters for the PPO, which can be found in the appendix.
+For each of these methods, we run PPO for 2000 episodes of 1024 steps, and track the reward\footnote{The true environment reward, not including pseudo-rewards given by the LLM in reward shaping} received for each one.
+We do this eight times per method, and compare the average results against each other.
 
@@
 
-As we can see from figure \ref{doorkeyresults}, LLM integration gives a noticable improvement to the performance of the model, not just in the early stages of training, but throughout the entire training process.
+As we can see from Figure \ref{doorkeyresults}, LLM integration gives a noticable improvement to the performance of the model, not just in the early stages of training, but throughout the entire training process.
 Averaging over all episodes of all eight runs per model, we find that LLM reward shaping receives $24\%$ more reward per episode than PPO trained without LLM rewards.
@@
 Given that there is no reward given for a partially completed task, this is not unexpected.
+
+We also notice a significant performance increase when using LLM policy optimization, even somewhat more than from reward shaping.
+Averaging over all episodes of all runs, we observe a NUMBER\% increase in reward per episode compared to the PPO baseline, and a NUMBER\% increase when compared to LLM policy shaping.
+The standard deviation when using LLM policy optimization is also similarly large to PPO and reward shaping (NUMBER).
+By caching LLM responses similarly to when performing reward shaping, this method too adds very little computational cost to the training in total.
+
+***ADD PARAGRAPH ABOUT USING BOTH TOGETHER*** However, the results from both of these methods are eclipsed by the results from combining LLM reward shaping and policy optimization together.
+Averaging over 8 runs, we observe a NUMBER\% increase in reward from the environment compared to PPO alone, NUMBER\% higher than reward shaping and NUMBER\%higher than policy optimization, with a similar standard deviation of NUMBER.
 
@@
 
+\subsection{Unlock Pickup environment}
+
+The Unlock Pickup environment is an even more complex environment than DoorKey, thus giving rewards to the agent even more sparsely.
+As may be anticipated, this added complexity to the state space makes the LLM contribution even more significant to the RL performance.
+
+Compared to the baseline PPO when averaged per episode over all 8 runs, LLM reward shaping gave a NUMBER\% increase in reward from the environment, with a standard deviation of NUMBER (PPO alone had a standard deviation of NUMBER).
+LLM policy optimization gave a NUMBER\% increase compared to the PPO baseline and a NUMBER\% increase compared to reward shaping.
+Also here giving the highest average reward per episode, the combination of reward shaping and policy optimization gave NUMBER\% more reward compared to the PPO baseline, NUMBER\% more than reward shaping and NUMBER\% higher than policy optimization.
+
 \section{Conclusion}
@@
 It works man, dogski did it!
+
+As shown in the previously referenced work (\ref{background}), the integration of a Large Language Model into the Reinforcement Learning framework can improve network performance.
+In the Minigrid environments we have tested on, DoorKey and Pickup Unlock, we have demonstrated so.
+This has been done both through the LLM shaping the RL reward function, and through the LLM influencing the action probabilities of the RL agent, as well as through combining these two methodologies.
+
+For the DoorKey environment, we found that all three methods surpassed the performance of our PPO baseline.
 
@@
 In the further development of this and similar algorithms, we also consider it essential to put effort into trying to understand the causality behind certain actions being taken, as is being done in the field of explainable artificial intelligence (XAI).
-When involving one major ML framework like that of LLMs into RL, we would ideally want to see exactly where and how much the LLM contributes to the RL networks being updated.
+When involving a major ML framework like that of LLMs into RL, we would ideally want to see exactly where and how much the LLM contributes to the RL networks being updated.
 This would, in turn, help in the work to mitigate whatever biases and variance that may be added to the RL framework from integrating the LLM.

### 2023-10-31 src/ActorCritic.py
Add stuff
Signed-off-by: Jonatan Hanssen <jonatan.hanssen@outlook.com>

@@ -70,10 +70,10 @@ class Agent(nn.Module):
         )
 
         goal = env.reset()[0]["mission"]
-        print(goal)
         if not consigliere:
             self.consigliere = llama2_policy(goal, cos_sim_threshold=0, similarity_modifier=0.1) if llama else None
-        else self.consigliere = consigliere
+        else:
+            self.consigliere = consigliere
 
 
     def get_value(self, observation):

### 2023-10-31 src/utils.py

@@ -157,7 +157,6 @@ class llama2_base:
 
     def reset_cache(self):
         self.cache = dict()
-        self.sim_cache = dict()
 
     def save_cache(self):
         if self.cache_file:

### 2023-11-01 src/utils.py
smol bean
Signed-off-by: Jonatan Hanssen <jonatan.hanssen@outlook.com>

@@ -130,7 +130,6 @@ class llama2_base:
 
         self.dialog.append({"role": "user", "content": observation})
 
-
         if observation in self.cache.keys():
             answer = self.cache[observation]
 
@@ -188,7 +187,6 @@ class llama2_base:
         # print(f"{caption=}")
         best_suggestion = None
         for suggestion in self.suggestions.splitlines():
-
             if caption + suggestion in self.sim_cache.keys():
                 cos_sim = self.sim_cache[caption + suggestion]
 
@@ -280,7 +278,7 @@ def caption_action(action, obs_matrix):
 
             if inventory[0] > 3:
                 caption += f" with{inventory_state} {inventory_color} {inventory_item}"
-            
+
             caption += "."
 
     return caption
@@ -299,7 +297,14 @@ class llama2_reward_shaper(llama2_base):
         sim_cache_file="sim_cache_reward.json",
     ):
         super().__init__(
-            goal, cos_sim_threshold, similarity_modifier, temperature, top_p, rl_temp, cache_file, sim_cache_file
+            goal,
+            cos_sim_threshold,
+            similarity_modifier,
+            temperature,
+            top_p,
+            rl_temp,
+            cache_file,
+            sim_cache_file,
         )
 
 
@@ -316,7 +321,14 @@ class llama2_policy(llama2_base):
         sim_cache_file="sim_cache_policy.json",
     ):
         super().__init__(
-            goal, cos_sim_threshold, similarity_modifier, temperature, top_p, rl_temp, cache_file, sim_cache_file
+            goal,
+            cos_sim_threshold,
+            similarity_modifier,
+            temperature,
+            top_p,
+            rl_temp,
+            cache_file,
+            sim_cache_file,
         )
 
         self.dialog = [
@@ -372,7 +384,6 @@ class llama2_policy(llama2_base):
                 "role": "assistant",
                 "content": "pick up red key.",
             },
-
         ]
 
     def compare(self, action, obs_matrix):
@@ -386,10 +397,6 @@ class llama2_policy(llama2_base):
             a reward if action and one of the suggested actions is semantically similar
         """
         caption = caption_action(action, obs_matrix)
-
-
-
-
 
         max_cos_sim = 0
         # print(f"{caption=}")
@@ -416,12 +423,9 @@ class llama2_policy(llama2_base):
             # print(f"{best_suggestion=}")
             # print()
 
-
             return max_cos_sim * self.similarity_modifier
[…]

### 2023-11-01 src/utils.py
Smool been fix
Signed-off-by: Jonatan Hanssen <jonatan.hanssen@outlook.com>

@@ -469,23 +469,23 @@ def obs_to_string(obs_matrix, positions=True, you=True):
             string = f"see a{state} {color} {item}"
 
             if positions:
-                if rel_y == 0:
+                if rel_x == 0:
                     longitude = ""
 
-                elif rel_y > 0:
-                    longitude = f"{rel_y} square{'s' if rel_y != 1 else ''} RIGHT"
+                elif rel_x > 0:
+                    longitude = f"{rel_x} square{'s' if rel_x != 1 else ''} RIGHT"
 
                 else:
                     longitude = (
-                        f"{abs(rel_y)} square{'s' if abs(rel_y) != 1 else ''} LEFT"
+                        f"{abs(rel_x)} square{'s' if abs(rel_x) != 1 else ''} LEFT"
                     )
 
-                if rel_x == 0:
+                if rel_y == 0:
                     latitude = ""
 
                 else:
                     latitude = (
-                        f"{abs(rel_x)} square{'s' if abs(rel_x) != 1 else ''} FORWARD"
+                        f"{abs(rel_y)} square{'s' if abs(rel_y) != 1 else ''} FORWARD"
                     )
 
                 if longitude and latitude:

### 2023-11-01 src/utils.py
Prepare for merge

@@ -232,22 +232,22 @@ def caption_action(action, obs_matrix):
 
     ###  Captioning action based on cell and inventory ###
 
-    caption = "do nothing."
+    caption = "do nothing"
 
     if action == "done":
-        caption = "mark the game as completed."
+        caption = "mark the game as completed"
 
     if action == "forward":
         if front_item == "empty":
-            caption = "move forward."
+            caption = "move forward"
         elif front_item == "door" and front_state == " open":
-            caption = "move forward."
+            caption = "move forward"
 
     if action == "right":
-        caption = "turn right."
+        caption = "turn right"
 
     if action == "left":
-        caption = "turn left."
+        caption = "turn left"
 
     if action == "pick":
         if front_cell[0] < 4:  # nothing to pick up
@@ -257,7 +257,7 @@ def caption_action(action, obs_matrix):
             pass
 
         else:
-            caption = f"pick up{front_state} {front_color} {front_item}."
+            caption = f"pick up{front_state} {front_color} {front_item}"
 
     elif action == "drop":
         if inventory[0] < 4:  # nothing to drop
@@ -267,19 +267,19 @@ def caption_action(action, obs_matrix):
             pass
 
         else:
-            caption = f"drop{inventory_state} {inventory_color} {inventory_item}."
+            caption = f"drop{inventory_state} {inventory_color} {inventory_item}"
 
     elif action == "toggle":
         if front_item == "box":
-            caption = f"destroy {front_color} box."
+            caption = f"destroy {front_color} box"
         else:
             if front_cell[0] >= 4:
                 caption = f"use{front_state} {front_color} {front_item}"
 
             if inventory[0] > 3:
                 caption += f" with{inventory_state} {inventory_color} {inventory_item}"
-
-            caption += "."
+            
+    caption += "."
 
     return caption
 
@@ -334,11 +334,11 @@ class llama2_policy(llama2_base):
         self.dialog = [
             {
                 "role": "system",
[…]

### 2023-11-01 src/utils.py
Longditude dongtidude
Signed-off-by: Jonatan Hanssen <jonatan.hanssen@outlook.com>

@@ -278,7 +278,7 @@ def caption_action(action, obs_matrix):
 
             if inventory[0] > 3:
                 caption += f" with{inventory_state} {inventory_color} {inventory_item}"
-            
+
     caption += "."
 
     return caption
@@ -506,7 +506,10 @@ def obs_to_string(obs_matrix, positions=True, you=True):
                     string += f" {longitude}"
 
                 elif not longitude and latitude:
-                    string += f" {latitude}"
+                    if rel_y == -1:
+                        string += " directly in front"
+                    else:
+                        string += f" {latitude}"
 
             observation_strings.append(string)
 

### 2023-11-01 src/multi_ppo.py
Add multi_ppo
Signed-off-by: Jonatan Hanssen <jonatan.hanssen@outlook.com>

@@ -1,4 +1,78 @@
 from ppo import PPO
 from utils import llama2_policy
 
-consigliere = llama2_policy(goal, cos_sim_threshold=0, similarity_modifier=0.1)
+consigliere = llama2_policy("use the key to open the door and then get to the goal", cos_sim_threshold=0, similarity_modifier=0.1)
+
+print("Starting policy 2")
+ppo = PPO(
+        "hyperparams.json",
+        result_file="data/doorkey_big_policy2",
+        reward=False,
+        policy=True,
+        consigliere=consigliere,
+    )
+ppo.agent.consigliere.reset_cache()
+ppo.train()
+
+print("Starting policy 3")
+ppo = PPO(
+        "hyperparams.json",
+        result_file="data/doorkey_big_policy3",
+        reward=False,
+        policy=True,
+        consigliere=consigliere,
+    )
+ppo.agent.consigliere.reset_cache()
+ppo.train()
+
+print("Starting policy 4")
+ppo = PPO(
+        "hyperparams.json",
+        result_file="data/doorkey_big_policy4",
+        reward=False,
+        policy=True,
+        consigliere=consigliere,
+    )
+ppo.agent.consigliere.reset_cache()
+ppo.train()
+
+print("Starting policy 5")
+ppo = PPO(
+        "hyperparams.json",
+        result_file="data/doorkey_big_policy5",
+        reward=False,
+        policy=True,
+        consigliere=consigliere,
+    )
+ppo.agent.consigliere.reset_cache()
+ppo.train()
+
+print("Starting policy 6")
+ppo = PPO(
+        "hyperparams.json",
+        result_file="data/doorkey_big_policy6",
+        reward=False,
+        policy=True,
+        consigliere=consigliere,
+    )
+ppo.train()
+
+print("Starting policy 7")
+ppo = PPO(
+        "hyperparams.json",
+        result_file="data/doorkey_big_policy7",
+        reward=False,
+        policy=True,
+        consigliere=consigliere,
+    )
+ppo.train()
+
+print("Starting policy 8")
+ppo = PPO(
+        "hyperparams.json",
+        result_file="data/doorkey_big_policy8",
+        reward=False,
+        policy=True,
+        consigliere=consigliere,
+    )
+ppo.train()

### 2023-11-01 src/ActorCritic.py
add softmax for logits in llm policy

@@ -71,7 +71,7 @@ class Agent(nn.Module):
 
         goal = env.reset()[0]["mission"]
         if not consigliere:
-            self.consigliere = llama2_policy(goal, cos_sim_threshold=0, similarity_modifier=0.1) if llama else None
+            self.consigliere = llama2_policy(goal, cos_sim_threshold=0.7, similarity_modifier=0.1) if llama else None
         else:
             self.consigliere = consigliere
 
@@ -92,8 +92,7 @@ class Agent(nn.Module):
 
         if self.consigliere:
             #print(observation.shape)
-            #logits = torch.nn.functional.softmax(logits, dim=-1)
-            #logits /= torch.norm(logits)
+            # logits /= torch.norm(logits)
             if len(observation.shape) == 2:
                 advisor_values_list = list()
                 for single_obs in observation:
@@ -112,7 +111,7 @@ class Agent(nn.Module):
             # print(f"{logits=}")
             # print(f"{advisor_values=}")
             # # anti adrian propaganda
-            logits += advisor_values
+            # logits += advisor_values
             # adrian good vote adrian
             # logits *= advisor_values
 
@@ -121,7 +120,12 @@ class Agent(nn.Module):
         # print(probs.probs)
 
         if action is None:
-            action = probs.sample()
+            if self.consigliere:
+                newprobs = torch.nn.functional.softmax(probs.probs + advisor_values, dim=-1)
+                newprobs = Categorical(logits=logits)
+                action = newprobs.sample()
+            else:
+                action = probs.sample()
         return (
             action,
             probs.log_prob(action),

### 2023-11-01 src/multi_ppo.py

@@ -3,10 +3,21 @@ from utils import llama2_policy
 
 consigliere = llama2_policy("use the key to open the door and then get to the goal", cos_sim_threshold=0, similarity_modifier=0.1)
 
+print("Starting policy 1")
+ppo = PPO(
+        "hyperparams.json",
+        result_file="data/doorkey_policy1",
+        reward=False,
+        policy=True,
+        consigliere=consigliere,
+    )
+ppo.agent.consigliere.reset_cache()
+ppo.train()
+
 print("Starting policy 2")
 ppo = PPO(
         "hyperparams.json",
-        result_file="data/doorkey_big_policy2",
+        result_file="data/doorkey_policy2",
         reward=False,
         policy=True,
         consigliere=consigliere,
@@ -17,7 +28,7 @@ ppo.train()
 print("Starting policy 3")
 ppo = PPO(
         "hyperparams.json",
-        result_file="data/doorkey_big_policy3",
+        result_file="data/doorkey_policy3",
         reward=False,
         policy=True,
         consigliere=consigliere,
@@ -28,7 +39,7 @@ ppo.train()
 print("Starting policy 4")
 ppo = PPO(
         "hyperparams.json",
-        result_file="data/doorkey_big_policy4",
+        result_file="data/doorkey_policy4",
         reward=False,
         policy=True,
         consigliere=consigliere,
@@ -39,7 +50,7 @@ ppo.train()
 print("Starting policy 5")
 ppo = PPO(
         "hyperparams.json",
-        result_file="data/doorkey_big_policy5",
+        result_file="data/doorkey_policy5",
         reward=False,
         policy=True,
         consigliere=consigliere,
@@ -50,7 +61,7 @@ ppo.train()
 print("Starting policy 6")
 ppo = PPO(
         "hyperparams.json",
-        result_file="data/doorkey_big_policy6",
+        result_file="data/doorkey_policy6",
         reward=False,
         policy=True,
         consigliere=consigliere,
@@ -60,7 +71,7 @@ ppo.train()
 print("Starting policy 7")
 ppo = PPO(
         "hyperparams.json",
-        result_file="data/doorkey_big_policy7",
+        result_file="data/doorkey_policy7",
         reward=False,
         policy=True,
         consigliere=consigliere,
@@ -70,7 +81,7 @@ ppo.train()
 print("Starting policy 8")
 ppo = PPO(
         "hyperparams.json",
-        result_file="data/doorkey_big_policy8",
[…]

### 2023-11-01 src/utils.py

@@ -362,7 +362,7 @@ class llama2_policy(llama2_base):
             },
             {
                 "role": "user",
-                "content": "My goal is: open the purple door. I see a locked purple door FORWARD and have a purple key in my inventory. What actions do you suggest?",
+                "content": "My goal is: open the purple door. I see a locked purple door directly in front and have a purple key in my inventory. What actions do you suggest?",
             },
             {
                 "role": "assistant",
@@ -370,7 +370,7 @@ class llama2_policy(llama2_base):
             },
             {
                 "role": "user",
-                "content": "My goal is: open the purple door. I see a locked purple door FORWARD. What actions do you suggest?",
+                "content": "My goal is: open the purple door. I see a locked purple door directly in front. What actions do you suggest?",
             },
             {
                 "role": "assistant",
@@ -386,7 +386,7 @@ class llama2_policy(llama2_base):
             },
             {
                 "role": "user",
-                "content": "My goal is: open the red door. I see a locked red door RIGHT and FORWARD and a red key FORWARD. What actions do you suggest?",
+                "content": "My goal is: open the red door. I see a locked red door FORWARD and a red key directly in front. What actions do you suggest?",
             },
             {
                 "role": "assistant",

### 2023-11-02 src/ActorCritic.py
Turd time's the charm
Signed-off-by: Jonatan Hanssen <jonatan.hanssen@outlook.com>

@@ -71,7 +71,7 @@ class Agent(nn.Module):
 
         goal = env.reset()[0]["mission"]
         if not consigliere:
-            self.consigliere = llama2_policy(goal, cos_sim_threshold=0.7, similarity_modifier=0.1) if llama else None
+            self.consigliere = llama2_policy(goal, cos_sim_threshold=0.7, similarity_modifier=0.3) if llama else None
         else:
             self.consigliere = consigliere
 
@@ -106,7 +106,7 @@ class Agent(nn.Module):
             else:
                 unflat_obs = observation.reshape((7, 7, 3)).to(torch.int64)
                 #self.consigliere.suggest(unflat_obs)
-                advisor_values = self.consigliere.give_values(np.array(unflat_obs.cpu()))
+                advisor_values = torch.tensor(self.consigliere.give_values(np.array(unflat_obs.cpu())), dtype=torch.float64)
 
             # print(f"{logits=}")
             # print(f"{advisor_values=}")
@@ -121,9 +121,18 @@ class Agent(nn.Module):
 
         if action is None:
             if self.consigliere:
-                newprobs = torch.nn.functional.softmax(probs.probs + advisor_values, dim=-1)
-                newprobs = Categorical(logits=logits)
-                action = newprobs.sample()
+                if torch.norm(advisor_values) == 0:
+                    action = probs.sample()
+                else:
+                    scale_factor = (torch.norm(logits) / torch.norm(advisor_values))
+                    # print(f"{advisor_values=}")
+                    # print(f"{scale_factor=}")
+                    # print(f"{advisor_values * scale_factor=}")
+                    # print(f"{logits=}")
+                    newprobs = Categorical(logits=logits + advisor_values * scale_factor)
+                    action = newprobs.sample()
+                    # print(f"{probs.probs=}")
+                    # print(f"{newprobs.probs=}")
             else:
                 action = probs.sample()
         return (

### 2023-11-02 src/ppo.py

@@ -26,8 +26,8 @@ class PPO:
         self.results_file = (
             self.args["results_file"] if not result_file else result_file
         )
-        self.llama_policy = self.args["llama_policy"] if not policy else policy
-        self.llama_reward = self.args["llama_reward"] if not reward else reward
+        self.llama_policy = self.args["llama_policy"] if policy is None else policy
+        self.llama_reward = self.args["llama_reward"] if reward is None else reward
 
         self.env = gym.make(
             self.args["env_name"], render_mode="rgb_array", max_steps=self.args["steps"]
@@ -287,10 +287,10 @@ class PPO:
 if __name__ == "__main__":
     parser = argparse.ArgumentParser()
     parser.add_argument(
-        "-f", "--result_file", type=str, help="Output file for env rewards"
+        "-f", "--result_file", type=str, help="Output file for env rewards", default=None
     )
-    parser.add_argument("-r", "--reward", type=bool, help="Use LLM reward shaping")
-    parser.add_argument("-p", "--policy", type=bool, help="Use LLM policy")
+    parser.add_argument("-r", "--reward", type=bool, help="Use LLM reward shaping", default=None)
+    parser.add_argument("-p", "--policy", type=bool, help="Use LLM policy", default=None)
     args = parser.parse_args()
     print(args)
 

### 2023-11-02 src/utils.py

@@ -179,7 +179,7 @@ class llama2_base:
         caption = caption_action(action, obs_matrix)
 
         if not caption or caption in self.caption_set:
-            return 0
+            return 0.0
 
         self.caption_set.add(caption)
 
@@ -208,7 +208,7 @@ class llama2_base:
 
             return max_cos_sim * self.similarity_modifier
         else:
-            return 0
+            return 0.0
 
 
 def caption_action(action, obs_matrix):

### 2023-11-02 src/ActorCritic.py
Add annealing
Signed-off-by: Jonatan Hanssen <jonatan.hanssen@outlook.com>

@@ -71,7 +71,7 @@ class Agent(nn.Module):
 
         goal = env.reset()[0]["mission"]
         if not consigliere:
-            self.consigliere = llama2_policy(goal, cos_sim_threshold=0.7, similarity_modifier=0.3) if llama else None
+            self.consigliere = llama2_policy(goal, cos_sim_threshold=0.7, similarity_modifier=1) if llama else None
         else:
             self.consigliere = consigliere
 
@@ -82,7 +82,7 @@ class Agent(nn.Module):
         ).flatten()
         return self.critic(observation.to(torch.float64))
 
-    def get_action_and_value(self, observation, action=None):
+    def get_action_and_value(self, observation, action=None, rollout=None):
         observation_onehot = (
             nn.functional.one_hot(observation.to(torch.int64), num_classes=11)
             .to(torch.float64)
@@ -106,7 +106,7 @@ class Agent(nn.Module):
             else:
                 unflat_obs = observation.reshape((7, 7, 3)).to(torch.int64)
                 #self.consigliere.suggest(unflat_obs)
-                advisor_values = torch.tensor(self.consigliere.give_values(np.array(unflat_obs.cpu())), dtype=torch.float64)
+                advisor_values = self.consigliere.give_values(np.array(unflat_obs.cpu())).to(torch.float64)
 
             # print(f"{logits=}")
             # print(f"{advisor_values=}")
@@ -124,12 +124,17 @@ class Agent(nn.Module):
                 if torch.norm(advisor_values) == 0:
                     action = probs.sample()
                 else:
+                    anneal = ((500 - rollout) / 500) ** 2 if rollout else 1
+                    if anneal < 0:
+                        anneal = 0
+                    # print(anneal)
                     scale_factor = (torch.norm(logits) / torch.norm(advisor_values))
                     # print(f"{advisor_values=}")
                     # print(f"{scale_factor=}")
                     # print(f"{advisor_values * scale_factor=}")
                     # print(f"{logits=}")
-                    newprobs = Categorical(logits=logits + advisor_values * scale_factor)
+                    advisor_values = advisor_values.to(torch.device("cuda"))
+                    newprobs = Categorical(logits=logits + advisor_values * scale_factor * anneal)
                     action = newprobs.sample()
                     # print(f"{probs.probs=}")
                     # print(f"{newprobs.probs=}")

### 2023-11-02 src/ppo.py

@@ -79,7 +79,7 @@ class PPO:
             (
                 env_reward,
                 advisor_reward_cum,
-            ) = self.next_episode()  # Perform steps and store relevant values
+            ) = self.next_episode(rollout)  # Perform steps and store relevant values
             rewards.append(env_reward)
             # if not env_reward:
             #     print("No reward received, skipping update of networks")
@@ -109,7 +109,7 @@ class PPO:
 
             i += 1
 
-    def next_episode(self):
+    def next_episode(self, rollout):
         """Steps through a single episode and calculates what is needed to
         perform PPO update
         """
@@ -124,7 +124,7 @@ class PPO:
             self.observations[step] = observation
 
             with torch.no_grad():
-                action, logprob, _, value = self.agent.get_action_and_value(observation)
+                action, logprob, _, value = self.agent.get_action_and_value(observation, rollout=rollout)
 
             self.values[step] = value.flatten()
             self.actions[step] = action

### 2023-11-02 src/ActorCritic.py
Fix anneal
Signed-off-by: Jonatan Hanssen <jonatan.hanssen@outlook.com>

@@ -124,9 +124,13 @@ class Agent(nn.Module):
                 if torch.norm(advisor_values) == 0:
                     action = probs.sample()
                 else:
-                    anneal = ((500 - rollout) / 500) ** 2 if rollout else 1
-                    if anneal < 0:
-                        anneal = 0
+                    max_rollout = 500
+                    if rollout:
+                        anneal = ((max_rollout - rollout) / max_rollout) ** 2
+                        if rollout > max_rollout:
+                            anneal = 0
+                    else:
+                        anneal = 1
                     # print(anneal)
                     scale_factor = (torch.norm(logits) / torch.norm(advisor_values))
                     # print(f"{advisor_values=}")

### 2023-11-03 src/utils.py
added more examples for llm policy

@@ -362,7 +362,7 @@ class llama2_policy(llama2_base):
             },
             {
                 "role": "user",
-                "content": "My goal is: open the purple door. I see a locked purple door directly in front and have a purple key in my inventory. What actions do you suggest?",
+                "content": "My goal is: use the key to open the door and get to the goal. I see a locked purple door directly in front and have a purple key in my inventory. What actions do you suggest?",
             },
             {
                 "role": "assistant",
@@ -370,7 +370,7 @@ class llama2_policy(llama2_base):
             },
             {
                 "role": "user",
-                "content": "My goal is: open the purple door. I see a locked purple door directly in front. What actions do you suggest?",
+                "content": "My goal is: use the key to open the door and get to the goal. I see a locked purple door directly in front. What actions do you suggest?",
             },
             {
                 "role": "assistant",
@@ -386,11 +386,35 @@ class llama2_policy(llama2_base):
             },
             {
                 "role": "user",
-                "content": "My goal is: open the red door. I see a locked red door FORWARD and a red key directly in front. What actions do you suggest?",
+                "content": "My goal is: use the key to open the door and get to the goal. I see a locked red door FORWARD and a red key directly in front. What actions do you suggest?",
             },
             {
                 "role": "assistant",
                 "content": "pick up red key.",
+            },
+            {
+                "role": "user",
+                "content": "My goal is: use the key to open the door and get to the goal. I see a green goal directly in front. What actions do you suggest?",
+            },
+            {
+                "role": "assistant",
+                "content": "move forward.",
+            },
+            {
+                "role": "user",
+                "content": "My goal is: use the key to open the door and get to the goal. I see a green goal RIGHT and a locked blue door directly in front and have a blue key in my inventory. What actions do you suggest?",
+            },
+            {
+                "role": "assistant",
+                "content": "turn right.",
+            },
+            {
+                "role": "user",
+                "content": "My goal is: use the key to open the door and get to the goal. I see a green goal LEFT and a yellow key FORWARD and a locked yellow door FORWARD. What actions do you suggest?",
+            },
+            {
+                "role": "assistant",
+                "content": "turn left.",
             },
         ]
 

### 2023-11-03 doc/paper.tex
new stuff in paper

@@
 \begin{IEEEkeywords}
-large language models, reinforcement learning, minigrid, proximal policy optimization
+large language models, reinforcement learning, minigrid, proximal policy influencing
 \end{IEEEkeywords}
@@
 After sorting available tasks into interesting and boring ones, it assigns sampling weights to them, giving much higher weights to interesting tasks than boring ones, thus directly influencing the agent's policy by altering probabilities of which action it then chooses to do.
-In \cite{idm}, another paper proposing using the LLM as a policy optimizer, has a similar approach in which the state observation is tokenized and passed to the LLM, though in their experiment, goals and action histories are also tokenized and passed.
+In \cite{idm}, another paper proposing using the LLM as a policy influencer, has a similar approach in which the state observation is tokenized and passed to the LLM, though in their experiment, goals and action histories are also tokenized and passed.
 Here, too, the LLM response is fed back into a task-specific RL model, which in turn influences policy probabilities of certain actions being taken.
@@
 Example instances of these environments can be found in the appendix.
-For each environment we test our three different models; PPO with no LLM integration, PPO with LLM reward shaping and PPO with LLM policy optimization ***CONSIDER ALSO TESTING REWARD SHAPER + CONSIGLIERE TOGETHER***.
+For each environment we test our three different models; PPO with no LLM integration, PPO with LLM reward shaping and PPO with LLM policy influencing ***CONSIDER ALSO TESTING REWARD SHAPER + CONSIGLIERE TOGETHER***.
 All methods use the same hyperparameters for the PPO, which can be found in the appendix.
@@
 
-We also notice a significant performance increase when using LLM policy optimization, even somewhat more than from reward shaping.
-Averaging over all episodes of all runs, we observe a NUMBER\% increase in reward per episode compared to the PPO baseline, and a NUMBER\% increase when compared to LLM policy shaping.
-The standard deviation when using LLM policy optimization is also similarly large to PPO and reward shaping (NUMBER).
+% We also notice a significant performance increase when using LLM policy influencing, even somewhat more than from reward shaping.
+Averaging over all episodes of all runs, we observe a NUMBER\% increase in reward per episode compared to the PPO baseline, and a NUMBER\% increase when compared to LLM reward shaping.
+The standard deviation when using LLM policy influencing is also similarly large to PPO and reward shaping (NUMBER).
 By caching LLM responses similarly to when performing reward shaping, this method too adds very little computational cost to the training in total.
 
-***ADD PARAGRAPH ABOUT USING BOTH TOGETHER*** However, the results from both of these methods are eclipsed by the results from combining LLM reward shaping and policy optimization together.
-Averaging over 8 runs, we observe a NUMBER\% increase in reward from the environment compared to PPO alone, NUMBER\% higher than reward shaping and NUMBER\%higher than policy optimization, with a similar standard deviation of NUMBER.
+Using LLM policy influencing in the DoorKey environment has resulted in somewhat worse results than using the PPO baseline.
+Averaging per episode over all eight runs shows that LLM policy influencing only achieves $89\%$ of the reward amount per episode that the baseline achieves.
+Judging from the graph, it may seem that policy influencing may have given some initial learning advantage, but ultimately slows learning for the agent further on.
+The method also gives the highest standard deviation of reward given per episode of the three compared so far, at $0.25$.
+As with reward shaping, we have also used caching of LLM responses to significantly decrease the computational overhead cost to a minimum.
+
+***TODO: add paragraph about using both together*** 
+% However, the results from both of these methods are eclipsed by the results from combining LLM reward shaping and policy influencing together.
+Averaging over 8 runs, we observe a NUMBER\% increase in reward from the environment compared to PPO alone, NUMBER\% higher than reward shaping and NUMBER\%higher than policy influencing, with a similar standard deviation of NUMBER.
 
@@
 Compared to the baseline PPO when averaged per episode over all 8 runs, LLM reward shaping gave a NUMBER\% increase in reward from the environment, with a standard deviation of NUMBER (PPO alone had a standard deviation of NUMBER).
-LLM policy optimization gave a NUMBER\% increase compared to the PPO baseline and a NUMBER\% increase compared to reward shaping.
-Also here giving the highest average reward per episode, the combination of reward shaping and policy optimization gave NUMBER\% more reward compared to the PPO baseline, NUMBER\% more than reward shaping and NUMBER\% higher than policy optimization.
+LLM policy influencing gave a NUMBER\% increase compared to the PPO baseline and a NUMBER\% increase compared to reward shaping.
+Also here giving the highest average reward per episode, the combination of reward shaping and policy influencing gave NUMBER\% more reward compared to the PPO baseline, NUMBER\% more than reward shaping and NUMBER\% higher than policy influencing.
 
-\section{Conclusion}
+\section{Discussion, Future Work and Conclusion}
 
-It works man, dogski did it!
+% It works man, dogski did it!
 
-As shown in the previously referenced work (\ref{background}), the integration of a Large Language Model into the Reinforcement Learning framework can improve network performance.
+% As shown in the previously referenced work (\ref{background}), the integration of a Large Language Model into the Reinforcement Learning framework can improve network performance.
 In the Minigrid environments we have tested on, DoorKey and Pickup Unlock, we have demonstrated so.
@@
 
-For the DoorKey environment, we found that all three methods surpassed the performance of our PPO baseline.
+% For the DoorKey environment, we found that all three methods surpassed the performance of our PPO baseline.
+
+As shown in the previously referenced work (\ref{background}), the integration of a Large Language Model into the Reinforcement Learning framework can improve network performance.
+There are various ways to do this, and our work attempts to compare two main methodologies of integrating the LLM, namely through reward shaping and policy influencing.
+We have implemented algorithms to perform this on top of a PPO baseline, and have measured varying results across environments of different complexities.
+
+Our reward shaping algorithm did best in the more complex environments (DoorKey and Unlock Pickup), while policy influencing performed better in the simpler Empty environment.
+A possible explanation for this could lie in the details of how the algorithms impact choice of action.
+LLM reward shaping does not directly impact decision making at all; it only impacts the reward function, meaning the mechanism that dictates which action is taken is still solely the actor network present in PPO.
+This means that reward shaping does not introduce any risk of the agent choosing a suboptimal action directly due to being told to do so by the LLM.
+Learning which action to take is still done through the same process of trial and error as in classical RL.
+
+However, the causality behind which action is taken is different in LLM policy influencing; the LLM suggestions are added directly to the logits used to calculate probabilities of each action, meaning that the LLM directly influences the decision making process.
+Every time the LLM suggests performing a suboptimal action, this impacts the action probabilities, regardless whether the PPO baseline has learned that another action is better or not.
+Over time, this may lead to the PPO not being able to explore other actions than the ones proposed by the LLM, thus making the PPO learn slower than if it was allowed to choose actions on its own.
+In other words, the LLM may have too much impact in the decision making when performing policy influencing, making the algorithm prone to errors from the LLM.
+
+A natural suggestion for how to improve the performance of LLM policy influencing in the future thus becomes to decrease the influence from the LLM over time, letting the PPO choose on its own as learning progresses.
+One way of doing this is through annealing of the LLM values added to the logits.
+This can be combined with experimenting with a hyperparameter determining how the LLM values should be weighted at the beginning of training.
+We have attempted to do this, but have encountered programmatic issues regarding scaling; the logits change in magnitude over time, and a fitting way of dynamically changing the LLM contributions to the logits has to be found.
 

### 2023-11-03 doc/paper.tex
Update paper
Signed-off-by: Jonatan Hanssen <jonatan.hanssen@outlook.com>

--- 2023-10-31 doc/paper.tex
@@
 
-To evaluate the effectiveness of the integration of LLMs into Reinforcement Learning algorithms, we use the \textbf{MiniGrid} and \textbf{BabyAI} libraries.
-Here, we handpick specific environments with varying levels of complexity, to evaluate the effectiveness of our proposed techniques against a standard Actor-Critic \ref{Proximal policy optimzation}, and against each other.
-The environments selected are all based in a small sized grid-world, where the agent is surrounded by walls, and has to perform at least one task to complete the game.
+To evaluate the effectiveness of the integration of LLMs into Reinforcement Learning algorithms, we use the MiniGrid library.
+Here, we handpick specific environments with varying levels of complexity, to evaluate the effectiveness of our proposed techniques against a standard Actor-Critic PPO, and against each other.
+% The environments selected are all based in a small sized grid-world, where the agent is surrounded by walls, and has to perform at least one task to complete the game.
 How difficult an environment is to beat depends on factors such as the shape of the walls, whether the agent has to solve subtasks to finish the game, and the types of subtasks encountered.
@@
 
-Our baseline is the \textbf{Proximal Policy Optimization} (PPO) algorithm, which has been established as a state-of-art approach to solving reinforcement learning problems over the last few years.
-The main idea behind this algorithm is that the new policy is clipped so that it does not move too quickly and too far from the old policy, i.e.
-the new policy cannot diverge too much from the old policy.
-This helps to avoid exceedingly large updates to policy gradients, thus also avoiding the problem of vanishing and exploding gradients.
-This has elevated the ability of RL-techniques to tackle sparse environments, and outperform older approches such as DQN.
-For a more detailed discussion of the mathematics and implementation details, see \cite{OpenAIPPO}
+Our baseline is the PPO algorithm, which is a policy gradient method that has been established as a state-of-art approach to solving reinforcement learning problems over the last few years.
+The main idea behind this algorithm is that the new policy is clipped so that it does not move too quickly and too far from the old policy, which helps to avoid changing the policy too drastically during the update.
+With this simple concept, PPO has outperformed previous methods such as Deep-Q Networks and A2C, while being relatively simple to implement\cite{ppo}.
 
@@
 Then, the LLM is used as a tool for semantic similarity comparison, comparing its own formerly proposed action to the action performed by the agent.
-Finally, high semantic similarity gives an additional reward which is implemented in the reward function of the agent.
+Finally, high semantic similarity gives an additional reward which is added to the reward of the environment.
+Compared to the policy influencing method above, this approach is less direct, as the policy is only indirectly affected by the LLM through these additional rewards.
 
-\subsection{Large Language Model used}
+\subsection{Language Models}
 
-For the purpose of our research, we have decided utilize a LLM from the Llama open-source family of language models developed by Meta AI, first released in Febraury, 2023.
-Specifically, we chose the newest of the models named Llama 2 \cite{llama}, which has been trained using 40\% more data than its predecessor Llama 1, and according to \cite{meta} has double the context length.
-In the experiments presented below, we use the smallest version of Llama2, with 7 billion parameters, which we download to run locally and reduce the bottleneck effect of having to call the API-version.
+In both these cases, we need two language models; one to generate the suggestions and one to generate semantic similarity values.
+For text generation, we have decided utilize an LLM from the Llama open-source family of language models developed by Meta AI, first released in Febraury, 2023.
+Specifically, we chose Llama 2, which has been trained using 40\% more data than its predecessor Llama 1, and has double the context length\cite{llama}.
+In the experiments presented below, we use the smallest version of Llama2, with 7 billion parameters, which we download to run locally and reduce the bottleneck effect of having to call an API to a model hosted by someone else.
 The reason for not using any of the larger Llama 2 models is simply the hardware limitation we face, with the 7B model requiring 19GB of VRAM to be run.
 
-Many research papers nowadays propose fine-tuning of LLMs as means of leveraging the capabilites of the language models for decision-making purposes even further.
+Some research papers propose fine-tuning of LLMs as means of leveraging the capabilites of the language models for decision-making purposes even further \cite{grounding}.
 Our plan of action, on the other hand, aims at evaluating the efficacy of the LLM without any previous pre-tuning.
@@
 
+Instead of fine tuning the LLM, we use few-shot learning when prompting, giving the LLM a few examples of observations and correct corresponding action recommendations.
+This is an effective way to guide the LLM for our specific purposes without requiring us to retrain the entire network.
+
+For generating semantic similarity values, we use Sentence-Bert \cite{bert} for generating encoding vectors, and perform a cosine similarity calculation of the two sentence encodings.
+This gives a value between $-1$ and $1$, where two sentences with the exact same encoding have a cosine similarity of $1$.
+
 \section{Experiments}
@@
 We perform experiments to compare different methods of integrating LLMs with the PPO algorithm, specifically comparing their ability to guide the early exploration of an RL model in a reward-sparse environment.
+Even state-of-art RL algorithms may struggle in such environments.
+Therefore, it is the perfect arena to see if the high generalization capabilites of pre-trained Large Language Models will lead to better results over using only RL methods.
 
@@
 
-We select NUMBER different Minigrid environments; DoorKey, OTHERENV, OTHERENV.
-Example instances of these environments can be found in the appendix.
-For each environment we test our three different models; PPO with no LLM integration, PPO with LLM reward shaping and PPO with LLM policy optimization ***CONSIDER ALSO TESTING REWARD SHAPER + CONSIGLIERE TOGETHER***.
+We select three different Minigrid environments with increasing complexity; Empty, DoorKey and UnlockPickup.
+For each environment we test our three different models; PPO with no LLM integration, PPO with LLM reward shaping and PPO with LLM policy influencing.
 All methods use the same hyperparameters for the PPO, which can be found in the appendix.
-For each of these methods, we run PPO for 2000 episodes of 1024 steps, and track the reward\footnote{The true environment reward, not including pseudo-rewards given by the LLM in reward shaping} received for each one.
-We do this eight times per method, and compare the average results against each other.
+For each of these methods, we run PPO for 2000 (200 for Empty) episodes of 1024 steps, and track the reward\footnote{The true environment reward, not including pseudo-rewards given by the LLM in reward shaping} received for each one.
+We do this eight times per method, and compare the average results per episode against each other.
+
+\subsection{Empty Environment}
+
+This environment is the simplest possible, consisting of an empty room with a single goal in the bottom right corner.
+This environment contains no stochasticity; the agent and goal always start in the same positions.
+Despite its simplicity, there is still potential for LLM integration to improve the early phases of training, because the room is 15 by 15 tiles, which means that receiving any reward requires a significant amount of exploration.
+The environment is seen in figure \ref{emptyenv}.
+
+\begin{figure}[h]
+\centerline{\includegraphics[width=3.25in]{figure/emptyenv.png}}
+\caption{The Minigrid Empty Environment}
+\label{emptyenv}
+\end{figure}
+
+\subsubsection{DoorKey environment}
+
+This environment consists of two rooms separated by a locked door.
+Fig \ref{doorkeyenv} shows an example configuration.
+The agent starts in one room, while the goal square is in the other.
+The agent must pick up a key in the starting room, and use it on the door to gain access to the room containing the goal.
+This environment thus contains two subgoals; picking up the key and using it to unlock the door.
+
+\subsubsection{UnlockPickup environment}
+
+This environment is quite similar to the DoorKey environment.
+Fig \ref{unlockpickupenv} shows an example configuration.
+The difference is that after unlocking the door, the agent must now pick up a chest instead of walking to a goal square.
+Because the agent can only hold one item at a time, this requires dropping the key on an empty square, which introduces another subtask.
+Chests can also be opened like doors, but this removes the chest from the game and thus causes and instant failure of the task, as it can no longer be picked up.
+Thus, the agent must learn that using a key on a door is a valuable subtask, while using it on a chest is a bad idea.
+This added complexity makes it a much more difficult task than the previous two.
 
 \section{Results}
+
+\subsection{Empty Environment}
+
+In this simple environment, we see that both LLM reward shaping and LLM policy influencing outperform the baseline model.
+Averaging over all episodes for all eight runs per model, we find LLM policy influencing to perform best, receiving $30\%$ more reward than the baseline.
+LLM reward shaping similarly outperforms the baseline, receiving $10\%$ more reward.
+Given that policy influencing is a more direct approach, it makes sense that this method performs best, as the simple environment makes it likely that the LLM will suggest the best action the majority of the time.
+
+\begin{figure}[h]
+\centerline{\includegraphics[width=3.25in]{figure/emptyresults.png}}
+\caption{Reward per episode for the three different methods in the Empty environment}
+\label{doorkeyresults}
+\end{figure}
 
@@
 \centerline{\includegraphics[width=3.25in]{figure/doorkeyresults.png}}
-\caption{Reward per episode for the three (two) different methods in the DoorKey environment}
+\caption{Reward per episode for the three different methods in the DoorKey environment}
 \label{doorkeyresults}
@@
 
-The Unlock Pickup environment is an even more complex environment than DoorKey, thus giving rewards to the agent even more sparsely.
+The Unlock Pickup environment is an even more complex environment than DoorKey, requiring a much longer string of actions to give out a reward.
 As may be anticipated, this added complexity to the state space makes the LLM contribution even more significant to the RL performance.
 
-Compared to the baseline PPO when averaged per episode over all 8 runs, LLM reward shaping gave a NUMBER\% increase in reward from the environment, with a standard deviation of NUMBER (PPO alone had a standard deviation of NUMBER).
-LLM policy optimization gave a NUMBER\% increase compared to the PPO baseline and a NUMBER\% increase compared to reward shaping.
-Also here giving the highest average reward per episode, the combination of reward shaping and policy optimization gave NUMBER\% more reward compared to the PPO baseline, NUMBER\% more than reward shaping and NUMBER\% higher than policy optimization.
+\begin{figure}[h]
+\centerline{\includegraphics[width=3.25in]{figure/unlockpickupresults.png}}
+\caption{Reward per episode for the three different methods in the UnlockPickup environment}
+\label{unlockpickupresults}
+\end{figure}
+
+Looking at figure \ref{unlockpickupresults}, we can see that all methods struggle.
+However, we see that the LLM aided methods are faring much better than vanilla PPO, which struggles to learn the correct policy in this complex environment.
+Given that terrible performance of the baseline, LLM reward shaping achieved a percentage increase of over $700\%$, which is not that impressive when the average reward from the baseline was close to zero.
+The baseline achieved an average reward of $0.0089$ (with a standard deviation of $0.015$), while LLM reward shaping achieved an average reward of $0.062$ (with a standard deviation of $0.081$).
+
+% LLM policy optimization gave a NUMBER\% increase compared to the PPO baseline and a NUMBER\% increase compared to reward shaping.
 
@@
 
-It works man, dogski did it!
+As shown in the previously referenced work (\ref{background}), the integration of a Large Language Model into the Reinforcement Learning framework can improve model performance.
+In the Minigrid environments we have tested on, we have shown that LLM reward shaping gives increased performance across the board, which shows it to be a good candidate for reward shaping in many environments with sparse rewards.
+Our method of policy integration has not shown similar increases, giving lower rewards on average than the baseline.
 
-As shown in the previously referenced work (\ref{background}), the integration of a Large Language Model into the Reinforcement Learning framework can improve network performance.
-In the Minigrid environments we have tested on, DoorKey and Pickup Unlock, we have demonstrated so.
-This has been done both through the LLM shaping the RL reward function, and through the LLM influencing the action probabilities of the RL agent, as well as through combining these two methodologies.
+For the DoorKey environment, we found that LLM reward shaping gives a very significant $24\%$ increase in average rewards, aiding the agent in all stages of training.
+LLM policy influencing has sadly not given the same increase, underperforming in the more difficult environments.
 
-For the DoorKey environment, we found that all three methods surpassed the performance of our PPO baseline.
+For the UnlockPickup environment, we find that all methods struggle to complete the goal, but that LLM-aided methods also here outperform vanilla PPO, achieving METRIC.
 
@@
 
-% \begin{table*}[h]
-% \caption{PPO hyperparameters}
-% \begin{center}
-% \label{hyperparams}
-% \begin{tabular}{c | l l l l l l l l} 
-% % & & \\ % put some space after the caption
-% % \hline
-% Parameter & Gamma & Steps/episode & Epochs/episode & Batches/epoch & Clip & Value loss coeff. & Entropy coeff. & Learning rate \\
-% \hline
-% Value & 0.99 & 1024 & 10 & 8 & 0.2 & 0.3 & 0 & 3.0e-4 \\
-% \hline
-% \end{tabular}
-% \end{center}
-% \end{table*}
+\begin{figure}[h]
+\centerline{\includegraphics[width=3.25in]{figure/unlockpickupenv.png}}
+\caption{The Minigrid UnlockPickup environment}
+\label{unlockpickupenv}
+\end{figure}
+
 \begin{table*}[h]

### 2023-11-03 doc/paper.tex

@@
 
-% LLM policy optimization gave a NUMBER\% increase compared to the PPO baseline and a NUMBER\% increase compared to reward shaping.
+\section{Discussion, Future Work and Conclusion}
 
-\section{Conclusion}
+As shown in the previously referenced work (\ref{background}), the integration of a Large Language Model into the Reinforcement Learning framework can improve network performance.
+There are various ways to do this, and our work attempts to compare two main methodologies of integrating the LLM, namely through reward shaping and policy influencing.
+We have implemented algorithms to perform this on top of a PPO baseline, and have measured varying results across environments of different complexities.
 
-As shown in the previously referenced work (\ref{background}), the integration of a Large Language Model into the Reinforcement Learning framework can improve model performance.
-In the Minigrid environments we have tested on, we have shown that LLM reward shaping gives increased performance across the board, which shows it to be a good candidate for reward shaping in many environments with sparse rewards.
-Our method of policy integration has not shown similar increases, giving lower rewards on average than the baseline.
+Our reward shaping algorithm did best in the more complex environments (DoorKey and Unlock Pickup), while policy influencing performed better in the simpler Empty environment.
+A possible explanation for this could lie in the details of how the algorithms impact choice of action.
+LLM reward shaping does not directly impact decision making at all; it only impacts the reward function, meaning the mechanism that dictates which action is taken is still solely the actor network present in PPO.
+This means that reward shaping does not introduce any risk of the agent choosing a suboptimal action directly due to being told to do so by the LLM.
+Learning which action to take is still done through the same process of trial and error as in classical RL.
 
-For the DoorKey environment, we found that LLM reward shaping gives a very significant $24\%$ increase in average rewards, aiding the agent in all stages of training.
-LLM policy influencing has sadly not given the same increase, underperforming in the more difficult environments.
+However, the causality behind which action is taken is different in LLM policy influencing; the LLM suggestions are added directly to the logits used to calculate probabilities of each action, meaning that the LLM directly influences the decision making process.
+Every time the LLM suggests performing a suboptimal action, this impacts the action probabilities, regardless whether the PPO baseline has learned that another action is better or not.
+Over time, this may lead to the PPO not being able to explore other actions than the ones proposed by the LLM, thus making the PPO learn slower than if it was allowed to choose actions on its own.
+In other words, the LLM may have too much impact in the decision making when performing policy influencing, making the algorithm prone to errors from the LLM.
 
-For the UnlockPickup environment, we find that all methods struggle to complete the goal, but that LLM-aided methods also here outperform vanilla PPO, achieving METRIC.
+A natural suggestion for how to improve the performance of LLM policy influencing in the future thus becomes to decrease the influence from the LLM over time, letting the PPO choose on its own as learning progresses.
+One way of doing this is through annealing of the LLM values added to the logits.
+This can be combined with experimenting with a hyperparameter determining how the LLM values should be weighted at the beginning of training.
+We have attempted to do this, but have encountered programmatic issues regarding scaling; the logits change in magnitude over time, and a fitting way of dynamically changing the LLM contributions to the logits has to be found.
 

### 2023-11-03 doc/paper.tex
final mods to paper draft

@@
 Large Language Models (LLMs) are trained on large corpora of human written text, and have been shown to capture parts of this intuition in many tasks.
-In recent years, LLMs have been succesfully used to aid RL agents in more efficient exploration, and have allowed them to solve problems which previous methods have been unable to.
+In recent years, LLMs have been successfully used to aid RL agents in more efficient exploration, and have allowed them to solve problems which previous methods have been unable to.
 We compare different methods for integrating an LLM into the Deep Learning RL algorithm Proximal Policy Optimization (PPO), and compare their efficiency against each other.
-We find that BEST METHOD gives the best sample efficiency, outperforming normal PPO by PERCENTAGE in METRIC.
 
@@
 \begin{IEEEkeywords}
-large language models, reinforcement learning, minigrid, proximal policy influencing
+large language models, reinforcement learning, minigrid, proximal policy optimization 
 \end{IEEEkeywords}
@@
 In recent years, LLMs have shown remarkable capabilities in problem solving and planning \cite{sparks}, qualities which traditional RL agents often lack.
-Thus, a new area of research has emerged, which attemps to use the vast amount of human knowledge encoded in these models to improve the performance of reinforcement learning algorithms \cite{survey}.
+Thus, a new area of research has emerged, which attempts to use the vast amount of human knowledge encoded in these models to improve the performance of reinforcement learning algorithms \cite{survey}.
 These LLM assisted RL agents have been able to outperform state-of-the-art RL methods in many problems \cite{omni, ellm, idm}.
@@
 
-For the implementation details of our PPO and the hyperparameters used, take a look at Appendix X.X.
+For the implementation details of our PPO and the hyperparameters used, see the Appendix.
 
@@
 The agent is trained as in the baseline.
-The agent's state observation is encoded to natural text, and sent as part of a prompt to the LLM.
-This may be done at every timestep, or after some interval of timesteps.
+The agent's state observation is encoded to natural text, and sent as part of a prompt to the LLM at every timestep.
 In addition to the state observation, the LLM is also asked to respond with a list of recommended actions for the agent to perform.
@@
 In both these cases, we need two language models; one to generate the suggestions and one to generate semantic similarity values.
+
 For text generation, we have decided utilize an LLM from the Llama open-source family of language models developed by Meta AI, first released in Febraury, 2023.
@@
 Our plan of action, on the other hand, aims at evaluating the efficacy of the LLM without any previous pre-tuning.
-From the perspective of reinforcement learning, is is the appropriate course of action, as the LLM should not possess any previous knowledge about the task at hand, just like the agent should not.
+From the perspective of reinforcement learning, this is the appropriate course of action, as the LLM should not possess any previous knowledge about the task at hand, just like the agent should not.
 In addtion, the act of fine-tuning LLM is a rather time-comsuming process, requiring the generation and structuring of a new dataset, and also vast hardware resources.
@@
 An important characteristic of all these environments is that they only give reward on the completion of the goal of the environment.
-In other words, the environment only ever gives out a single reward per episode, at the end of the episode.
-Goals in minigrid are often simple tasks, such as "unlock the door", "go to the goal", "pick up the red chest".
+In other words, the environment only ever gives out a single reward per episode, given that the agent completes its task.
+Goals in Minigrid are often simple tasks, such as "unlock the door", "go to the goal", "pick up the red chest", etc.
 Completing these goals often requires the completion of subtasks; going to the goal might require opening a door, and opening the door might require picking up a key.
@@
 
-\subsection{Empty Environment}
+\subsubsection{Empty environment}
 
@@
 
-As we can see from Figure \ref{doorkeyresults}, LLM integration gives a noticable improvement to the performance of the model, not just in the early stages of training, but throughout the entire training process.
+As we can see from Figure \ref{doorkeyresults}, LLM reward shaping gives a noticable improvement to the performance of the model, not just in the early stages of training, but throughout the entire training process.
 Averaging over all episodes of all eight runs per model, we find that LLM reward shaping receives $24\%$ more reward per episode than PPO trained without LLM rewards.
@@
 
-***TODO: add paragraph about using both together*** 
+% ***TODO: add paragraph about using both together*** 
 % However, the results from both of these methods are eclipsed by the results from combining LLM reward shaping and policy influencing together.
@@
 This can be combined with experimenting with a hyperparameter determining how the LLM values should be weighted at the beginning of training.
-We have attempted to do this, but have encountered programmatic issues regarding scaling; the logits change in magnitude over time, and a fitting way of dynamically changing the LLM contributions to the logits has to be found.
+We have attempted to do this, but have encountered programmatic issues regarding scaling; the logits change in magnitude over time, and a fitting way of dynamically changing the LLM contributions to the logits has to be found.We believe this is what causes the noticable dip in performance that happens after some initial rollouts in the policy influencing results.
+
+Overall, we found that the LLM can be a valuable asset in the RL framework, but that a balance must be struck regarding how significantly it should influence the learning process.
 
@@
 
-I would like to acknowledge myself for being a absolute legend.
+We would like to acknowledge the ROBIN research group at the University of Oslo Dept.
+of Informatics, who have arranged for us to be able to undertake this research and gain valuable experience.
+In particular, we would like to thank our supervisors Katrine Nergård and Kai Olav Ellefsen for their help and guidance in this project.
 

### 2023-11-03 doc/paper.tex
DRAFT
Signed-off-by: Jonatan Hanssen <jonatan.hanssen@outlook.com>

--- 2023-11-03 doc/paper.tex
@@
 Looking at figure \ref{unlockpickupresults}, we can see that all methods struggle.
-However, we see that the LLM aided methods are faring much better than vanilla PPO, which struggles to learn the correct policy in this complex environment.
+However, we see that the LLM reward shaping performs much better than vanilla PPO, which struggles to learn the correct policy in this complex environment.
 Given that terrible performance of the baseline, LLM reward shaping achieved a percentage increase of over $700\%$, which is not that impressive when the average reward from the baseline was close to zero.
 The baseline achieved an average reward of $0.0089$ (with a standard deviation of $0.015$), while LLM reward shaping achieved an average reward of $0.062$ (with a standard deviation of $0.081$).
+
+{\it TODO: Add results from policy influencing}
 

### 2023-11-06 src/ActorCritic.py
The dogski system
Signed-off-by: Jonatan Hanssen <jonatan.hanssen@outlook.com>

@@ -71,7 +71,7 @@ class Agent(nn.Module):
 
         goal = env.reset()[0]["mission"]
         if not consigliere:
-            self.consigliere = llama2_policy(goal, cos_sim_threshold=0.7, similarity_modifier=1) if llama else None
+            self.consigliere = llama2_policy(goal, cos_sim_threshold=0.7, similarity_modifier=5) if llama else None
         else:
             self.consigliere = consigliere
 
@@ -124,7 +124,7 @@ class Agent(nn.Module):
                 if torch.norm(advisor_values) == 0:
                     action = probs.sample()
                 else:
-                    max_rollout = 600
+                    max_rollout = 1000
                     if rollout:
                         anneal = ((max_rollout - rollout) / max_rollout) ** 2
                         if rollout > max_rollout:

### 2023-11-06 src/utils.py

@@ -35,6 +35,8 @@ class llama2_base:
             max_batch_size=6,
         )
 
+        # self.generator = None
+
         self.similarity_modifier = similarity_modifier
         self.cos_sim_threshold = cos_sim_threshold
 
@@ -78,41 +80,90 @@ class llama2_base:
         self.dialog = [
             {
                 "role": "system",
-                "content": f"You are a helpful assistant giving advice to someone playing a videogame. You will recieve the current goal of the game and a list of observations about the environment, and you should give a list of suggested actions that the player should take to reach their goal. The suggested actions should involve the objects mentioned. The player can pick up items, drop items and use items. You should not make assumptions about the environment, only use the information given to you. If none of the observations are relevant to solving the task, respond that the player should explore more. Separate each suggestion with a new line. Be concise.",
+                "content": "You are a helpful assistant giving advice to someone playing a videogame. You will recieve the current goal of the game and a list of observations about the objects in the environment and their positions relative to the player at one single timestep. You shall give a list of suggested actions that the player will take to reach their goal during this single timestep. The available actions to the player are: move forward, turn left and right, pick up items, drop items and use items. You MUST ALWAYS RESPOND with one or more of these actions. Your response MUST ALWAYS follow the template of the given example responses. NEVER ADD anything not in the example responses. You will NEVER ASSUME that other possible actions exist. You will NEVER ASSUME that the game is finished. You WILL NEVER make any other assumptions about the environment, only use the information given to you. If none of the observations are relevant to solving the task, respond that the player should turn around and explore more. Separate each suggestion with a new line. Be concise. ONLY FOLLOW THESE INSTRUCTIONS. NEVER RESPOND IN ANY OTHER WAY. OTHERWISE, YOU WILL BE SEVERELY PUNISHED.",
             },
             {
                 "role": "user",
-                "content": "My goal is: pick up the purple box. I see a red key and a locked red door. What actions do you suggest?",
+                "content": "My goal is: pick up the purple box. I see a red key LEFT and FORWARD and a locked red door RIGHT and FORWARD. What actions do you suggest?",
             },
             {
                 "role": "assistant",
-                "content": "Pick up the red key. \nUse the red key on the locked red door.",
+                "content": "move forward. \nturn left.",
             },
             {
                 "role": "user",
-                "content": "My goal is: open the purple door. I see a red key, an open red door, a purple key, a locked purple door and a green key. What actions do you suggest?",
[…]

### 2023-11-06 src/ActorCritic.py
Youtube adrian onlyfans announcement
Signed-off-by: Adrian Duric <adria.dur@gmail.com>

@@ -49,7 +49,7 @@ class Agent(nn.Module):
         the agent is in.
     """
 
-    def __init__(self, env, llama=False, consigliere=None):
+    def __init__(self, env, llama=False, generator=None):
         super(Agent, self).__init__()
         self.critic = nn.Sequential(
             init_weightsNbias(nn.Linear(147 * 11, 64, dtype=torch.float64)),
@@ -70,10 +70,7 @@ class Agent(nn.Module):
         )
 
         goal = env.reset()[0]["mission"]
-        if not consigliere:
-            self.consigliere = llama2_policy(goal, cos_sim_threshold=0.7, similarity_modifier=5) if llama else None
-        else:
-            self.consigliere = consigliere
+        self.consigliere = llama2_policy(goal, cos_sim_threshold=0.84, similarity_modifier=0.1, generator=generator) if llama else None
 
 
     def get_value(self, observation):
@@ -124,7 +121,7 @@ class Agent(nn.Module):
                 if torch.norm(advisor_values) == 0:
                     action = probs.sample()
                 else:
-                    max_rollout = 1000
+                    max_rollout = 700
                     if rollout:
                         anneal = ((max_rollout - rollout) / max_rollout) ** 2
                         if rollout > max_rollout:

### 2023-11-06 src/ppo.py

@@ -18,7 +18,7 @@ base_path = os.path.dirname(__file__)
 
 class PPO:
     def __init__(
-        self, param_file, result_file=None, reward=None, policy=None, consigliere=None
+        self, param_file, result_file=None, reward=None, policy=None, generator=None, allow_farming=False
     ):
         self.args = read_params(param_file)
 
@@ -35,7 +35,7 @@ class PPO:
 
         self.device = torch.device("cuda" if torch.cuda.is_available() else "cpu")
         # self.device = torch.device("cpu")
-        self.agent = Agent(self.env, self.llama_policy, consigliere=consigliere).to(
+        self.agent = Agent(self.env, self.llama_policy, generator=generator).to(
             self.device
         )
         self.optimizer = optim.Adam(
@@ -48,6 +48,8 @@ class PPO:
                 self.env.reset()[0]["mission"],
                 similarity_modifier=0.005,
                 cos_sim_threshold=0.84,
+                generator=generator,
+                allow_farming=allow_farming,
             )
             if self.llama_reward
             else None
@@ -291,13 +293,30 @@ if __name__ == "__main__":
     )
     parser.add_argument("-r", "--reward", type=bool, help="Use LLM reward shaping", default=None)
     parser.add_argument("-p", "--policy", type=bool, help="Use LLM policy", default=None)
+    parser.add_argument("-a", "--allow_farming", action="store_true")
+    parser.add_argument("-d", "--doobie", action="store_true")
+
     args = parser.parse_args()
     print(args)
+
+    if args.doobie:
+        generator = 1
+    else:
+        generator = Llama.build(
+            ckpt_dir=os.path.join(base_path, "../llama-2-7b-chat"),
+            tokenizer_path=os.path.join(
+                base_path, "../llama-2-7b-chat/tokenizer.model"
+            ),
+            max_seq_len=2048,
+            max_batch_size=6,
+        )
 
     ppo = PPO(
         "hyperparams.json",
         result_file=args.result_file,
         reward=args.reward,
         policy=args.policy,
+        generator=generator,
+        allow_farming=args.allow_farming,
     )
     ppo.train()

### 2023-11-06 src/utils.py

@@ -25,15 +25,20 @@ class llama2_base:
         rl_temp=0,
         cache_file=None,
         sim_cache_file=None,
+        generator=None,
+        allow_farming=False,
     ):
-        self.generator = Llama.build(
-            ckpt_dir=os.path.join(base_path, "../llama-2-7b-chat"),
-            tokenizer_path=os.path.join(
-                base_path, "../llama-2-7b-chat/tokenizer.model"
-            ),
-            max_seq_len=2048,
-            max_batch_size=6,
-        )
+        if generator:
+            self.generator = generator
+        else:
+            self.generator = Llama.build(
+                ckpt_dir=os.path.join(base_path, "../llama-2-7b-chat"),
+                tokenizer_path=os.path.join(
+                    base_path, "../llama-2-7b-chat/tokenizer.model"
+                ),
+                max_seq_len=2048,
+                max_batch_size=6,
+            )
 
         # self.generator = None
 
@@ -47,6 +52,7 @@ class llama2_base:
         self.rl_temp = rl_temp
         self.suggestions = None
         self.goal = goal
+        self.allow_farming = allow_farming
 
         self.cache_file = cache_file
         self.cache_misses = 0
@@ -188,12 +194,16 @@ class llama2_base:
         else:
             # print("Cache miss")
             self.cache_misses += 1
-            result = self.generator.chat_completion(
-                [self.dialog],  # type: ignore
-                max_gen_len=100,
-                temperature=self.temperature,
-                top_p=self.top_p,
-            )
+            try:
+                result = self.generator.chat_completion(
+                    [self.dialog],  # type: ignore
+                    max_gen_len=100,
+                    temperature=self.temperature,
+                    top_p=self.top_p,
+                )
+            except AttributeError:
+                print("im smoking doobies")
+                result = [{"generation": {"content": "i smoking that zaza"}}]
 
             answer = result[0]["generation"]["content"]
 
@@ -232,8 +242,8 @@ class llama2_base:
         # print(f"{caption=}")
         # print(f"{self.suggestions=}")
 
-        # if not caption or caption in self.caption_set:
-        #     return 0.0
+        if caption in self.caption_set and self.allow_farming == False:
+            return 0.0
 
         self.caption_set.add(caption)
 
@@ -349,6 +359,8 @@ class llama2_reward_shaper(llama2_base):
         rl_temp=0,
         cache_file="cache_reward.json",
[…]

### 2023-11-07 doc/paper.tex
update paper
Signed-off-by: Jonatan Hanssen <jonatan.hanssen@outlook.com>

@@
 \usepackage{xcolor}
+\usepackage{float}
 \usepackage{hyperref}
@@
 These values are joined with the logits of the actions, thus influencing the probability weights of each action being taken so that semantically similar actions are weighted higher than others.
-We call this the Actor-Advisor-Critic Model.
+We call this the Actor-Advisor-Critic model.
 
@@
 Compared to the policy influencing method above, this approach is less direct, as the policy is only indirectly affected by the LLM through these additional rewards.
+
+\subsubsection{PPO with both LLM reward shaping and LLM policy influencing}
+
+Finally, we combine the two LLM integrations, hoping to combine the direct influence of our policy integration with the more indirect reward shaping.
+Thus, we are affecting the policy in two ways: Through reward shaping, we are causing the policy to update to better attain these pseudorewards, which will encourage completing of subgoals which are necessary to complete the actual goal of the environment, but which do not give out intermediate rewards.
+Through policy influencing, we are nudging the agent to take favorable actions by directly altering the action probabilities, which will make it more likely that the agent completes the given task and receives an actual reward from the environment.
+Thus, the agent will more quickly discover the correct policy.
 
@@
 We select three different Minigrid environments with increasing complexity; Empty, DoorKey and UnlockPickup.
-For each environment we test our three different models; PPO with no LLM integration, PPO with LLM reward shaping and PPO with LLM policy influencing.
+For each environment we test our four different models; PPO with no LLM integration, PPO with LLM reward shaping and PPO with LLM policy influencing.
 All methods use the same hyperparameters for the PPO, which can be found in the appendix.
@@
 \centerline{\includegraphics[width=3.25in]{figure/emptyresults.png}}
-\caption{Reward per episode for the three different methods in the Empty environment}
+\caption{Reward per episode for the four different methods in the Empty environment}
 \label{doorkeyresults}
 \end{figure}
+
+\begin{table}[h]
+\caption{All results on Empty environment}
+\begin{center}
+\label{emptytable}
+\begin{tabular}{c | c c c}
+Method & Mean reward & Std reward & Increase over baseline \\
+\hline
+Baseline & 0.99 & 1 & 0\% \\
+Reward shaping & 0.99 & 1 & 30\% \\
+Policy influencing & 0.99 & 1 & -7\% \\
+Both & 0.99 & 1 & 36\% \\
+\end{tabular}
+\end{center}
+\end{table}
 
@@
 
+\begin{table}[h]
+\caption{All results on DoorKey environment}
+\begin{center}
+\label{doorkeytable}
+\begin{tabular}{c | c c c}
+Method & Mean reward & Std reward & Increase over baseline \\
+\hline
+Baseline & 0.99 & 1 & 0\% \\
+Reward shaping & 0.99 & 1 & 30\% \\
+Policy influencing & 0.99 & 1 & -7\% \\
+Both & 0.99 & 1 & 36\% \\
+\end{tabular}
+\end{center}
+\end{table}
+
 % ***TODO: add paragraph about using both together*** 
@@
 \centerline{\includegraphics[width=3.25in]{figure/doorkeyresults.png}}
-\caption{Reward per episode for the three different methods in the DoorKey environment}
+\caption{Reward per episode for the four different methods in the DoorKey environment}
 \label{doorkeyresults}
@@
 \centerline{\includegraphics[width=3.25in]{figure/unlockpickupresults.png}}
-\caption{Reward per episode for the three different methods in the UnlockPickup environment}
+\caption{Reward per episode for the four different methods in the UnlockPickup environment}
 \label{unlockpickupresults}
@@
 The baseline achieved an average reward of $0.0089$ (with a standard deviation of $0.015$), while LLM reward shaping achieved an average reward of $0.062$ (with a standard deviation of $0.081$).
+
+\begin{table}[h]
+\caption{All results on UnlockPickup environment}
+\begin{center}
+\label{unlockpickuptable}
+\begin{tabular}{c | c c c}
+Method & Mean reward & Std reward & Increase over baseline \\
+\hline
+Baseline & 0.99 & 1 & 0\% \\
+Reward shaping & 0.99 & 1 & 30\% \\
+Policy influencing & 0.99 & 1 & -7\% \\
+Both & 0.99 & 1 & 36\% \\
+\end{tabular}
+\end{center}
+\end{table}
 
@@
 
-\begin{table*}[h]
+\begin{table}[h]
 \caption{PPO hyperparameters}
@@
 \end{center}
-\end{table*}
+\end{table}
 

### 2023-11-07 doc/paper.tex
added to paper

@@
 
-In this simple environment, we see that both LLM reward shaping and LLM policy influencing outperform the baseline model.
-Averaging over all episodes for all eight runs per model, we find LLM policy influencing to perform best, receiving $30\%$ more reward than the baseline.
-LLM reward shaping similarly outperforms the baseline, receiving $10\%$ more reward.
-Given that policy influencing is a more direct approach, it makes sense that this method performs best, as the simple environment makes it likely that the LLM will suggest the best action the majority of the time.
+In this simple environment, we see that both LLM reward shaping, LLM policy influencing and A2C2 outperform the baseline model.
+Averaging over all episodes for all eight runs, we find that reward shaping increases the mean reward received by $10\%$ over the baseline, with a lower standard deviation in rewards received per run of $0.12$.
+Policy influencing scores even higher at $32\%$ over the baseline, with an even lower standard deviation of $0.09$.
+Finally, A2C2 performs the best of the four with a $37\%$ increase in average reward over the baseline model, and the lowest standard deviation of $0.04$
+
+Notably, both policy influencing and A2C2 receive much higher rewards than reward shaping and the baseline right from the first episode.
+This makes sense as the environment is so simple that the LLM seems not to have any problem giving sensible instructions to directly influence the policy most of the time.
+However, we notice from the graph that policy influencing alone consistently gives a significant dip in received reward after the first few episodes, which it later compensates for, following a similar trajectory to that of reward shaping after about 50 episodes.
+We believe this could be a sign of the policy influencing being somewhat overly influential in changing the logits early on, because the logits initially have low values.
+As the PPO learns the environment, the magnitude of the logits increases, which at some point leads to conflicting contributions to the logits from the LLM and from the PPO.
+As the dip stops and performance increases again, this is likely due to the PPO becoming more influencing than the LLM after having started to learn the environment properly.
+We also notice that this dip is completely removed with A2C2, which could signify that the added reward shaping helps the PPO learn faster, enabling it to increase the influence from PPO on the logits faster.
 
@@
 \hline
-Baseline & 0.99 & 1 & 0\% \\
-Reward shaping & 0.99 & 1 & 30\% \\
-Policy influencing & 0.99 & 1 & -7\% \\
-Both & 0.99 & 1 & 36\% \\
+Reward shaping & 0.76 & 0.12 & 10\% \\
+Policy influencing & 0.91 & 0.09 & 32\% \\
+Both & 0.95 & 0.04 & 37\% \\
+Baseline & 0.69 & 0.28 & 0\% \\
 \end{tabular}

### 2023-11-07 doc/paper.tex
paper dog
Signed-off-by: Jonatan Hanssen <jonatan.hanssen@outlook.com>

--- 2023-11-07 doc/paper.tex
@@
 \usepackage{xcolor}
+\usepackage{tabularx}
 \usepackage{float}
@@
 These values are joined with the logits of the actions, thus influencing the probability weights of each action being taken so that semantically similar actions are weighted higher than others.
-We call this the Actor-Advisor-Critic model.
+This is a very direct approach to influencing the agent; if an action is favoured by the LLM in a timestep, this will affect the actions immediately by increasing the chance that this action is sampled.
+Thus, this method functions as a sort of coach, giving immediate and direct suggestions throughout the episodes.
+Throughout the paper, we will refer to this as LLM policy influencing.
 
@@
 Then, the LLM is used as a tool for semantic similarity comparison, comparing its own formerly proposed action to the action performed by the agent.
-Finally, high semantic similarity gives an additional reward which is added to the reward of the environment.
-Compared to the policy influencing method above, this approach is less direct, as the policy is only indirectly affected by the LLM through these additional rewards.
+Finally, high semantic similarity gives an additional pseudo-reward which is added to the reward of the environment.
+Compared to the policy influencing method above, this approach is less direct, as the policy is only indirectly affected by the LLM through these additional rewards, which only take effect when the policy is updated through backpropogation after an entire episode is completed.
+Thus, this method functions more as an advisor, influencing the policy in a more long term capacity than the coaching of the policy influencer.
+Throughout the paper, we will refer to this as LLM reward shaping.
 
@@
 Finally, we combine the two LLM integrations, hoping to combine the direct influence of our policy integration with the more indirect reward shaping.
-Thus, we are affecting the policy in two ways: Through reward shaping, we are causing the policy to update to better attain these pseudorewards, which will encourage completing of subgoals which are necessary to complete the actual goal of the environment, but which do not give out intermediate rewards.
+Thus, we are affecting the policy in two ways: Through reward shaping, we are causing the policy to update to better attain the aforementioned pseudo-rewards, which will encourage completing of subgoals which are necessary to complete the actual goal of the environment, but which do not give out intermediate rewards.
 Through policy influencing, we are nudging the agent to take favorable actions by directly altering the action probabilities, which will make it more likely that the agent completes the given task and receives an actual reward from the environment.
 Thus, the agent will more quickly discover the correct policy.
+We call this the \textbf{Actor-Advisor-Critic-Coach} method, or \textbf{A2C2} for short.
 
@@
 
-As we can see from Figure \ref{doorkeyresults}, LLM reward shaping gives a noticable improvement to the performance of the model, not just in the early stages of training, but throughout the entire training process.
-Averaging over all episodes of all eight runs per model, we find that LLM reward shaping receives $24\%$ more reward per episode than PPO trained without LLM rewards.
-Furthermore, this is an improvement that comes at very little computational cost, because LLM responses can be cached between runs.
-Of the eight runs on this environment using LLM reward shaping, the last three runs never queried the LLM at all, only relying on the cache of previous answers.
-In all methods, we see that the standard deviation is very large ($0.28$ and $0.31$ for LLM reward shaping and baseline, respectively).
-Given that there is no reward given for a partially completed task, this is not unexpected.
-
-% We also notice a significant performance increase when using LLM policy influencing, even somewhat more than from reward shaping.
-Averaging over all episodes of all runs, we observe a NUMBER\% increase in reward per episode compared to the PPO baseline, and a NUMBER\% increase when compared to LLM reward shaping.
-The standard deviation when using LLM policy influencing is also similarly large to PPO and reward shaping (NUMBER).
-By caching LLM responses similarly to when performing reward shaping, this method too adds very little computational cost to the training in total.
-
-Using LLM policy influencing in the DoorKey environment has resulted in somewhat worse results than using the PPO baseline.
-Averaging per episode over all eight runs shows that LLM policy influencing only achieves $89\%$ of the reward amount per episode that the baseline achieves.
-Judging from the graph, it may seem that policy influencing may have given some initial learning advantage, but ultimately slows learning for the agent further on.
-The method also gives the highest standard deviation of reward given per episode of the three compared so far, at $0.25$.
-As with reward shaping, we have also used caching of LLM responses to significantly decrease the computational overhead cost to a minimum.
+As we can see from Figure \ref{doorkeyresults}, we see impressive results from our LLM integration in this environment as well, especially from reward shaping and A2C2.
+For the mean and standard deviation of all methods, see table \ref{doorkeytable}.
 
@@
 
-% ***TODO: add paragraph about using both together*** 
-% However, the results from both of these methods are eclipsed by the results from combining LLM reward shaping and policy influencing together.
-Averaging over 8 runs, we observe a NUMBER\% increase in reward from the environment compared to PPO alone, NUMBER\% higher than reward shaping and NUMBER\%higher than policy influencing, with a similar standard deviation of NUMBER.
+LLM reward shaping gives a noticable improvement to the performance of the model, not just in the early stages of training, but throughout the entire training process.
+Averaging over all episodes of all eight runs per model, we find that LLM reward shaping receives $24\%$ more reward per episode than PPO trained without LLM rewards.
+Furthermore, this is an improvement that comes at very little computational cost, because LLM responses can be cached between runs.
+Of the eight runs on this environment using LLM reward shaping, the last three runs never queried the LLM at all, only relying on the cache of previous answers.
+
+Looking at LLM policy influencing, we see that it has resulted in somewhat worse results than using the PPO baseline.
+Averaging per episode over all eight runs shows that LLM policy influencing only achieves $89\%$ of the reward amount per episode that the baseline achieves.
+Judging from the graph, it may seem that policy influencing may have given some initial learning advantage, but ultimately slows learning for the agent further on.
+The method also gives the highest standard deviation of reward given per episode of the three compared so far, at $0.25$.
+As with reward shaping, we have also used caching of LLM responses to significantly decrease the computational overhead cost to a minimum.
+
+However, the results from both of these methods are eclipsed by the results from combining LLM reward shaping and policy influencing together in our A2C2 model.
+Averaging over 8 runs, we observe a $34\%$ increase in reward from the environment compared to PPO alone, $8\%$ higher than reward shaping and $43\%$ higher than policy influencing, with the lowest standard deviation of all methods, at $0.22$.
+Clearly, the combination of immediate direct suggestions from the policy influencer and long term indirect suggestions from the reward shaper is a favourable way of integrating the LLM with PPO, which gives results which are better than the sum of its parts, not only in the simple Empty environment, but also in more complicated environments like DoorKey.
 
@@
 The Unlock Pickup environment is an even more complex environment than DoorKey, requiring a much longer string of actions to give out a reward.
-As may be anticipated, this added complexity to the state space makes the LLM contribution even more significant to the RL performance.
+As may be anticipated, this added complexity makes the average reward much lower, and we see that all methods struggle (figure \ref{unlockpickupresults}).
 
@@
 
-Looking at figure \ref{unlockpickupresults}, we can see that all methods struggle.
-However, we see that the LLM reward shaping performs much better than vanilla PPO, which struggles to learn the correct policy in this complex environment.
-Given that terrible performance of the baseline, LLM reward shaping achieved a percentage increase of over $700\%$, which is not that impressive when the average reward from the baseline was close to zero.
+As we can see, LLM performs the best overall, and performs much better than vanilla PPO, which struggles to learn the correct policy in this complex environment.
+Given the terrible performance of the baseline, LLM reward shaping achieved a percentage increase of over $700\%$, which is not that impressive when the average reward from the baseline was close to zero.
 The baseline achieved an average reward of $0.0089$ (with a standard deviation of $0.015$), while LLM reward shaping achieved an average reward of $0.062$ (with a standard deviation of $0.081$).
+See table \ref{unlockpickuptable}.
 
@@
 
-{\it TODO: Add results from policy influencing}
+As we can see, policy influencing struggles in this environment as well, performing even worse than the baseline.
+Policy influencing received a mean average score of NUMBER, which is PERCENT lower than the baseline.
+In this complicated environment, it may that suboptimal responses from the LLM have high chance to ruin the agent's chance to complete the goal, for example by suggesting to open the box instead of picking it up.
+As explained earlier, this instantly loses the game by removing the box from the game.
+
+Our A2C2 performs much better than just policy influencing, but we are not here able to combine the two LLM integrations to receive a performance higher than the sum of its parts.
+Here we receive a score of NUMBER, which is higher than the baseline and policy influencing, but not higher than reward shaping by itself.
+It may be that in this case, the performance of policy influencing is so poor that the combination is not able to utilize the best of both methods, instead only being dragged down by poor suggestions.
+
+\subsection{Inspecting the LLM responses}
+
+Having looked at the performance of our different models, we can see that the suggestions from the LLM have positive effects in many cases, while also sometimes being detrimental.
+We now inspect some sample responses from the LLM, to gain deeper understanding of how the LLM is effecting the agent.
+
+\subsubsection{Empty}
+
+From table \ref{emptyresponses} we see that the suggestions are very good, which explain the great performance all LLM integrations show.
+
+\begin{table}[h]
+\caption{All results on UnlockPickup environment}
+\begin{center}
+\label{unlockpickuptable}
+\begin{tabular}{c | c}
+Prompt & Response \\
+\hline
+My goal is: get to the green goal square. I see a green goal LEFT and FORWARD." & "move forward."
+\end{tabular}
+\end{center}
+\end{table}
 
@@
 There are various ways to do this, and our work attempts to compare two main methodologies of integrating the LLM, namely through reward shaping and policy influencing.
-We have implemented algorithms to perform this on top of a PPO baseline, and have measured varying results across environments of different complexities.
+We have implemented algorithms to perform this on top of a PPO baseline, as well as a combination of the two, and have measured varying results across environments of different complexities.
 
@@
 However, the causality behind which action is taken is different in LLM policy influencing; the LLM suggestions are added directly to the logits used to calculate probabilities of each action, meaning that the LLM directly influences the decision making process.
-Every time the LLM suggests performing a suboptimal action, this impacts the action probabilities, regardless whether the PPO baseline has learned that another action is better or not.
+Every time the LLM suggests performing a suboptimal action, this impacts the sampled action, regardless whether the PPO baseline has learned that another action is better or not.
 Over time, this may lead to the PPO not being able to explore other actions than the ones proposed by the LLM, thus making the PPO learn slower than if it was allowed to choose actions on its own.

### 2023-11-07 src/ppo.py
Add goal update, + black formatting
Signed-off-by: Jonatan Hanssen <jonatan.hanssen@outlook.com>

@@ -18,7 +18,13 @@ base_path = os.path.dirname(__file__)
 
 class PPO:
     def __init__(
-        self, param_file, result_file=None, reward=None, policy=None, generator=None, allow_farming=False
+        self,
+        param_file,
+        result_file=None,
+        reward=None,
+        policy=None,
+        generator=None,
+        allow_farming=False,
     ):
         self.args = read_params(param_file)
 
@@ -81,7 +87,9 @@ class PPO:
             (
                 env_reward,
                 advisor_reward_cum,
-            ) = self.next_episode(rollout)  # Perform steps and store relevant values
+            ) = self.next_episode(
+                rollout
+            )  # Perform steps and store relevant values
             rewards.append(env_reward)
             # if not env_reward:
             #     print("No reward received, skipping update of networks")
@@ -121,12 +129,19 @@ class PPO:
         observation = torch.Tensor(observation_dict["image"].flatten()).to(self.device)
         env_reward = 0
         advisor_reward_cum = 0
+        if self.reward_shaper:
+            self.reward_shaper.goal = observation_dict["mission"]
+
+        if self.agent.consigliere:
+            self.agent.consigliere.goal = observation_dict["mission"]
 
         for step in range(self.args["steps"]):
             self.observations[step] = observation
 
             with torch.no_grad():
-                action, logprob, _, value = self.agent.get_action_and_value(observation, rollout=rollout)
+                action, logprob, _, value = self.agent.get_action_and_value(
+                    observation, rollout=rollout
+                )
 
             self.values[step] = value.flatten()
             self.actions[step] = action
@@ -289,10 +304,18 @@ class PPO:
 if __name__ == "__main__":
     parser = argparse.ArgumentParser()
     parser.add_argument(
-        "-f", "--result_file", type=str, help="Output file for env rewards", default=None
+        "-f",
+        "--result_file",
+        type=str,
+        help="Output file for env rewards",
+        default=None,
     )
-    parser.add_argument("-r", "--reward", type=bool, help="Use LLM reward shaping", default=None)
-    parser.add_argument("-p", "--policy", type=bool, help="Use LLM policy", default=None)
+    parser.add_argument(
+        "-r", "--reward", type=bool, help="Use LLM reward shaping", default=None
+    )
+    parser.add_argument(
+        "-p", "--policy", type=bool, help="Use LLM policy", default=None
+    )
     parser.add_argument("-a", "--allow_farming", action="store_true")
     parser.add_argument("-d", "--doobie", action="store_true")
 

### 2023-11-07 doc/paper.tex
Update paperino
Signed-off-by: Jonatan Hanssen <jonatan.hanssen@outlook.com>

@@
 From table \ref{emptyresponses} we see that the suggestions are very good, which explain the great performance all LLM integrations show.
+However, we also see that this is a very simple environment, where one would expect an LLM to be able to respond coherently.
 
 \begin{table}[h]
-\caption{All results on UnlockPickup environment}
+\caption{Sample responses from LLM in Empty environment.}
 \begin{center}
-\label{unlockpickuptable}
-\begin{tabular}{c | c}
+\label{emptyresponses}
+\begin{tabularx}{3.25in}{X | X}
 Prompt & Response \\
 \hline
-My goal is: get to the green goal square. I see a green goal LEFT and FORWARD." & "move forward."
-\end{tabular}
+My goal is: get to the green goal square. I see a green goal left and forward & move forward. \\
+My goal is: get to the green goal square. I see a green goal left. & move left. \\
+My goal is: get to the green goal square. I see a green goal directly in front. & move forward.
+\end{tabularx}
+\end{center}
+\end{table}
+
+\subsubsection{DoorKey}
+
+From table \ref{doorkeyresponses} we can see that some suggestions are good, such as suggestion to pick up a key when it is directly in front, and moving towards the goal although other observations are also present.
+We also see some suboptimal responses, such as suggesting to pick up a key when it is not directly in front and suggesting to pick up a key when it is already in the players inventory.
+These poor responses can explain why policy influencing can have a detrimental impact in this environment, as these actions will always be sampled slightly more often.
+
+\begin{table}[h]
+\caption{Sample responses from LLM in DoorKey environment. Good responses above line and suboptimal responses below. GOAL is equal to {\it use the key to open the door and then get to the goal}}
+\begin{center}
+\label{doorkeyresponses}
+\begin{tabularx}{3.25in}{X | X}
+Prompt & Response \\
+\hline
+My goal is: GOAL. I see a closed yellow door left, see a yellow key directly in front. & pick up yellow key. \\
+My goal is: GOAL. I see a green goal left and forward, see a open yellow door left, see a yellow key right and forward. & move forward. \\
+My goal is: GOAL. I see a green goal left and forward, see a open yellow door left, see a yellow key right. & move forward. \\
+\hline
+My goal is: GOAL. I see a yellow key left and forward, see a locked yellow door left and forward. & pick up yellow key. \\
+My goal is: GOAL. I see a green goal left and forward, have a yellow key in my inventory, see a closed yellow door right and forward. & pick up yellow key.
+\end{tabularx}
+\end{center}
+\end{table}
+
+\subsubsection{UnlockPickup}
+
+On our final environment, we can see from table \ref{unlockpickupresponses}, such as suggestion to pick up a key when it is directly in front, and moving towards the goal although other observations are also present.
+We also see some suboptimal responses, such as suggesting to pick up a key when it is not directly in front and suggesting to pick up a key when it is already in the players inventory.
+These poor responses can explain why policy influencing can have a detrimental impact in this environment, as these actions will always be sampled slightly more often.
+
+\begin{table}[h]
+\caption{Sample responses from LLM in UnlockPickup environment. Good responses above line and suboptimal responses below.}
+\begin{center}
+\label{unlockpickupresponses}
+\begin{tabularx}{3.25in}{X | X}
+Prompt & Response \\
+\hline
+My goal is: pick up the blue box. I see a locked red door LEFT and FORWARD, see a red key directly in front. & pick up red key.
+\hline
+My goal is: pick up the grey box.
+I have a grey key in my inventory.
+& Congratulations, you have successfully reached your goal!
+You find a small, mysterious box inside the grey box.
+It's made of a strange, otherworldly material that seems to glow with an ethereal light.
+As you examine it closer, you can feel a faint hum of energy emanating from it.
+\\
+My goal is: pick up the green box.
+I have a purple key in my inventory.
+& Great!
+With the purple key in your inventory, you can now unlock the green box.
+You approach the green box and use the purple key to unlock it.
+Inside, you find a shiny coin!
+\end{tabularx}
 \end{center}

### 2023-11-07 doc/paper.tex
paper dog ass

--- 2023-11-07 doc/paper.tex
@@
 This is a very direct approach to influencing the agent; if an action is favoured by the LLM in a timestep, this will affect the actions immediately by increasing the chance that this action is sampled.
-Thus, this method functions as a sort of coach, giving immediate and direct suggestions throughout the episodes.
+Thus, this method functions analogously to how a coach instructs a player, giving direct suggestions to the agent immediately as the agent is about to perform actions at all timesteps of each episode.
 Throughout the paper, we will refer to this as LLM policy influencing.
@@
 Specifically, we chose Llama 2, which has been trained using 40\% more data than its predecessor Llama 1, and has double the context length\cite{llama}.
-In the experiments presented below, we use the smallest version of Llama2, with 7 billion parameters, which we download to run locally and reduce the bottleneck effect of having to call an API to a model hosted by someone else.
+In the experiments presented below, we use the smallest version of Llama 2, with 7 billion parameters, which we download to run locally and reduce the bottleneck effect of having to call an API to a model hosted by someone else.
 The reason for not using any of the larger Llama 2 models is simply the hardware limitation we face, with the 7B model requiring 19GB of VRAM to be run.
@@
 
-\subsection{Empty Environment}
+This section outlines the results of our experiments in each of our selected environments. We also show example outputs from the LLM and examine how these may influence our results.
+
+\subsection{Results per environment}
+
+\subsubsection{Empty environment}
+\label{emptyresultstext}
 
@@
 
-\subsection{DoorKey environment}
+\subsubsection{DoorKey environment}
 
@@
 
-\subsection{Unlock Pickup environment}
+\subsubsection{Unlock Pickup environment}
 
@@
 
-As we can see, policy influencing struggles in this environment as well, performing even worse than the baseline.
+As is clear from the graph, policy influencing struggles in this environment as well, performing even worse than the baseline.
 Policy influencing received a mean average score of NUMBER, which is PERCENT lower than the baseline.
@@
 
-Our A2C2 performs much better than just policy influencing, but we are not here able to combine the two LLM integrations to receive a performance higher than the sum of its parts.
+Our A2C2 performs much better than only policy influencing, but in this case we are not able to combine the two LLM integrations to receive a performance higher than the sum of its parts.
 Here we receive a score of NUMBER, which is higher than the baseline and policy influencing, but not higher than reward shaping by itself.
@@
 \subsection{Inspecting the LLM responses}
+\label{llmresponse}
 
@@
 As shown in the previously referenced work (\ref{background}), the integration of a Large Language Model into the Reinforcement Learning framework can improve network performance.
-There are various ways to do this, and our work attempts to compare two main methodologies of integrating the LLM, namely through reward shaping and policy influencing.
-We have implemented algorithms to perform this on top of a PPO baseline, as well as a combination of the two, and have measured varying results across environments of different complexities.
+There are various ways to do this, and our work attempts to compare two main methodologies of integrating the LLM as well as the combination of the two, namely through reward shaping and policy influencing.
+We have implemented algorithms to perform this on top of a PPO baseline, and have measured several interesting results across environments of varying complexities.
 
-Our reward shaping algorithm did best in the more complex environments (DoorKey and Unlock Pickup), while policy influencing performed better in the simpler Empty environment.
+Amongst the two main methodologies, our reward shaping algorithm did best in the more complex environments (DoorKey and Unlock Pickup), while policy influencing performed better in the simpler Empty environment.
 A possible explanation for this could lie in the details of how the algorithms impact choice of action.
@@
 
-A natural suggestion for how to improve the performance of LLM policy influencing in the future thus becomes to decrease the influence from the LLM over time, letting the PPO choose on its own as learning progresses.
+Regardless of these shortcomings in the performance of policy influencing, it seems to have contributed positively when combining the two methods in A2C2, which outperformed each of the methods in both the Empty and DoorKey environments.
+As mentioned in \ref{emptyresultstext}, it seems that because the reward shaping makes the PPO learn the environment faster, it lessens the conflict between the influences of the LLM and the PPO itself on the logits.
+This allows the PPO to exert the most influence on the logits earlier on.
+However, this conflict is not entirely negated, as witnessed in the UnlockPickup environment where reward shaping still vastly outperforms both policy influencing and A2C2.
+
+A natural suggestion for how to improve the performance of both LLM policy influencing and A2C2 thus becomes to find a way to appropriately decrease the influence from the LLM over time, letting the PPO choose on its own as learning progresses.
 One way of doing this is through annealing of the LLM values added to the logits.
 This can be combined with experimenting with a hyperparameter determining how the LLM values should be weighted at the beginning of training.
-We have attempted to do this, but have encountered programmatic issues regarding scaling; the logits change in magnitude over time, and a fitting way of dynamically changing the LLM contributions to the logits has to be found.We believe this is what causes the noticable dip in performance that happens after some initial rollouts in the policy influencing results.
+We have attempted to do this, but have encountered programmatic issues regarding scaling; the logits change in magnitude over time, making it harder to adjust the rate change in LLM influence at a desired pace.
+Therefore, a fitting way of dynamically changing the LLM contributions to the logits at some specified rate has to be found.
+
+Another potential improvement, perhaps the most obvious one, is to utilize a better LLM for generating responses, both for reward shaping and policy influencing.
+The one we used, the 7B model of Llama 2, varies greatly in the quality of responses it produces, as demonstrated in \ref{llmresponse}.
+While is cost-effective both financially and in terms of computational resources, we expect that better results can be achieved with higher-performance models.
+This is especially true for policy influencing, where performance depends heavily on the LLM understanding the environment in detail.
 
 Overall, we found that the LLM can be a valuable asset in the RL framework, but that a balance must be struck regarding how significantly it should influence the learning process.
+Our results suggest that both policy influencing and reward shaping can be beneficial in environments of varying complexity, but the two, or their combination, can contribute differently depending on the environment.
+In simpler environments like Empty, where the LLM shows good understanding of the environment and can make detailed suggestions, policy influencing is capable of contributing the most.
+However, as environment complexity increases and the LLM may not be as precise in policy influencing, it seems to pay off prioritizing reward shaping and having policy influencing contribute less.
 

### 2023-11-07 doc/paper.tex

@@
 \hline
-My goal is: pick up the blue box. I see a locked red door LEFT and FORWARD, see a red key directly in front. & pick up red key.
+My goal is: pick up the blue box. I see a locked red door LEFT and FORWARD, see a red key directly in front. & pick up red key. \\
 \hline
@@
 Our results suggest that both policy influencing and reward shaping can be beneficial in environments of varying complexity, but the two, or their combination, can contribute differently depending on the environment.
-In simpler environments like Empty, where the LLM shows good understanding of the environment and can make detailed suggestions, policy influencing is capable of contributing the most.
-However, as environment complexity increases and the LLM may not be as precise in policy influencing, it seems to pay off prioritizing reward shaping and having policy influencing contribute less.
+In simpler environments like Empty, where the LLM shows good understanding of the environment and can make accurate suggestions, policy influencing is capable of contributing the most.
+However, as environment complexity increases and the LLM may not be as precise in policy influencing, it seems to pay off to prioritize the more indirect method of reward shaping and having policy influencing contribute less.
 

### 2023-11-09 src/plot_rewards.py
Update plotrewards
Signed-off-by: Jonatan Hanssen <jonatan.hanssen@outlook.com>

@@ -8,6 +8,7 @@ parser = argparse.ArgumentParser()
 
 parser.add_argument("environment")
 parser.add_argument("-w", "--window", type=int, default=20)
+parser.add_argument("-l", "--limit", type=float, default=1.0)
 args = parser.parse_args()
 
 
@@ -87,7 +88,7 @@ print(f"Increase both over reward: {means_list[2] / means_list[0]}")
 print(f"Increase both over policy: {means_list[2] / means_list[1]}")
 
 
-plt.ylim(0, 1)
+plt.ylim(0, args.limit)
 plt.title(f"Average reward per episode\n on {args.environment} environment")
 plt.xlabel("Episodes")
 plt.ylabel("Reward")

### 2023-11-09 doc/paper.tex
paper mods

@@
 
-A more complicated environment can be seen in figure \ref{blockedunlockpickupenv}.
-Completing the goal requires picking up the ball and dropping it somewhere else, picking up the key, using it to unlock the door, dropping the key (the agent can only carry one item at once) and finally picking up the box.
-With no intermediate rewards, this is a very difficult task for traditional RL methods.
-The exact configuration of Minigrid environments is random; the position of the agent, the placement of the walls and items within the grid all vary from episode to episode.
-
-\begin{figure}[h]
-\centerline{\includegraphics[width=3.25in]{figure/blockedunlockpickupenv.png}}
-\caption{The Minigrid BlockedUnlockPickup environment}
-\label{blockedunlockpickupenv}
-\end{figure}
-
+With no intermediate rewards, more complex environments like the one shown in Figure \ref{unlockpickupenv} quickly become difficult to learn for traditional RL methods.
 Some environments can only be effectively solved by relying on language (for example, environments where the goal text describes a specific object to be picked up, which changes every time), while others can be solved without (for example, if the goal always is to pick up a key).
@@
 Despite its simplicity, there is still potential for LLM integration to improve the early phases of training, because the room is 15 by 15 tiles, which means that receiving any reward requires a significant amount of exploration.
-The environment is seen in figure \ref{emptyenv}.
+The environment is seen in Figure \ref{emptyenv}.
 
@@
 This environment consists of two rooms separated by a locked door.
-Fig \ref{doorkeyenv} shows an example configuration.
+Figure \ref{doorkeyenv} shows an example configuration.
 The agent starts in one room, while the goal square is in the other.
@@
 \subsubsection{UnlockPickup environment}
+\label{unlockpickupdescription}
 
 This environment is quite similar to the DoorKey environment.
-Fig \ref{unlockpickupenv} shows an example configuration.
+Figure \ref{unlockpickupenv} shows an example configuration.
 The difference is that after unlocking the door, the agent must now pick up a chest instead of walking to a goal square.
@@
 This added complexity makes it a much more difficult task than the previous two.
+
+\begin{figure}[h]
+\centerline{\includegraphics[width=3.25in]{figure/unlockpickupenv.png}}
+\caption{The Minigrid UnlockPickup environment}
+\label{unlockpickupenv}
+\end{figure}
 
@@
 In this complicated environment, it may that suboptimal responses from the LLM have high chance to ruin the agent's chance to complete the goal, for example by suggesting to open the box instead of picking it up.
-As explained earlier, this instantly loses the game by removing the box from the game.
+As explained in \ref{unlockpickupdescription}, this instantly loses the game by removing the box from the game.
 
-Our A2C2 performs much better than only policy influencing, but in this case we are not able to combine the two LLM integrations to receive a performance higher than the sum of its parts.
+Our A2C2 performs much better than policy influencing, but in this case we are not able to combine the two LLM integrations to receive a performance higher than the sum of its parts.
 Here we receive a score of NUMBER, which is higher than the baseline and policy influencing, but not higher than reward shaping by itself.
@@
 
-\begin{figure}[h]
-\centerline{\includegraphics[width=3.25in]{figure/unlockpickupenv.png}}
-\caption{The Minigrid UnlockPickup environment}
-\label{unlockpickupenv}
-\end{figure}
-
 \begin{table}[h]

### 2023-11-09 doc/paper.tex
Paper dogs
Signed-off-by: Jonatan Hanssen <jonatan.hanssen@outlook.com>

--- 2023-11-07 doc/paper.tex
@@
 \hline
-Baseline & 0.99 & 1 & 0\% \\
-Reward shaping & 0.99 & 1 & 30\% \\
-Policy influencing & 0.99 & 1 & -7\% \\
-Both & 0.99 & 1 & 36\% \\
+Reward shaping & 0.48 & 0.28 & 24\% \\
+Policy influencing & 0.36 & 0.31 & -6\% \\
+Both & 0.53 & 0.25 & 38\% \\
+Baseline & 0.38 & 0.31 & 0\% \\
 \end{tabular}
@@
 Looking at LLM policy influencing, we see that it has resulted in somewhat worse results than using the PPO baseline.
-Averaging per episode over all eight runs shows that LLM policy influencing only achieves $89\%$ of the reward amount per episode that the baseline achieves.
+Averaging per episode over all eight runs shows that LLM policy influencing only achieves $93\%$ of the reward amount per episode that the baseline achieves.
 Judging from the graph, it may seem that policy influencing may have given some initial learning advantage, but ultimately slows learning for the agent further on.
-The method also gives the highest standard deviation of reward given per episode of the three compared so far, at $0.25$.
 As with reward shaping, we have also used caching of LLM responses to significantly decrease the computational overhead cost to a minimum.
 
-However, the results from both of these methods are eclipsed by the results from combining LLM reward shaping and policy influencing together in our A2C2 model.
-Averaging over 8 runs, we observe a $34\%$ increase in reward from the environment compared to PPO alone, $8\%$ higher than reward shaping and $43\%$ higher than policy influencing, with the lowest standard deviation of all methods, at $0.22$.
+Regardless, the results from both of these methods are eclipsed by the results from combining LLM reward shaping and policy influencing together in our A2C2 model.
+Averaging over 8 runs, we observe a $38\%$ increase in reward from the environment compared to PPO alone, which is significantly higher than both policy influencing and reward shaping.
+Furthermore, it has the lowest standard deviation of all methods, at $0.25$.
 Clearly, the combination of immediate direct suggestions from the policy influencer and long term indirect suggestions from the reward shaper is a favourable way of integrating the LLM with PPO, which gives results which are better than the sum of its parts, not only in the simple Empty environment, but also in more complicated environments like DoorKey.
@@
 The Unlock Pickup environment is an even more complex environment than DoorKey, requiring a much longer string of actions to give out a reward.
-As may be anticipated, this added complexity makes the average reward much lower, and we see that all methods struggle (figure \ref{unlockpickupresults}).
+As may be anticipated, this added complexity makes the average reward much lower, and can see from Fig.
+\ref{unlockpickupresults} that all methods struggle (keep in mind the the y-axis now only goes to $0.15$, as opposed to the maximum reward possible of $1$, which the other graphs use).
 
@@
 \centerline{\includegraphics[width=3.25in]{figure/unlockpickupresults.png}}
-\caption{Reward per episode for the four different methods in the UnlockPickup environment}
+\caption{Reward per episode for the four different methods in the UnlockPickup environment. Note the y-axis, which only goes to $0.15$ as opposed to $1$}
 \label{unlockpickupresults}
@@
 
-As we can see, LLM performs the best overall, and performs much better than vanilla PPO, which struggles to learn the correct policy in this complex environment.
-Given the terrible performance of the baseline, LLM reward shaping achieved a percentage increase of over $700\%$, which is not that impressive when the average reward from the baseline was close to zero.
-The baseline achieved an average reward of $0.0089$ (with a standard deviation of $0.015$), while LLM reward shaping achieved an average reward of $0.062$ (with a standard deviation of $0.081$).
+As we can see, LLM reward shaping performs the best overall, and performs much better than vanilla PPO, which struggles to learn the correct policy in this complex environment.
+Given the terrible performance of the baseline, LLM reward shaping achieved a percentage increase of $275\%$, which is not that impressive when the average reward from the baseline was close to zero.
+The baseline achieved an average reward of $0.0097$, while LLM reward shaping achieved an average reward of $0.037$.
 See table \ref{unlockpickuptable}.
@@
 \hline
-Baseline & 0.99 & 1 & 0\% \\
-Reward shaping & 0.99 & 1 & 30\% \\
-Policy influencing & 0.99 & 1 & -7\% \\
-Both & 0.99 & 1 & 36\% \\
+Reward shaping & 0.037 & 0.082 & 275\% \\
+Policy influencing & 0.014 & 0.034 & 43\% \\
+Both & 0.031 & 0.072 & 222\% \\
+Baseline & 0.0097 & 0.025 & 0\% \\
 \end{tabular}
@@
 
-As is clear from the graph, policy influencing struggles in this environment as well, performing even worse than the baseline.
-Policy influencing received a mean average score of NUMBER, which is PERCENT lower than the baseline.
+Looking at figure \ref{unlockpickupresults}, we can see that policy influencing also struggles in this environment.
+From table \ref{unlockpickuptable}, we see a $43\%$ increase, although this is not very impressive given the low score of the baseline, and it is difficult to see an upward trend in the amount of reward received, which can be clearly seen in the reward shaping method.
+Policy influencing received a mean average score of $0.014$.
 In this complicated environment, it may that suboptimal responses from the LLM have high chance to ruin the agent's chance to complete the goal, for example by suggesting to open the box instead of picking it up.
 As explained earlier, this instantly loses the game by removing the box from the game.
+Thus, while the results are better than the baseline, they are still very poor and we do not consider policy influencing to be particularly beneficial in this environment.
 
 Our A2C2 performs much better than only policy influencing, but in this case we are not able to combine the two LLM integrations to receive a performance higher than the sum of its parts.
-Here we receive a score of NUMBER, which is higher than the baseline and policy influencing, but not higher than reward shaping by itself.
+Here we receive a score of $0.072$, which is higher than the baseline and policy influencing, but not higher than reward shaping by itself.
 It may be that in this case, the performance of policy influencing is so poor that the combination is not able to utilize the best of both methods, instead only being dragged down by poor suggestions.
@@
 
-On our final environment, we can see from table \ref{unlockpickupresponses}, such as suggestion to pick up a key when it is directly in front, and moving towards the goal although other observations are also present.
-We also see some suboptimal responses, such as suggesting to pick up a key when it is not directly in front and suggesting to pick up a key when it is already in the players inventory.
-These poor responses can explain why policy influencing can have a detrimental impact in this environment, as these actions will always be sampled slightly more often.
+On our final environment, we can see (Table \ref{unlockpickupresponses}) beneficial responses similar to the ones given in DoorKey.
+Here, we also see some absurd responses, which show that the LLM has completely misunderstood the prompt and few-shot examples.
+These responses are not particularly common, but they do occur sometimes and reveal some of the problems with running a relatively small LLM locally, as opposed to a larger model such as GPT4.
 

### 2023-11-09 doc/paper.tex
smolfiks

@@
 With no intermediate rewards, more complex environments like the one shown in Figure \ref{unlockpickupenv} quickly become difficult to learn for traditional RL methods.
-Some environments can only be effectively solved by relying on language (for example, environments where the goal text describes a specific object to be picked up, which changes every time), while others can be solved without (for example, if the goal always is to pick up a key).
+In addition, some environments can only be effectively solved by relying on language (for example, environments where the goal text describes a specific object to be picked up, which changes every time), while others can be solved without (for example, if the goal always is to pick up a key).
 We only use environments that can be solved without language, so that the baseline PPO algorithm is not at an unfair disadvantage.
@@
 
-\subsection{Societal Impact and Potential Harmful Consequences}
+\subsection*{Societal Impact and Potential Harmful Consequences}
 
@@
 
-\subsection{Impact Mitigation Measures}
+\subsection*{Impact Mitigation Measures}
 

### 2023-11-09 doc/paper.tex
add footnote
Signed-off-by: Jonatan Hanssen <jonatan.hanssen@outlook.com>
update paper?
Signed-off-by: Jonatan Hanssen <jonatan.hanssen@outlook.com>

@@
 We compare different methods for integrating an LLM into the Deep Learning RL algorithm Proximal Policy Optimization (PPO), and compare their efficiency against each other.
+\footnote{Code available at \href{https://github.com/jonatan-hanssen/in5490-project}{GitHub}}
 
@@
 The agent's state observation is encoded to natural text, and sent as part of a prompt to the LLM at every timestep.
-In addition to the state observation, the LLM is also asked to respond with a list of recommended actions for the agent to perform.
+After being given the state observation, the LLM is asked to respond with a list of recommended actions for the agent to perform.
 The agent then acts independently of the LLM for that timestep, as dictated by the baseline RL algorithm it is trained with.
 As it does so, its chosen action is encoded to natural text.
-Then, the LLM is used as a tool for semantic similarity comparison, comparing its own formerly proposed action to the action performed by the agent.
+Then, another LLM is used for semantic similarity comparison, comparing the proposed action to the action performed by the agent.
 Finally, high semantic similarity gives an additional pseudo-reward which is added to the reward of the environment.
@@
 
-In both these cases, we need two language models; one to generate the suggestions and one to generate semantic similarity values.
+In all these cases, we need two language models; one to generate the suggestions and one to generate semantic similarity values.
 
-For text generation, we have decided utilize an LLM from the Llama open-source family of language models developed by Meta AI, first released in Febraury, 2023.
+For text generation, we have decided to utilize an LLM from the Llama open-source family of language models developed by Meta AI, first released in Febraury, 2023.
 Specifically, we chose Llama 2, which has been trained using 40\% more data than its predecessor Llama 1, and has double the context length\cite{llama}.
-In the experiments presented below, we use the smallest version of Llama 2, with 7 billion parameters, which we download to run locally and reduce the bottleneck effect of having to call an API to a model hosted by someone else.
+In the experiments presented below, we use the smallest version of Llama 2, with 7 billion parameters, which we download to run locally and reduce the bottleneck effect and cost of having to call an API to a model hosted by someone else.
 The reason for not using any of the larger Llama 2 models is simply the hardware limitation we face, with the 7B model requiring 19GB of VRAM to be run.
 
-Some research papers propose fine-tuning of LLMs as means of leveraging the capabilites of the language models for decision-making purposes even further \cite{grounding}.
+Some research papers propose fine-tuning of LLMs as a means of leveraging the capabilites of the language models for decision-making purposes even further \cite{grounding}.
 Our plan of action, on the other hand, aims at evaluating the efficacy of the LLM without any previous pre-tuning.
@@
 
-For generating semantic similarity values, we use Sentence-Bert \cite{bert} for generating encoding vectors, and perform a cosine similarity calculation of the two sentence encodings.
+For generating semantic similarity values, we follow \cite{ellm} and use Sentence-Bert \cite{bert} for generating encoding vectors, and perform a cosine similarity calculation of the two sentence encodings.
 This gives a value between $-1$ and $1$, where two sentences with the exact same encoding have a cosine similarity of $1$.
@@
 We select three different Minigrid environments with increasing complexity; Empty, DoorKey and UnlockPickup.
-For each environment we test our four different models; PPO with no LLM integration, PPO with LLM reward shaping and PPO with LLM policy influencing.
+For each environment we test our four different models; PPO with no LLM integration, PPO with LLM reward shaping, PPO with LLM policy influencing and PPO with both reward shaping and policy influencing (A2C2).
 All methods use the same hyperparameters for the PPO, which can be found in the appendix.
@@
 
-In this simple environment, we see that both LLM reward shaping, LLM policy influencing and A2C2 outperform the baseline model.
+In this simple environment, we see from Figure \ref{emptyresults} that both LLM reward shaping, LLM policy influencing and A2C2 outperform the baseline model.
 Averaging over all episodes for all eight runs, we find that reward shaping increases the mean reward received by $10\%$ over the baseline, with a lower standard deviation in rewards received per run of $0.12$.
 Policy influencing scores even higher at $32\%$ over the baseline, with an even lower standard deviation of $0.09$.
-Finally, A2C2 performs the best of the four with a $37\%$ increase in average reward over the baseline model, and the lowest standard deviation of $0.04$
+Finally, A2C2 performs the best of the four with a $37\%$ increase in average reward over the baseline model, and the lowest standard deviation of $0.04$.
+See Table \ref{emptytable}.
 
@@
 \caption{Reward per episode for the four different methods in the Empty environment}
-\label{doorkeyresults}
+\label{emptyresults}
 \end{figure}
@@
 Both & 0.95 & 0.04 & 37\% \\
-Baseline & 0.69 & 0.28 & 0\% \\
+Baseline & 0.69 & 0.28 & - \\
 \end{tabular}
@@
 
-As we can see from Figure \ref{doorkeyresults}, we see impressive results from our LLM integration in this environment as well, especially from reward shaping and A2C2.
+As we can see from Figure \ref{doorkeyresults}, LLM integration gives impressive results in this environment as well, especially from reward shaping and A2C2.
 For the mean and standard deviation of all methods, see table \ref{doorkeytable}.
+
+\begin{figure}[h]
+\centerline{\includegraphics[width=3.25in]{figure/doorkeyresults.png}}
+\caption{Reward per episode for the four different methods in the DoorKey environment}
+\label{doorkeyresults}
+\end{figure}
 
@@
 Both & 0.53 & 0.25 & 38\% \\
-Baseline & 0.38 & 0.31 & 0\% \\
+Baseline & 0.38 & 0.31 & - \\
 \end{tabular}
@@
 
-\begin{figure}[h]
-\centerline{\includegraphics[width=3.25in]{figure/doorkeyresults.png}}
-\caption{Reward per episode for the four different methods in the DoorKey environment}
-\label{doorkeyresults}
-\end{figure}
-
 \subsubsection{Unlock Pickup environment}
@@
 The Unlock Pickup environment is an even more complex environment than DoorKey, requiring a much longer string of actions to give out a reward.
-As may be anticipated, this added complexity makes the average reward much lower, and can see from Fig.
-\ref{unlockpickupresults} that all methods struggle (keep in mind the the y-axis now only goes to $0.15$, as opposed to the maximum reward possible of $1$, which the other graphs use).
+As may be anticipated, this added complexity makes the average reward much lower, and we can see from Figure \ref{unlockpickupresults} that all methods struggle (keep in mind the the y-axis now only goes to $0.15$, as opposed to the maximum reward possible of $1$, which the other graphs use).
 
@@
 Both & 0.031 & 0.072 & 222\% \\
-Baseline & 0.0097 & 0.025 & 0\% \\
+Baseline & 0.0097 & 0.025 & - \\
 \end{tabular}
@@
 Looking at figure \ref{unlockpickupresults}, we can see that policy influencing also struggles in this environment.
-From table \ref{unlockpickuptable}, we see a $43\%$ increase, although this is not very impressive given the low score of the baseline, and it is difficult to see an upward trend in the amount of reward received, which can be clearly seen in the reward shaping method.
-Policy influencing received a mean average score of $0.014$.
-In this complicated environment, it may that suboptimal responses from the LLM have high chance to ruin the agent's chance to complete the goal, for example by suggesting to open the box instead of picking it up.
+From table \ref{unlockpickuptable}, we see a $43\%$ increase, although this is not very impressive given the low score of the baseline, and it is difficult to see an upward trend in the amount of reward received, which can be clearly seen in both reward shaping and A2C2.
+Policy influencing received a mean score of $0.014$.
+In this complicated environment, it may be that suboptimal responses from the LLM have high chance to ruin the agent's chance to complete the goal, for example by suggesting to open the box instead of picking it up.
 As explained earlier, this instantly loses the game by removing the box from the game.
@@
 Having looked at the performance of our different models, we can see that the suggestions from the LLM have positive effects in many cases, while also sometimes being detrimental.
-We now inspect some sample responses from the LLM, to gain deeper understanding of how the LLM is effecting the agent.
+We now inspect some sample responses from the LLM, to gain deeper understanding of how the LLM is affecting the agent.
 
@@
 \hline
-My goal is: get to the green goal square. I see a green goal left and forward & move forward. \\
+My goal is: get to the green goal square. I see nothing interesting. & move forward. \\
+My goal is: get to the green goal square. I see a green goal left and forward. & move forward. \\
 My goal is: get to the green goal square. I see a green goal left. & move left. \\
@@
 From table \ref{doorkeyresponses} we can see that some suggestions are good, such as suggestion to pick up a key when it is directly in front, and moving towards the goal although other observations are also present.
-We also see some suboptimal responses, such as suggesting to pick up a key when it is not directly in front and suggesting to pick up a key when it is already in the players inventory.
+We also see some suboptimal responses, such as suggesting to pick up a key when it is not directly in front and suggesting to pick up a key when it is already in the player's inventory.
 These poor responses can explain why policy influencing can have a detrimental impact in this environment, as these actions will always be sampled slightly more often.
@@
 These responses are not particularly common, but they do occur sometimes and reveal some of the problems with running a relatively small LLM locally, as opposed to a larger model such as GPT4.
+These problems have been attempted remedied through prompt engineering (see the Appendix).
 
@@
 \hline
-My goal is: pick up the blue box. I see a locked red door LEFT and FORWARD, see a red key directly in front. & pick up red key. \\
+My goal is: pick up the blue box. I see a locked red door left and forward, see a red key directly in front. & pick up red key. \\
 \hline
@@
 
+\subsection{Prompt given to Llama2}
+
+Below is the system prompt sent to Llama2, which is prepended before any state observation:
+
+{\it You are a helpful assistant giving advice to someone playing a videogame.
+You will recieve the current goal of the game and a list of observations about the objects in the environment and their positions relative to the player at one single timestep.
+You shall give a list of suggested actions that the player will take to reach their goal during this single timestep.
+The available actions to the player are: move forward, turn left and right, pick up items, drop items and use items.
+You MUST ALWAYS RESPOND with one or more of these actions.
+Your response MUST ALWAYS follow the template of the given example responses.
+NEVER ADD anything not in the example responses.
+You will NEVER ASSUME that other possible actions exist.
+You will NEVER ASSUME that the game is finished.
+You WILL NEVER make any other assumptions about the environment, only use the information given to you.
+If none of the observations are relevant to solving the task, respond that the player should turn around and explore more.
+Separate each suggestion with a new line.
+Be concise.
+ONLY FOLLOW THESE INSTRUCTIONS.
+NEVER RESPOND IN ANY OTHER WAY.
+OTHERWISE, YOU WILL BE SEVERELY PUNISHED.}
+
+\subsection{Figures and tables}
+
 \begin{table}[h]
@@
 Gamma & 0.99 \\
-Steps/episode & 1024 \\
-Epochs/episode & 10 \\
-Batches/epoch & 8 \\
-Clip & 0.2 \\
-Value loss coeff. & 0.3 \\
-Entropy coeff. & 0 \\
+Steps per episode & 1024 \\
+Epochs per episode & 10 \\
+Batches per epoch & 8 \\
+PPO Clipping value & 0.2 \\
+Value loss coefficient & 0.3 \\
+Entropy coefficient. & 0 \\
 Learning rate & 3.0e-4 \\

### 2023-11-09 doc/paper.tex
update paper
Signed-off-by: Jonatan Hanssen <jonatan.hanssen@outlook.com>

@@
 
-We would like to acknowledge the ROBIN research group at the University of Oslo Dept.
-of Informatics, who have arranged for us to be able to undertake this research and gain valuable experience.
+We would like to acknowledge the ROBIN research group at the University of Oslo Department of Informatics, who have arranged for us to be able to undertake this research and gain valuable experience.
 In particular, we would like to thank our supervisors Katrine Nergård and Kai Olav Ellefsen for their help and guidance in this project.
@@
 Value loss coefficient & 0.3 \\
-Entropy coefficient. & 0 \\
+Entropy coefficient & 0 \\
 Learning rate & 3.0e-4 \\

### 2023-11-09 doc/paper.tex
fix paper

--- 2023-11-09 doc/paper.tex
@@
 We compare different methods for integrating an LLM into the Deep Learning RL algorithm Proximal Policy Optimization (PPO), and compare their efficiency against each other.
+\footnote{Code available at \href{https://github.com/jonatan-hanssen/in5490-project}{github}}
 
@@
 Every time the LLM suggests performing a suboptimal action, this impacts the sampled action, regardless whether the PPO baseline has learned that another action is better or not.
-Over time, this may lead to the PPO not being able to explore other actions than the ones proposed by the LLM, thus making the PPO learn slower than if it was allowed to choose actions on its own.
-In other words, the LLM may have too much impact in the decision making when performing policy influencing, making the algorithm prone to errors from the LLM.
+Furthermore, as mentioned in Section \ref{emptyresultstext}, the logits increase in magnitude over time, while the contribution from the LLM does not scale with it.
+Since we cannot control the change in magnitude, this leads to unpredictable behavior from the model.
 
@@
 Therefore, a fitting way of dynamically changing the LLM contributions to the logits at some specified rate has to be found.
+
+A natural suggestion for how to improve the performance of both LLM policy influencing and A2C2 thus becomes to find a way to dynamically scale the influence from the LLM over time, so that the default scaling of logit contribution between the PPO and the LLM remains constant.
+This would allow for more predictability in which of the two contributes the most as learning progresses, and would make it possible to accurately customize how the rate of contribution from the LLM changes, for instance through simulated annealing.
 

### 2023-11-09 doc/paper.tex
fix it for bixit

@@
 The one we used, the 7B model of Llama 2, varies greatly in the quality of responses it produces, as demonstrated in \ref{llmresponse}.
-While is cost-effective both financially and in terms of computational resources, we expect that better results can be achieved with higher-performance models.
+While it is cost-effective both financially and in terms of computational resources, we expect that better results can be achieved with higher-performance models.
 This is especially true for policy influencing, where performance depends heavily on the LLM understanding the environment in detail.

### 2023-11-10 doc/paper.tex
Capitialize reference to figures
Signed-off-by: Jonatan Hanssen <jonatan.hanssen@outlook.com>

@@
 As we can see from Figure \ref{doorkeyresults}, LLM integration gives impressive results in this environment as well, especially from reward shaping and A2C2.
-For the mean and standard deviation of all methods, see table \ref{doorkeytable}.
+For the mean and standard deviation of all methods, see Table \ref{doorkeytable}.
 
@@
 The baseline achieved an average reward of $0.0097$, while LLM reward shaping achieved an average reward of $0.037$.
-See table \ref{unlockpickuptable}.
+See Table \ref{unlockpickuptable}.
 
@@
 
-Looking at figure \ref{unlockpickupresults}, we can see that policy influencing also struggles in this environment.
-From table \ref{unlockpickuptable}, we see a $43\%$ increase, although this is not very impressive given the low score of the baseline, and it is difficult to see an upward trend in the amount of reward received, which can be clearly seen in both reward shaping and A2C2.
+Looking at Figure \ref{unlockpickupresults}, we can see that policy influencing also struggles in this environment.
+From Table \ref{unlockpickuptable}, we see a $43\%$ increase, although this is not very impressive given the low score of the baseline, and it is difficult to see an upward trend in the amount of reward received, which can be clearly seen in both reward shaping and A2C2.
 Policy influencing received a mean score of $0.014$.
@@
 
-From table \ref{emptyresponses} we see that the suggestions are very good, which explain the great performance all LLM integrations show.
+From Table \ref{emptyresponses} we see that the suggestions are very good, which explain the great performance all LLM integrations show.
 However, we also see that this is a very simple environment, where one would expect an LLM to be able to respond coherently.
@@
 
-From table \ref{doorkeyresponses} we can see that some suggestions are good, such as suggestion to pick up a key when it is directly in front, and moving towards the goal although other observations are also present.
+From Table \ref{doorkeyresponses} we can see that some suggestions are good, such as suggestion to pick up a key when it is directly in front, and moving towards the goal although other observations are also present.
 We also see some suboptimal responses, such as suggesting to pick up a key when it is not directly in front and suggesting to pick up a key when it is already in the player's inventory.
@@
 Regardless of these shortcomings in the performance of policy influencing, it seems to have contributed positively when combining the two methods in A2C2, which outperformed each of the methods in both the Empty and DoorKey environments.
-As mentioned in \ref{emptyresultstext}, it seems that because the reward shaping makes the PPO learn the environment faster, it lessens the conflict between the influences of the LLM and the PPO itself on the logits.
+As mentioned in Section \ref{emptyresultstext}, it seems that because the reward shaping makes the PPO learn the environment faster, it lessens the conflict between the influences of the LLM and the PPO itself on the logits.
 This allows the PPO to exert the most influence on the logits earlier on.
@@
 Another potential improvement, perhaps the most obvious one, is to utilize a better LLM for generating responses, both for reward shaping and policy influencing.
-The one we used, the 7B model of Llama 2, varies greatly in the quality of responses it produces, as demonstrated in \ref{llmresponse}.
+The one we used, the 7B model of Llama 2, varies greatly in the quality of responses it produces, as demonstrated in Section \ref{llmresponse}.
 While it is cost-effective both financially and in terms of computational resources, we expect that better results can be achieved with higher-performance models.

### 2023-11-10 src/ActorCritic.py
Add ability to send thresholds
Signed-off-by: Jonatan Hanssen <jonatan.hanssen@outlook.com>

@@ -49,7 +49,7 @@ class Agent(nn.Module):
         the agent is in.
     """
 
-    def __init__(self, env, llama=False, generator=None):
+    def __init__(self, env, llama=False, generator=None, policy_sim_thres=None, policy_sim_mod=None):
         super(Agent, self).__init__()
         self.critic = nn.Sequential(
             init_weightsNbias(nn.Linear(147 * 11, 64, dtype=torch.float64)),
@@ -70,7 +70,9 @@ class Agent(nn.Module):
         )
 
         goal = env.reset()[0]["mission"]
-        self.consigliere = llama2_policy(goal, cos_sim_threshold=0.84, similarity_modifier=0.1, generator=generator) if llama else None
+        threshold = 0.84 if policy_sim_thres is None else policy_sim_thres
+        modifier = 0.01 if policy_sim_mod is None else policy_sim_mod
+        self.consigliere = llama2_policy(goal, cos_sim_threshold=threshold, similarity_modifier=modifier, generator=generator) if llama else None
 
 
     def get_value(self, observation):
@@ -88,7 +90,6 @@ class Agent(nn.Module):
         logits = self.actor(observation_onehot)
 
         if self.consigliere:
-            #print(observation.shape)
             # logits /= torch.norm(logits)
             if len(observation.shape) == 2:
                 advisor_values_list = list()
@@ -105,16 +106,8 @@ class Agent(nn.Module):
                 #self.consigliere.suggest(unflat_obs)
                 advisor_values = self.consigliere.give_values(np.array(unflat_obs.cpu())).to(torch.float64)
 
-            # print(f"{logits=}")
-            # print(f"{advisor_values=}")
-            # # anti adrian propaganda
-            # logits += advisor_values
-            # adrian good vote adrian
-            # logits *= advisor_values
-
         probs = Categorical(logits=logits)
 
-        # print(probs.probs)
 
         if action is None:
             if self.consigliere:
@@ -128,17 +121,10 @@ class Agent(nn.Module):
                             anneal = 0
                     else:
                         anneal = 1
-                    # print(anneal)
                     scale_factor = (torch.norm(logits) / torch.norm(advisor_values))
-                    # print(f"{advisor_values=}")
-                    # print(f"{scale_factor=}")
-                    # print(f"{advisor_values * scale_factor=}")
-                    # print(f"{logits=}")
                     advisor_values = advisor_values.to(torch.device("cuda"))
[…]

### 2023-11-10 src/ppo.py

@@ -25,6 +25,8 @@ class PPO:
         policy=None,
         generator=None,
         allow_farming=False,
+        policy_sim_thres=None,
+        policy_sim_mod=None,
     ):
         self.args = read_params(param_file)
 
@@ -41,7 +43,7 @@ class PPO:
 
         self.device = torch.device("cuda" if torch.cuda.is_available() else "cpu")
         # self.device = torch.device("cpu")
-        self.agent = Agent(self.env, self.llama_policy, generator=generator).to(
+        self.agent = Agent(self.env, self.llama_policy, generator=generator, policy_sim_thres=policy_sim_thres, policy_sim_mod=policy_sim_mod).to(
             self.device
         )
         self.optimizer = optim.Adam(

### 2023-11-13 doc/paper.tex
Update paper with peer review changes
Signed-off-by: Jonatan Hanssen <jonatan.hanssen@outlook.com>

@@
 
-The agent is trained as in the baseline.
+In this method, the agent is trained using the same method as in the baseline.
 The agent's state observation is encoded to natural text, and sent as part of a prompt to the LLM at every timestep.
@@
 Looking at LLM policy influencing, we see that it has resulted in somewhat worse results than using the PPO baseline.
-Averaging per episode over all eight runs shows that LLM policy influencing only achieves $93\%$ of the reward amount per episode that the baseline achieves.
+Averaging over all eight runs shows that LLM policy influencing only achieves $93\%$ of the reward amount the baseline receives.
 Judging from the graph, it may seem that policy influencing may have given some initial learning advantage, but ultimately slows learning for the agent further on.
@@
 
-As we can see, LLM reward shaping performs the best overall, and performs much better than vanilla PPO, which struggles to learn the correct policy in this complex environment.
+As we can see, LLM reward shaping performs the best overall, and performs much better than the baseline PPO, which struggles to learn the correct policy in this complex environment.
 Given the terrible performance of the baseline, LLM reward shaping achieved a percentage increase of $275\%$, which is not that impressive when the average reward from the baseline was close to zero.
@@
 
-As shown in the previously referenced work (\ref{background}), the integration of a Large Language Model into the Reinforcement Learning framework can improve network performance.
+As shown in the previously referenced work (Section \ref{background}), the integration of a Large Language Model into the Reinforcement Learning framework can improve network performance.
 There are various ways to do this, and our work attempts to compare two main methodologies of integrating the LLM as well as the combination of the two, namely through reward shaping and policy influencing.

### 2023-11-13 doc/paper.tex
Add appendix ref
Signed-off-by: Jonatan Hanssen <jonatan.hanssen@outlook.com>

@@
 
-For the implementation details of our PPO and the hyperparameters used, see the Appendix.
+For the hyperparameters used in our PPO implementation, see the \nameref{sec:appendixtext}.
 
@@
 For text generation, we have decided to utilize an LLM from the Llama open-source family of language models developed by Meta AI, first released in Febraury, 2023.
-Specifically, we chose Llama 2, which has been trained using 40\% more data than its predecessor Llama 1, and has double the context length\cite{llama}.
+Specifically, we choose Llama 2, which has been trained using 40\% more data than its predecessor Llama 1, and has double the context length\cite{llama}.
 In the experiments presented below, we use the smallest version of Llama 2, with 7 billion parameters, which we download to run locally and reduce the bottleneck effect and cost of having to call an API to a model hosted by someone else.
@@
 These responses are not particularly common, but they do occur sometimes and reveal some of the problems with running a relatively small LLM locally, as opposed to a larger model such as GPT4.
-These problems have been attempted remedied through prompt engineering (see the Appendix).
+These problems have been attempted remedied through prompt engineering (see the \nameref{sec:appendixtext}).
 
@@
 \section*{Appendix}
+\label{sec:appendixtext}
 

### 2023-11-14 doc/paper.tex
Update paper
Signed-off-by: Jonatan Hanssen <jonatan.hanssen@outlook.com>

@@
 In recent years, LLMs have been successfully used to aid RL agents in more efficient exploration, and have allowed them to solve problems which previous methods have been unable to.
-We compare different methods for integrating an LLM into the Deep Learning RL algorithm Proximal Policy Optimization (PPO), and compare their efficiency against each other.
-\footnote{Code available at \href{https://github.com/jonatan-hanssen/in5490-project}{GitHub}}
+We compare different methods for integrating an LLM into the Deep Learning RL algorithm Proximal Policy Optimization (PPO), and compare their efficiency against each other.\footnote{Code available at \href{https://github.com/jonatan-hanssen/in5490-project}{GitHub}}
 
@@
 
-RL informed by natural language is a relatively new field, where research before 2018 have been limited to relatively small corpora or synthetic language \cite[1]{survey}.
+RL informed by natural language is a relatively new field, where research before 2018 had been limited to relatively small corpora or synthetic language \cite[1]{survey}.
 With the introduction of Large Language Models, many new possibilities have opened up, and in recent years there have been many successful integrations of these models into traditional RL \cite{omni, ellm, idm}.
@@
 
-The algorithm proposed in this paper involves prompting the LLM with the agent's state observation, a list of tasks the agent already does well, and a list of available tasks for the LLM to deem as either interesting or boring.
-After sorting available tasks into interesting and boring ones, it assigns sampling weights to them, giving much higher weights to interesting tasks than boring ones, thus directly influencing the agent's policy by altering probabilities of which action it then chooses to do.
+The algorithm proposed in this paper involves prompting the LLM with the agent's state observation, a list of tasks the agent already does well, and a list of available tasks for the LLM to deem as either "interesting" or "boring".
+After sorting available tasks into interesting and boring ones, it assigns sampling weights to them, giving much higher weights to interesting tasks than boring ones.
+Thus, it directly influences the agent's policy by altering the probabilities of the possible actions.
 In \cite{idm}, another paper proposing using the LLM as a policy influencer, has a similar approach in which the state observation is tokenized and passed to the LLM, though in their experiment, goals and action histories are also tokenized and passed.

### 2023-11-14 doc/paper.tex
update paper
Signed-off-by: Jonatan Hanssen <jonatan.hanssen@outlook.com>

@@
 Furthermore, in many problems the state-action spaces are enormous, meaning that uniform exploration is unlikely to find good solutions in a reasonable amount of time.
-Many methods have been developed to deal with this issue, for example by introducing auxilliary reward functions that use domain knowledge to reward actions which are considered good, or which reward the agent for learning novel skills.
+Many methods have been developed to deal with this issue, for example by introducing auxiliary reward functions that use domain knowledge to reward actions which are considered good, or which reward the agent for learning novel skills.
 However, hand picking which actions to reward can lead to imitation rather than optimal behaviour, and novelty is not always useful \cite[1]{ellm}.
@@
 
-These papers explore different methods of integrating LLMs into a standard reinforcement learning training loop, from altering the policy directly to introducing an auxilliary reward function.
+These papers explore different methods of integrating LLMs into a standard reinforcement learning training loop, from altering the policy directly to introducing an auxiliary reward function.
 In this paper, we compare both methods in the same environment using the same deep reinforcement algorithm (PPO), and compare their results to each other to explore how LLMs best can be integrated into RL.
@@
 Our baseline is the PPO algorithm, which is a policy gradient method that has been established as a state-of-art approach to solving reinforcement learning problems over the last few years.
-The main idea behind this algorithm is that the new policy is clipped so that it does not move too quickly and too far from the old policy, which helps to avoid changing the policy too drastically during the update.
+PPO simply uses an Actor-Critic network which is updated using policy gradient methods, with the modification that the policy is not allowed to change too drastically when updating.
 With this simple concept, PPO has outperformed previous methods such as Deep-Q Networks and A2C, while being relatively simple to implement\cite{ppo}.
@@
 Finally, high semantic similarity gives an additional pseudo-reward which is added to the reward of the environment.
-Compared to the policy influencing method above, this approach is less direct, as the policy is only indirectly affected by the LLM through these additional rewards, which only take effect when the policy is updated through backpropogation after an entire episode is completed.
+Compared to the policy influencing method above, this approach is less direct, as the policy is only indirectly affected by the LLM through these additional rewards, which only take effect when the policy is updated through backpropagation after an entire episode is completed.
 Thus, this method functions more as an advisor, influencing the policy in a more long term capacity than the coaching of the policy influencer.
@@
 Finally, we combine the two LLM integrations, hoping to combine the direct influence of our policy integration with the more indirect reward shaping.
-Thus, we are affecting the policy in two ways: Through reward shaping, we are causing the policy to update to better attain the aforementioned pseudo-rewards, which will encourage completing of subgoals which are necessary to complete the actual goal of the environment, but which do not give out intermediate rewards.
+Thus, we are affecting the policy in two ways: Through reward shaping, we are causing the policy to update to better attain the aforementioned pseudo-rewards, which will encourage completion of subgoals that are necessary to complete the actual goal of the environment, but which do not give out intermediate rewards.
 Through policy influencing, we are nudging the agent to take favorable actions by directly altering the action probabilities, which will make it more likely that the agent completes the given task and receives an actual reward from the environment.
 Thus, the agent will more quickly discover the correct policy.
-We call this the \textbf{Actor-Advisor-Critic-Coach} method, or \textbf{A2C2} for short.
+We call this the \textbf{Actor-Advisor-Coach-Critic} method, or \textbf{A2C2} for short.
 
@@
 
-Some research papers propose fine-tuning of LLMs as a means of leveraging the capabilites of the language models for decision-making purposes even further \cite{grounding}.
+Some research papers propose fine-tuning of LLMs as a means of leveraging the capabilities of the language models for decision-making purposes even further \cite{grounding}.
 Our plan of action, on the other hand, aims at evaluating the efficacy of the LLM without any previous pre-tuning.
 From the perspective of reinforcement learning, this is the appropriate course of action, as the LLM should not possess any previous knowledge about the task at hand, just like the agent should not.
-In addtion, the act of fine-tuning LLM is a rather time-comsuming process, requiring the generation and structuring of a new dataset, and also vast hardware resources.
+In addition, the act of fine-tuning LLM is a rather time-consuming process, requiring the generation and structuring of a new dataset, and also vast hardware resources.
 
@@
 Even state-of-art RL algorithms may struggle in such environments.
-Therefore, it is the perfect arena to see if the high generalization capabilites of pre-trained Large Language Models will lead to better results over using only RL methods.
+Therefore, it is the perfect arena to see if the high generalization capabilities of pre-trained Large Language Models will lead to better results over using only RL methods.
 
@@
 These environments are partially observable, which means that the agent only receives a seven by seven observation grid, not the entire grid.
-Figure \ref{doorkeyenv}, shows a typical minigrid environment.
+Figure \ref{doorkeyenv}, shows a typical Minigrid environment.
 In this case, the agent (the red arrow) only sees the left room, and does not know where the goal is.
@@
 Policy influencing & 0.91 & 0.09 & 32\% \\
-Both & 0.95 & 0.04 & 37\% \\
+A2C2 & 0.95 & 0.04 & 37\% \\
 Baseline & 0.69 & 0.28 & - \\
@@
 Policy influencing & 0.36 & 0.31 & -6\% \\
-Both & 0.53 & 0.25 & 38\% \\
+A2C2 & 0.53 & 0.25 & 38\% \\
 Baseline & 0.38 & 0.31 & - \\
@@
 
-LLM reward shaping gives a noticable improvement to the performance of the model, not just in the early stages of training, but throughout the entire training process.
+LLM reward shaping gives a noticeable improvement to the performance of the model, not just in the early stages of training, but throughout the entire training process.
 Averaging over all episodes of all eight runs per model, we find that LLM reward shaping receives $24\%$ more reward per episode than PPO trained without LLM rewards.
@@
 Policy influencing & 0.014 & 0.034 & 43\% \\
-Both & 0.031 & 0.072 & 222\% \\
+A2C2 & 0.031 & 0.072 & 222\% \\
 Baseline & 0.0097 & 0.025 & - \\

### 2023-11-14 doc/paper.tex
do good things

@@
 However, we notice from the graph that policy influencing alone consistently gives a significant dip in received reward after the first few episodes, which it later compensates for, following a similar trajectory to that of reward shaping after about 50 episodes.
-We believe this could be a sign of the policy influencing being somewhat overly influential in changing the logits early on, because the logits initially have low values.
-As the PPO learns the environment, the magnitude of the logits increases, which at some point leads to conflicting contributions to the logits from the LLM and from the PPO.
-As the dip stops and performance increases again, this is likely due to the PPO becoming more influencing than the LLM after having started to learn the environment properly.
-We also notice that this dip is completely removed with A2C2, which could signify that the added reward shaping helps the PPO learn faster, enabling it to increase the influence from PPO on the logits faster.
+We believe this happens due to the magnitude of the logits increasing over time, meaning the LLM suggestions contribute less and the baseline PPO starts to overtake the majority of influence in actions being taken.
+Because the policy hasn't trained sufficiently yet, this makes the agent perform worse than when the LLM contributed more.
+We discuss this more in depth in the \nameref{sec:appendixtext}.
+We also notice that this dip is almost completely removed with A2C2, which might be because reward shaping leads to the policy network being better trained by the time the logits increase.
 

### 2024-01-24 src/ActorCritic.py
Update code with comments
Signed-off-by: Jonatan Hanssen <jonatan.hanssen@outlook.com>

@@ -34,13 +34,6 @@ def load_model(model):
     model.load_state_dict(torch.load(file_path))
 
 
-def checkpoint():
-    # TODO: Implement a function for saving model when the perforamce reaches a certain level.
-    # Intention is to use the "save_model(model)" method
-    raise NotImplementedError()
-
-
-# TODO: Transfer model to device a.k.a. to GPU
 class Agent(nn.Module):
     """
     Descr:
@@ -72,6 +65,7 @@ class Agent(nn.Module):
         goal = env.reset()[0]["mission"]
         threshold = 0.84 if policy_sim_thres is None else policy_sim_thres
         modifier = 0.01 if policy_sim_mod is None else policy_sim_mod
+        # consigliere is a different name for coach
         self.consigliere = llama2_policy(goal, cos_sim_threshold=threshold, similarity_modifier=modifier, generator=generator) if llama else None
 
 
@@ -104,6 +98,9 @@ class Agent(nn.Module):
             else:
                 unflat_obs = observation.reshape((7, 7, 3)).to(torch.int64)
                 #self.consigliere.suggest(unflat_obs)
+
+                # consigliere, which is the coach, gives advisor values instead of coach values.
+                # This is done to confuse the reader
                 advisor_values = self.consigliere.give_values(np.array(unflat_obs.cpu())).to(torch.float64)
 
         probs = Categorical(logits=logits)

### 2024-01-24 src/multi_ppo.py

@@ -1,89 +1,33 @@
 from ppo import PPO
-from utils import llama2_policy
+from llama import Llama
+import os
 
-consigliere = llama2_policy("use the key to open the door and then get to the goal", cos_sim_threshold=0, similarity_modifier=0.1)
+base_path = os.path.dirname(__file__)
 
-print("Starting policy 1")
-ppo = PPO(
-        "hyperparams.json",
-        result_file="data/doorkey_policy1",
-        reward=False,
-        policy=True,
-        consigliere=consigliere,
-    )
-ppo.agent.consigliere.reset_cache()
-ppo.train()
+generator = Llama.build(
+    ckpt_dir=os.path.join(base_path, "../llama-2-7b-chat"),
+    tokenizer_path=os.path.join(
+        base_path, "../llama-2-7b-chat/tokenizer.model"
+    ),
+    max_seq_len=2048,
+    max_batch_size=6,
+)
 
-print("Starting policy 2")
-ppo = PPO(
-        "hyperparams.json",
-        result_file="data/doorkey_policy2",
-        reward=False,
-        policy=True,
-        consigliere=consigliere,
-    )
-ppo.agent.consigliere.reset_cache()
-ppo.train()
+# make the generator something stupid if you think it will never be called
+# because of caching. Good if you want to run it with out a big gpu
+# generator = "a straight up dog"
 
-print("Starting policy 3")
-ppo = PPO(
-        "hyperparams.json",
-        result_file="data/doorkey_policy3",
-        reward=False,
-        policy=True,
-        consigliere=consigliere,
-    )
-ppo.agent.consigliere.reset_cache()
-ppo.train()
 
-print("Starting policy 4")
-ppo = PPO(
-        "hyperparams.json",
-        result_file="data/doorkey_policy4",
-        reward=False,
-        policy=True,
-        consigliere=consigliere,
-    )
-ppo.agent.consigliere.reset_cache()
-ppo.train()
-
-print("Starting policy 5")
-ppo = PPO(
-        "hyperparams.json",
-        result_file="data/doorkey_policy5",
-        reward=False,
-        policy=True,
-        consigliere=consigliere,
-    )
-ppo.agent.consigliere.reset_cache()
-ppo.train()
-
-print("Starting policy 6")
-ppo = PPO(
-        "hyperparams.json",
-        result_file="data/doorkey_policy6",
-        reward=False,
-        policy=True,
-        consigliere=consigliere,
-    )
-ppo.train()
-
-print("Starting policy 7")
[…]

### 2024-01-24 src/ppo.py

@@ -318,8 +318,8 @@ if __name__ == "__main__":
     parser.add_argument(
         "-p", "--policy", type=bool, help="Use LLM policy", default=None
     )
-    parser.add_argument("-a", "--allow_farming", action="store_true")
-    parser.add_argument("-d", "--doobie", action="store_true")
+    parser.add_argument("-a", "--allow_farming", action="store_true", help="Allow for getting multiple rewards for the same observation. Should be off for reward shaping and on for policy shaping.")
+    parser.add_argument("-d", "--doobie", action="store_true", help="Fake the generator, hope that everything is cached so that it is never called. If it is called the program will fail. Only use this if you are sure that all possible observations have been cached.")
 
     args = parser.parse_args()
     print(args)

### 2024-01-24 src/utils.py

@@ -194,16 +194,12 @@ class llama2_base:
         else:
             # print("Cache miss")
             self.cache_misses += 1
-            try:
-                result = self.generator.chat_completion(
-                    [self.dialog],  # type: ignore
-                    max_gen_len=100,
-                    temperature=self.temperature,
-                    top_p=self.top_p,
-                )
-            except AttributeError:
-                print("im smoking doobies")
-                result = [{"generation": {"content": "i smoking that zaza"}}]
+            result = self.generator.chat_completion(
+                [self.dialog],  # type: ignore
+                max_gen_len=100,
+                temperature=self.temperature,
+                top_p=self.top_p,
+            )
 
             answer = result[0]["generation"]["content"]
 

### 2024-01-24 doc/paper.tex
Updape paper
Signed-off-by: Jonatan Hanssen <jonatan.hanssen@outlook.com>

@@
 
+\begin{figure}[h]
+\centerline{\includegraphics[width=3.25in]{figure/unlockpickupenv.png}}
+\caption{The Minigrid UnlockPickup environment}
+\label{unlockpickupenv}
+\end{figure}
+
 \subsection{Our chosen environments}
@@
 For each environment we test our four different models; PPO with no LLM integration, PPO with LLM reward shaping, PPO with LLM policy influencing and PPO with both reward shaping and policy influencing (A2C2).
-All methods use the same hyperparameters for the PPO, which can be found in the appendix.
+All methods use the same hyperparameters for the PPO, which can be found in the \nameref{sec:appendixtext}.
 For each of these methods, we run PPO for 2000 (200 for Empty) episodes of 1024 steps, and track the reward\footnote{The true environment reward, not including pseudo-rewards given by the LLM in reward shaping} received for each one.
@@
 This added complexity makes it a much more difficult task than the previous two.
-
-\begin{figure}[h]
-\centerline{\includegraphics[width=3.25in]{figure/unlockpickupenv.png}}
-\caption{The Minigrid UnlockPickup environment}
-\label{unlockpickupenv}
-\end{figure}
 
@@
 This allows the PPO to exert the most influence on the logits earlier on.
-However, this conflict is not entirely negated, as witnessed in the UnlockPickup environment where reward shaping still vastly outperforms both policy influencing and A2C2.
+However, this conflict is not entirely negated, as witnessed in the UnlockPickup environment where reward shaping still outperforms both policy influencing and A2C2.
 
